---
title: "Harmonize de Smith clinical data to LinkML v2 model - March 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()
```

Read in data:

```{r}
desmith_clinical_source <- read_xlsx(synGet("syn44271573")$path)
```

Replace headers with working_headers
See Google sheet `INCLUDE clinical data dictionaries` -> `X01-deSmith`
```{r}
gs4_deauth()

desmith_working_headers <- read_sheet(
  "https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw/edit#gid=115820483",
  sheet = "X01-deSmith")

desmith_clinical_source <- setNames(desmith_clinical_source, desmith_working_headers$working_header)
```


Cols that are <chr> due to "nk" but should contain numbers:
gestational_age_at_birth_weeks
birthweight_kg
age_at_last_known_vital_status_days
platelets_at_birth
peripheral_blood_blasts_at_birth

- replace nk with `NA`
- convert misclassified cols to numeric

**TODO: Fix issues with decimal places**
```{r}
# use datapasta to make vector
misclassified <- c("gestational_age_at_birth_weeks", "Birthweight", "age_at_last_known_vital_status_days", "platelets_at_birth", "peripheral_blood_blasts_at_birth")

desmith_clinical_harmonized <- desmith_clinical_source %>% 
  mutate(across(matches(misclassified), ~na_if(., "nk"))) %>% 
  mutate(across(matches(misclassified), ~as.numeric(.))) %>% 
  rename(participantExternalId = wpsid) 

``` 


Get ETL plan
```{r}
# source("harmonization_setup_2023-03.R")

desmith_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "X01-deSmith")
```



##### PARTICIPANT

Remove Condition & Measure cols
```{r}
desmith_participant <- desmith_clinical_harmonized %>% 
  select(participantExternalId, age_at_enrollment, sex_original, race_ethnicity, 
         vital_status, age_at_last_known_vital_status_days) %>% 
  distinct()
```


Sex, Race (Ethnicity = Unknown):
(note - error in Race of OX257 - "white" lowercase - has been added to ETL plan)

```{r}
desmith_sex_lookup_values <- desmith_etl_plan %>% 
  filter(model_slot == "sex") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

desmith_race_lookup_values <- desmith_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

desmith_ethnicity_lookup_values <- desmith_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

desmith_vital_lookup_values <- desmith_etl_plan %>% 
  filter(model_slot == "outcomesVitalStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

desmith_participant <- desmith_participant %>% 
  left_join(desmith_race_lookup_values, by = c("race_ethnicity" = "original_value")) %>% 
  rename(race = model_enum_label) %>%
  left_join(desmith_ethnicity_lookup_values, by = c("race_ethnicity" = "original_value")) %>% 
  rename(ethnicity = model_enum_label) %>% 
  left_join(desmith_sex_lookup_values, by = c("sex_original" = "original_value")) %>% 
  rename(sex = model_enum_label)  %>% 
  left_join(desmith_vital_lookup_values, by = c("vital_status" = "original_value")) %>%
  rename(outcomesVitalStatus = model_enum_label)

#%>% select(participantExternalId, race_ethnicity, race, ethnicity, sex_original, sex)
```

Get Global IDs:

```{r}
desmith_dewrangle_input <- desmith_participant %>% 
  distinct(participantExternalId) %>% 
  rename(descriptor = participantExternalId) %>% 
  mutate(fhir_resource_type = "patient")

write_csv(desmith_dewrangle_input, 
          here::here("output", "desmith_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

desmith_participant_global_ids <- read_csv(here::here("output", "global-descriptors-X01-deSmith.csv")) %>% 
  filter(fhir_resource_type == "Patient")

desmith_participant <- desmith_participant %>% 
  left_join(desmith_global_ids %>% select(descriptor, global_id),
            by = c("participantExternalId" = "descriptor")) 

#%>% select(participantExternalId, global_id) %>% filter(is.na(global_id))
```


Create final Participant table:

- Fill in info that is same for every participant (mutate)
- Rename columns that need renamed with schema labels (also mutate)
- select all Participant columns
- rename with display names

```{r}
desmith_participant <- desmith_participant %>% 
  mutate(studyCode = "X01-deSmith",
         participantGlobalId = global_id,
         familyId = NA,
         familyType = "Proband-only",
         fatherId = NA,
         motherId = NA,
         siblingId = NA,
         otherFamilyMemberId = NA,
         familyRelationship = "Proband",
         ethnicity = "Unknown",
         downSyndromeStatus = "T21",
         ageAtFirstPatientEngagement = age_at_enrollment,
         firstPatientEngagementEvent = "Enrollment",
         ageAtLastVitalStatus = age_at_last_known_vital_status_days) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)

##TODO: check against data
```

```{r}
write_csv(desmith_participant, here::here("output", "desmith_participant.csv"))
```


##### CONDITION

Get condition & measure codes
```{r}
condition_lookup_desmith <- condition_codes_v3 %>% filter(study == "X01-deSmith") %>% 
  unnest(source_column_value)
measure_lookup_desmith <- measure_codes_v3 %>% filter(study == "X01-deSmith")
```

```{r}
desmith_conditions <- desmith_clinical_harmonized %>% 
  select(participantExternalId, gata1_mutation, clinical_tam, ml_ds, ds_all, 
         maternal_diabetes, congenital_heart_defects, other_birth_defects,
         thrombocytopenia, anemia, polycythemia) %>%
  #all pts have Down syndrome
  mutate(ds_diagnosis = "Down syndrome") %>% 
  pivot_longer(cols = c(-participantExternalId),
               names_to = "source_text",
               values_to = "status") %>% 
  left_join(condition_lookup_desmith, by = c("source_text" = "source_column",
                                             "status" = "source_column_value")) %>%
  filter(!is.na(conditionMeasureSourceText)) %>% 
  select(-study)

```

Create empty/all same value columns
Select columns in order of DCA template
```{r}
desmith_conditions <- desmith_conditions %>% 
  arrange(participantExternalId) %>% 
  mutate(studyCode = "X01-deSmith",
         eventId = NA,
         eventType = NA,
         ageAtConditionMeasureObservation = NA,
         conditionInterpretation = "Observed",
         conditionStatus = "History Of",
         conditionDataSource = "Clinical",
         measureValue = NA,
         measureUnit = NA) %>%
  select(all_of(condition_schemalabels))

```

Measures:
**NOTE: decimal number vary when displayed in R and Excel**
**Actual numbers match up in output csv and Excel when you click on the cell**

```{r}
desmith_measures <- desmith_clinical_harmonized %>% 
  select(participantExternalId, gestational_age_at_birth_weeks, birthweight_kg, hb_at_birth,
         wbc_at_birth, platelets_at_birth, peripheral_blood_blasts_at_birth, hematocrit) %>%
  pivot_longer(cols = c(-participantExternalId),
               names_to = "source_text",
               values_to = "measureValue") %>% 
  left_join(measure_lookup_desmith, by = c("source_text" = "source_column")) %>% 
  mutate(studyCode = "X01-deSmith",
         eventId = NA,
         eventType = NA,
         ageAtConditionMeasureObservation = NA,
         conditionInterpretation = NA,
         conditionStatus = NA, 
         conditionDataSource = NA,
         hpoLabel = NA,
         hpoCode = NA,
         mondoLabel = NA,
         mondoCode = NA,
         maxoLabel = NA,
         maxoCode = NA,
  ) %>% 
  select(all_of(condition_schemalabels))

#TODO: how to preserve decimal places after pivot_longer
```


Join conditions & measures, rename with display names

```{r}
desmith_condition <- desmith_conditions %>% 
  rbind(desmith_measures) %>% 
  arrange(participantExternalId) %>% 
  setNames(condition_displaynames)
```


```{r}
write_csv(desmith_condition, here::here("output", "desmith_condition.csv"))
```


##### BIOSPECIMEN

"whole blood DNA from EDTA tube"

```{r}
desmith_biospecimen <- desmith_clinical_source %>% 
  select(wpsid, sample, analyte_type, experimental_strategy) %>% 
  mutate(studyCode = "X01-deSmith",
         participantExternalId = wpsid,
         sampleGlobalId = NA,
         sampleExternalId = paste0(wpsid, "_", analyte_type),
         sampleType = analyte_type,
         ageAtBiospecimenCollection = 0,
         parentSampleGlobalId = NA,
         parentSampleExternalId = NA,
         parentSampleType = NA,
         collectionGlobalId = NA,
         collectionExternalId = paste0(wpsid, "_", str_replace(sample, " b", "_b")),
         collectionSampleType = "Peripheral Whole Blood",
         containerGlobalId = NA,
         containerExternalId = NA,
         volume = NA,
         volumeUnit = NA,
         concentration = NA,
         concentrationUnit = NA,
         laboratoryProcedure = NA,
         biospecimenStorage = NA,
         sampleAvailability = "Unavailable") %>% 
  select(all_of(biospecimen_schemalabels)) %>% 
  setNames(biospecimen_displaynames)

```

Get Sample Global IDs:

```{r}
desmith_sample_dewrangle_input <- desmith_biospecimen %>% 
  select(descriptor = `Sample External ID`) %>% 
  mutate(fhir_resource_type = "specimen")

write_csv(desmith_sample_dewrangle_input, 
          here::here("output", "desmith_sample_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

desmith_sample_global_ids <- read_csv(here::here("output", "global-descriptors-X01-deSmith.csv")) %>% 
  filter(fhir_resource_type == "Specimen")

desmith_biospecimen <- desmith_biospecimen %>% 
  left_join(desmith_sample_global_ids %>% select(descriptor, global_id),
            by = c("Sample External ID" = "descriptor")) %>% 
  mutate(`Sample Global ID` = global_id) %>% 
  select(-global_id)

#%>% select(participantExternalId, global_id) %>% filter(is.na(global_id))
```

```{r}
write_csv(desmith_biospecimen, here::here("output", "desmith_biospecimen_DRAFT.csv"))
```

##### dbGaP SUBMISSION FILES

1. Subject-Consent Mapping File
SUBJECT_ID, CONSENT, SEX

2. Dictionary for the Subject-Consent Mapping File

```{r}
participants_to_delete <- c("OX112", "OX118", "OX239", "OX275", "OX367", "OX514")


desmith_subject_consent_DS <- desmith_clinical_source %>% 
  mutate(SUBJECT_ID = wpsid,
         CONSENT = "1",
         SEX = case_when(
           sex_original == "M" ~ "1",
           sex_original == "F" ~ "2",
           sex_original == "nk" ~ "UNK"
         )) %>% 
  select(SUBJECT_ID, CONSENT, SEX) 

#%>% filter(SUBJECT_ID %in% participants_to_delete)
```

3. Subject-Sample Mapping File
SUBJECT_ID, SAMPLE_ID

4. Dictionary for the Subject-Sample Mapping File
```{r}
desmith_subject_sample_DS <- desmith_clinical_source %>% 
  mutate(SUBJECT_ID = wpsid,
         SAMPLE_ID = paste0(wpsid, "_", analyte_type)) %>% 
  select(SUBJECT_ID, SAMPLE_ID)
```

```{r}
desmith_subject_consent_DS %>% 
  write_csv(here::here("output", "desmith_subject_consent_DS.csv"))

desmith_subject_sample_DS %>% 
  write_csv(here::here("output", "desmith_subject_sample_DS.csv"))
```



































