---
title: "Harmonize Khor demographic data to portal schema"
output: html_notebook
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```


Read in data:
```{r}
khor_data <- read_xlsx(synGet("syn32937504")$path)
```


## PARTICIPANT

Participant ID = External ID = BRI INCLUDE ID
```{r}
khor_transformed <- khor_data %>% 
  rename(`External ID` = `BRI INCLUDE ID`)
```

Columns that need to be transformed:

Sex = needs initial cap
Race
- White, Caucasian = White
- Black, African American = Black or African American
- American Indian, Alaska Native = American Indian or Alaska Native
- Do Not Know = Unknown
- NA = Unknown
- Other Race = Other
- everything with comma = More than one race
Ethnicity

```{r}
khor_transformed <- khor_transformed %>% 
  mutate(Gender = str_to_title(Gender)) %>% 
  mutate(Race = str_replace(Race, "White, Caucasian", "White"),
         Race = str_replace(Race, "Black, African American", "Black or African American"),
         Race = str_replace(Race, "American Indian, Alaska Native", "American Indian or Alaska Native"),
         Race = str_replace(Race, "Do Not Know", "Unknown"),
         Race = str_replace(Race, "Other Race", "Other")) %>% 
  mutate(Race = case_when(is.na(Race) ~ "Unknown",
                          str_detect(Race, ",") ~ "More than one race",
                          TRUE ~ Race)) %>% 
  rename(Ethnicity = `Ethnicity (hispanic/latino y/n)`) %>% 
  mutate(Ethnicity = case_when(
    Ethnicity == "yes" ~ "Hispanic or Latino",
    Ethnicity == "no" ~ "Not Hispanic or Latino",
    Ethnicity == "decline" ~ "Prefer not to answer",
    is.na(Ethnicity) ~ "Unknown"))
```


Final table:

- Rename/create the remaining empty cols needed to match DCA template
- Then select cols in order 

*** Use "Not provided" for empty columns where other datasets (e.g. HTP) have data - otherwise schematic ({pandas}) interprets cols with only "NA" as Double

```{r}
khor_participant <- khor_transformed %>% 
  mutate(`Component` = "Participant",
         `Study Code` = "BRI-DSR",
         `Participant ID` = `External ID`,
         `Family ID` = "Not provided",
         `Family Type` = "Proband-only",
         `Father ID` = "Not provided",
         `Mother ID` = "Not provided",
         `Family Relationship` = "Not provided",
         `Down Syndrome Status` = "T21",
         `Outcomes Vital Status` = "Not provided",
         `Age at Last Vital Status` = "Not provided") %>% 
  select(`Component`,
         `Study Code`,
         `Participant ID`,
         `External ID`,
         `Family ID`,
         `Family Type`,
         `Father ID`,
         `Mother ID`,
         `Family Relationship`,
         Sex = Gender,
         Race,
         Ethnicity,
         `Down Syndrome Status`,
         `Outcomes Vital Status`,
         `Age at Last Vital Status`)

```


```{r}
write_csv(khor_participant, here::here("output", "khor_participant.csv"))
```


##### CONDITION

Columns: 
Participant ID 
Condition Description 
Age at Condition Observation (Days) - blank for now 
Condition Interpretation - all "Observed" for now (ie. only positive assertions) 
Condition Source - all "Clinical" 
HPO Label 
HPO Code 
MONDO Label 
MONDO Code 
MAXO Label 
MAXO Code 
Other Label 
Other Code


Bring in previously annotated phenotypes csv downloaded from ETL Google sheet (https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit#gid=1006243162)

```{r}
pheno_lookup_khor <- read_csv(here::here("data", "pheno_codes_v2.csv")) %>% 
  filter(cohort == "Khor")
```

Add column "DS_Diagnosis" = Down syndrome
Pull conditions from clinical data
Look up annotations
```{r}
khor_conditions_subset <- khor_transformed %>% 
  mutate(ds_diagnosis = "Down syndrome") %>% 
  select(-c(2:6)) %>% 
  pivot_longer(cols = c(-`External ID`), 
               names_to = "Source Text",
               values_to = "Status") %>% 
  filter(!is.na(Status)) %>% 
  left_join(pheno_lookup_khor, by = c("Source Text" = "source_column"))
```

Create empty/all same value columns
Select columns in order of DCA template
```{r}
khor_condition <- khor_conditions_subset %>% 
  mutate(`Study Code` = "BRI-DSR",
         `Age at Condition Observation` = "Not provided",
         `Condition Interpretation` = "Observed",
         `Condition Source` = "Not provided") %>% 
  select(`Study Code`,
         `Participant ID` = `External ID`,
         `Condition Description` = display_text,
         `Age at Condition Observation`,
         `Condition Interpretation`,
         `Condition Source`,
         `HPO Label` = HPO_label,
         `HPO Code` = HPO_code,
         `MONDO Label` = MONDO_label,
         `MONDO Code` = MONDO_code,
         `MAXO Label` = MAXO_label,
         `MAXO Code` = MAXO_code,
         `Other Label` = Other_label,
         `Other Code` = Other_code)
```

```{r}
write_csv(khor_condition, here::here("output", "khor_condition.csv"))
```


##### BIOSPECIMEN - ADD AGES

Add ages from GEO series matrix to Meen's biospecimen/data file manifest

```{r}
meen_khor_manifest <- read_csv(here::here("data", "meen_khor_manifest.csv"))

geo_metadata <- as.matrix(read_delim(here::here("data", "khor_geo_metadata_partial.txt"), delim = "\t"))
```


```{r}
ages <- t(geo_metadata) %>% 
  as_tibble(rownames = "sample_id") %>% 
  janitor::row_to_names(1) %>% 
  select(1, 11) %>% 
  rename(c("Container ID" = 1, "age_at_draw" = 2)) %>% 
  mutate(age_at_draw = str_extract(age_at_draw, "[:digit:]+"))
  
```

```{r}
meen_khor_manifest_with_age <- meen_khor_manifest %>% 
  left_join(ages) %>% 
  mutate(`Age at Biospecimen Collection` = age_at_draw) %>% 
  select(-age_at_draw)
```

```{r}
write_csv(meen_khor_manifest_with_age, here::here("output", "meen_khor_manifest_with_age.csv"))
```







