---
title: "DS-DETERMINED Unharmonized Data Prep - CDM"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(janitor)

synLogin()

source(here::here("src/helpers/prep_unharmonized_data.R"))
```

# Read in data

- condition syn61348974
- death syn61348975
- demographic syn61348976
- diagnosis syn61348977
- encounter syn61348966
- immunization syn61348967
- lab_result_cm syn61348968
- med_admin syn61348969
- prescribing syn61348970
- pro_cm syn61348971
- procedures syn61348972
- vital syn61348973

```{r}
uh_data <- read_csv(synGet("syn61348973")$path)
uh_dict <- read_xlsx(synGet("syn61348956")$path, 
                     sheet = "FIELDS") %>% 
  filter(TABLE_NAME == "VITAL")
```

# Check for duplicate Participant IDs

- Condition: does not apply; multiple rows per participant
- Death: 0 (only 3 pts)
- Demopgrahic: 0
- Diagnosis: NA
- Encounter: NA
- Immunization: NA
- Lab Result: NA
- Med: NA
- Prescribing: NA
- Pro: NA
- Procedures: NA
- Vital: 

```{r}
uh_data %>% get_dupes(patid)
```

# Get fields that are in data but not in dictionary
--> request from contributor

- Condition: age_at_visit 
- Death: age_at_death
- Demographic: age_in_years
- Diagnosis: age_at_diagnosis
- Encounter: age_at_admit
- Immunization: age_at_vx_record
- Lab Result: lab_name (all NA), age_at_order
- Med Admin: age_at_med_started
- Prescribing: age_at_order
- Pro: pro_loinc, age_at_pro
- Procedures: age_at_procedure
- Vital: age_at_measure

```{r}
setdiff(colnames(uh_data), tolower(uh_dict$FIELD_NAME)) %>% 
  str_c(collapse = ", ")
```

# Get fields that are in dict but not in data
(no need to request from contributor unless any seem like oversights)

```{r}
setdiff(tolower(uh_dict$FIELD_NAME), colnames(uh_data))
```


# Get unique values from all columns
# Check for PHI
# compare enums/ranges to dictionary

- Condition: no PHI; enums ok
- Death: Ok
- Demog: no PHI; what does blank mean (as opposed to NI, UN, etc)
- Diagnosis: no PHI; pt SD-00635 has age -6
- Encounter: several negative age_at_admit; facilityid has names of clinics
(not PHI, flag for contributor but let them decide)
- Immunization: Flag vx_provider (first initial/last name)
- Lab Result: Ok
- Med: Ok
- Prescribing: Ok
- Pro: Ok
- Procedures: Ok
- Vital: Ok

```{r}
get_all_unique_values(uh_data) %>%
  write_csv(here::here("output", "dsd_cdm_vital_unique.csv"))
```

       
