---
title: "DS-DETERMINED Unharmonized Data Prep - SDI"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(janitor)

synLogin()

source(here::here("src/helpers/prep_unharmonized_data.R"))
```

# Read in data

```{r}
sdi_data <- read_csv(synGet("syn61348988")$path)
sdi_dict <- read_xlsx(synGet("syn61348987")$path)
```

# Check for duplicate Participant IDs

```{r}
sdi_data %>% get_dupes(record_id) #0
```

# Get fields that are in data but not in dictionary
--> request from contributor

```{r}
setdiff(colnames(sdi_data), tolower(sdi_dict$`Question identifier`)) %>% 
  str_c(collapse = ", ")
```

# Get fields that are in dict but not in data
- demog9 = city --> ok that it's removed

```{r}
setdiff(tolower(sdi_dict$`Question identifier`), colnames(sdi_data))
```

# In Excel, check free text fields for PHI

free text fields - no PHI found
- demog11_other
- demog12_other
- demog15a
- demog2_other
- demog5a
- demog6a
- demog16_other (not in dict)

# Get unique values from all columns and compare enums/ranges to dictionary

```{r}
get_all_unique_values(sdi_data) %>% 
  write_csv(here::here("output", "sdi_data_unique.csv"))
```


