---
title: "DS-DETERMINED Unharmonized Data Prep - DS-Connect"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(janitor)

synLogin()

source(here::here("src/helpers/prep_unharmonized_data.R"))
```

# Read in data

```{r}
dsc_data <- read_csv(synGet("syn61348981")$path)
dsc_dict <- read_xlsx(synGet("syn61348983")$path)
```

# Check for duplicate Participant IDs

```{r}
dsc_data %>% get_dupes(col4) #0
```

# Get fields that are in data but not in dictionary
--> request from contributor

```{r}
setdiff(colnames(dsc_data), tolower(dsc_dict$`column name`)) %>% 
  str_c(collapse = ", ")
```

# Get fields that are in dict but not in data

```{r}
setdiff(tolower(dsc_dict$`column name`), colnames(dsc_data)) #0
```


# Get unique values from all columns
# Check for PHI
# compare enums/ranges to dictionary

- no PHI found
- did not check all the enums/ranges bc there are so many cols

```{r}
get_all_unique_values(dsc_data) %>%
  write_csv(here::here("output", "dsd_dsc_data_unique.csv"))
```

# check if any of the un-consented DSC participants are in this DSD dataset

```{r}
dsc_withdrawn <- c(3191, 3362, 3419, 3427, 3574, 3638, 3675, 3678, 3754, 3768, 3798, 3838, 3971, 4012, 4188, 4215, 4239, 4310, 4377, 4393, 4453, 4559, 4606, 4628, 4659, 4681, 4761, 4834, 4875, 4960, 4964, 5004, 5018, 5093, 5140, 5175, 5202, 5264, 5294, 5322, 5391, 5433, 5539, 5585, 5670, 5794, 5869, 5892, 5932, 6015, 6057, 6248, 6295, 6314, 6416, 6468, 6500, 6510, 6695, 6779, 6923, 6993, 7007, 7217, 7296, 10485, 10552, 10563, 10608, 10724, 10855, 11318, 11507, 11655, 11832, 12284, 12308, 12529, 12688, 12856, 12857, 12881, 12887, 13026, 13046, 13194)

dsd_dsc <- read_xlsx(synGet("syn61348982")$path) %>% 
  pull(patient_id)
```


```{r}
length(dsd_dsc) #105
length(dsc_withdrawn) #86

intersect(dsd_dsc, dsc_withdrawn) #0

length(setdiff(dsc_withdrawn, dsd_dsc)) #86
length(setdiff(dsd_dsc, dsc_withdrawn)) #104 --> why??

# figuring out why
read_xlsx(synGet("syn61348982")$path) %>% 
  filter(patient_id %in% dsc_withdrawn)

read_xlsx(synGet("syn61348982")$path) %>% 
  get_dupes(patient_id) # 12849 -> SB-00034 and SB01930 :(((

read_xlsx(synGet("syn61348982")$path) %>% 
  get_dupes(subject_ID) #0
```


























