---
title: "Hakonarson VBR v6"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)
library(janitor)

synLogin()
```


## CHECKS

Get VBR files

```{r}
hakon_vbr <- read_xlsx(synGet("syn64314143")$path, sheet = "CAG_ALL DATA")
#hakon_aliquot_map <- read_csv(synGet("syn53355321")$path)


hakon_clin_dna <- read_xlsx(synGet("syn44291080")$path)
hakon_clin_rna <- read_xlsx(synGet("syn44291064")$path)

#delete the last few cols since they are for solid tumors
hakon_clinical_all <- hakon_clin_dna %>% 
  rbind(hakon_clin_rna) %>% 
  select(-(`Days to First Event`:`Tumor resection extent`))

```

Confirm:
- All sample IDs are unique
- Every Parent Sample ID is also a Sample ID?
- Every Sample ID has 1 Aliquot ID?

```{r}
n_distinct(hakon_vbr$`Sample External ID`) #6772
n_distinct(hakon_vbr$`Parent Sample External ID`) #2874
n_distinct(hakon_vbr$`Participant External ID`) #1152
n_distinct(hakon_clinical_all$`Participant ID`) #1152
n_distinct(hakon_vbr$`ALQ ID`) #6016
n_distinct(hakon_clinical_all$`Aliquot ID`) #1524

hakon_vbr %>% get_dupes(`Sample External ID`) #0

# Parent Sample IDs that are also Sample IDs
hakon_vbr %>% 
  filter(`Sample External ID` %in% `Parent Sample External ID`) %>% 
  pull(`Sample External ID`) %>% n_distinct() #1201

# Parent Sample IDs that are NOT Sample IDs
hakon_vbr %>% 
  filter(!`Parent Sample External ID` %in% `Sample External ID`) %>%
  pull(`Parent Sample External ID`) %>% n_distinct() #1673

# Sample ID-ALQ ID pairs
hakon_vbr %>% get_dupes(`Sample External ID`, `ALQ ID`) #0
hakon_vbr %>% get_dupes(`Parent Sample External ID`, `ALQ ID`) #1109
hakon_vbr %>% get_dupes(`ALQ ID`) %>% 
  #filter(str_detect(`Laboratory Procedure`, "pool"))
  pull(`Laboratory Procedure`) %>% unique()

# So eg. ALQ ID 40035746523 has 6 different Participant IDs & Sample IDs
# The samples with Lab Procedure == aliquoting/pooling have different Participant IDs
# Most of the duplicate ALQ IDs have different Sample IDs - are those pooled?
```

Check if Sample Type > Parent Sample Type > Collection Sample Type combinations make sense
- I think so? maybe save this for review by James
```{r}
hakon_vbr %>% 
  select(`Parent Sample Type`, `Sample Type`, `Laboratory Procedure`) %>% 
  unique() %>%
  arrange(`Parent Sample Type`, `Sample Type`) %>%  view()
```

Check Participant IDs in vbr vs clinical
```{r}
setdiff(hakon_vbr$`Participant External ID`, hakon_clinical_all$`Participant ID`) #0
setdiff(hakon_clinical_all$`Participant ID`, hakon_vbr$`Participant External ID`) #0

# All Pt IDs in vbr are in clinical data
hakon_vbr %>% 
  filter(!`Participant External ID` %in% hakon_clinical_all$`Participant ID`) #0
```


Check ALQ/Aliquot IDs in vbr vs clinical
```{r}
# 4 Aliquot IDs are not found among ALQ IDs
# 78953919 1235817282 1235817277   78944859
setdiff(hakon_clinical_all$`Aliquot ID`, hakon_vbr$`ALQ ID`)

hakon_vbr_filtered <- hakon_vbr %>% 
  select(`Participant External ID`, `ALQ ID`, `Sample Type`, `Collection Sample Type`) %>% 
  unique()

hakon_clinical_filtered <- hakon_clinical_all %>%
  select(`Participant ID`, `Aliquot ID`, `Analyte Type`, `Sample Composition`) %>% 
  unique() 

# 3 ALQ IDs have multiple Aliquot IDs
hakon_vbr_filtered %>% 
  left_join(hakon_clinical_filtered, by = c("Participant External ID" = "Participant ID")) %>% 
  select(`ALQ ID`, `Aliquot ID`) %>% 
  unique() %>% 
  get_dupes(`ALQ ID`) %>% 
  filter(dupe_count > 2)

#1394 Aliquot IDs have multiple ALQ IDs
hakon_vbr_filtered %>% 
  left_join(hakon_clinical_filtered, by = c("Participant External ID" = "Participant ID")) %>% 
  select(`ALQ ID`, `Aliquot ID`) %>% 
  unique() %>% 
  get_dupes(`Aliquot ID`) %>%
  #filter(`Aliquot ID` == "1202703232") %>% pull(`ALQ ID`) %>% unique()
  filter(dupe_count > 2) %>% 
  pull(`Aliquot ID`) %>% n_distinct()
```

```{r}


# Sample/Analyte Type match when ALQ/Aliquot ID match
hakon_vbr_filtered %>% 
  left_join(hakon_clinical_filtered, by = c("Participant External ID" = "Participant ID")) %>% 
  select(`Participant External ID`, `ALQ ID`, `Aliquot ID`, `Sample Type`, `Analyte Type`, `Collection Sample Type`, `Sample Composition`) %>% 
  filter(`ALQ ID` == `Aliquot ID`) %>% 
  filter(`Sample Type` != `Analyte Type`) 

# For matching ALQ/Aliquot pairs, 
# Most Collection Sample/Composition mismatches are Blood/PBMC,  which is fine
# but 3 aliquots/ALQ have Saliva vs PBMC
# 373942436, 373942405, 373942444
hakon_vbr_filtered %>% 
  left_join(hakon_clinical_filtered, by = c("Participant External ID" = "Participant ID")) %>% 
  select(`Participant External ID`, `ALQ ID`, `Aliquot ID`, `Sample Type`, `Analyte Type`, `Collection Sample Type`, `Sample Composition`) %>% 
  filter(`ALQ ID` == `Aliquot ID`) %>% 
  filter(`Collection Sample Type` != `Sample Composition`)  %>% 
  filter(`Collection Sample Type` == "Saliva")


```



=============================
### OLD v4 stuff

1/19/24: Adding in new aliquot mapping

Check aliquot mapping vs biospecimen catalog
```{r}
n_distinct(hakon_aliquot_map$participant_id) #1140

no_aliquot_map <- setdiff(hakon_vbr$`Participant External ID orig`, hakon_aliquot_map$participant_id)
```

There are 12 Participants that don't have aliquot mappings
```{r}
hakon_vbr %>% 
  left_join(hakon_aliquot_map,
            by = c("Participant External ID orig" = "participant_id")) %>% 
  filter(is.na(encrypted_container_id)) %>% 
  pull(`Participant External ID orig`) %>% unique()
```

Get clinical data for those Participants:
```{r}
hakon_clinical_all %>% 
  filter(`Participant ID` %in% no_aliquot_map) %>% 
  select(`Participant ID`, `Aliquot ID`, `Sample Composition`, `Analyte Type`) %>% 
  write_csv(here::here("output", "hakon_no_aliquot_mapping.csv"))
```


Excel
- make sure any blank Volume Unit or Concentration Unit doesn't also have a number
- if Sample Availability = No then Container Availability should also = No

Check that every Collection and Parent Sample is also a Sample
Check that all Container IDs are unique
```{r}
n_distinct(hakon_vbr$`Sample ID (Calc)`) #v5=4151 v7=4151
n_distinct(hakon_vbr$`Container ID`) #10855
n_distinct(hakon_vbr$`Parent Sample ID (Calc)`) #v5=4473 v7=3659
n_distinct(hakon_vbr$`Parent Sample Container ID`) #4473 #1588
n_distinct(hakon_vbr$`Collection ID`) #1152
n_distinct(hakon_vbr$`Collection Container ID`) #1588

```

Why is there >1 Container of a Collection?
```{r}
hakon_vbr %>% 
  select(`Collection ID`, `Collection Container ID`) %>%
  distinct() %>% 
 add_count(`Collection ID`) %>% 
  arrange(desc(n), `Collection ID`)
```

**2023-10-12:** Replace old v3 Sample IDs (e.g. aliquot_105366888_parent) 
with new Sample IDs from their VBR manifest (e.g. KFDS0451188158_Proband_DNA)
These will eventually be replaced with newer Sample IDs without spaces

```{r}
hakon_datafiles_v3 <- read_csv(synGet("syn52660168")$path)

hakon_dewrangle_specimens <- read_csv(here::here("data", "global-descriptors-new-x01-hakon_2023-10-12.csv")) %>% 
  filter(fhir_resource_type == "Specimen")
```

hakon biospecimens draft - get container IDs
```{r}
hakon_biospecimen_draft <- read_csv(here::here("output", "hakon_biospecimen_DRAFT.csv"))
```

```{r}
hakon_datafiles_v4 <- hakon_datafiles_v3 %>% 
  left_join(hakon_biospecimen_draft %>% 
              select(`Sample External ID`, `Sample Global ID`, 
                     `Container External ID`, `Container Global ID`) %>% 
              mutate(`Container External ID` = paste0("KFDS", `Container External ID`, sep = "")),
            by = "Sample External ID")

hakon_datafiles_v4 %>%
  left_join(hakon_vbr %>% select(`Sample ID (Calc)`, `Container ID`),
            by = c("Container External ID" = "Container ID")) %>% view()

```






