---
title: "Harmonize HTP VBR clinical data to LinkML v2 model - Aug 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()
```


## Updated clinical data to accompany VBR export
This includes data from all 700 v3 participants
Replace "NULL" and "Unknown" with `NA`
(TODO: Find out why there is NULL and Unknown in family member ID cols)

**TODO: Next time don't replace Unknown with NA in Race and Ethnicity**

```{r}
vbr_participants <- read_delim(synGet("syn52138217")$path, "\t") %>% 
  mutate(across(where(is.character), ~na_if(., "NULL"))) %>% 
  mutate(across(where(is.character), ~na_if(., "Unknown")))

all_cohorts_participant <- read_csv(synGet("syn51216460")$path)
htp_participants_v3 <- all_cohorts_participant %>% filter(`Study Code` == "HTP")
```

**Checks:**

Start by checking in Excel
- Participant ID - do any look wrong
- Family ID - do any look wrong; are any NA
- Family Type, Sex, Race, Ethnicity, First Event, Vital Status - 
are they spelled the same as before, any new values
- Father, Mother ID - usually just one; flag & ask if multiple
- Sibling ID - flag if there are a lot
- Other Family ID - should be filled in?
- Family Relationship - check if "Other" makes sense with Family Type;
all "Proband" should be T21
- DS Status - should not be Unknown
- Age at registration - should not be 0
- Age at last vital status - should not be 0, should be >= age at reg
- Generally do missing values (NULL/NA/Unknown/etc.) match previous versions

ISSUES:
- Relative IDs = difference between Unknown and NULL?
- Stop changing enums! e.g. family relationships, "D21-Only" vs "D21 Only"

```{r}
vbr_participants %>% janitor::get_dupes(ParticipantExternalID)

# 0 participants don't have minimal data AND don't have biospecimens AND not in v3
vbr_participants %>% 
  filter(DownSyndromeStatus == "Unknown") %>%
  filter(ParticipantExternalID %in% htp_participants_v3$`Participant External ID`)
filter(ParticipantExternalID %in% vbr_biospecimens$ParticipantExternalID)

# 0 don't have Family ID AND don't have samples AND not in v3
vbr_participants %>% 
  filter(is.na(FamilyID)) %>% 
  filter(ParticipantExternalID %in% htp_participants_v3$`Participant External ID`)
filter(ParticipantExternalID %in% vbr_biospecimens$ParticipantExternalID)

# check ages
vbr_participants %>%
  filter(AgeAtLastVitalStatusInDays < AgeAtFirstPatientEngagement)

# all participants in v3 data are in vbr_participants
setdiff(htp_participants_v3$`Participant External ID`, vbr_participants$ParticipantExternalID)

# 345 participants in vbr_participant not in v3 data
setdiff(vbr_participants$ParticipantExternalID, htp_participants_v3$`Participant External ID`)
```


Race, ethnicity, family type, family relationship

```{r}
htp_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "HTP",
         model_slot %in% c("race", "ethnicity", "familyType")) %>% 
  select(model_slot, model_enum_label, original_value) %>% 
  unnest(original_value)

vbr_participants_harmonized <- vbr_participants %>% 
  left_join(htp_etl_plan %>% filter(model_slot == "race") %>% 
              select(model_enum_label, original_value), 
            by = c("Race" = "original_value")) %>% 
  rename(race = model_enum_label) %>%
  relocate(race, .after = Race) %>% 
  left_join(htp_etl_plan %>% filter(model_slot == "ethnicity") %>% 
              select(model_enum_label, original_value), 
            by = c("Ethnicity" = "original_value")) %>% 
  rename(ethnicity = model_enum_label) %>% 
  relocate(ethnicity, .after = Ethnicity) %>% 
  left_join(htp_etl_plan %>% filter(model_slot == "familyType") %>% 
              select(model_enum_label, original_value), 
            by = c("FamilyType" = "original_value")) %>% 
  rename(familyType = model_enum_label) %>% 
  relocate(familyType, .after = FamilyType) %>%
  mutate(familyRelationship = case_when(
    FamilyRelationship == "Sister" ~ "Sibling",
    FamilyRelationship == "Brother" ~ "Sibling",
    FamilyRelationship == "Other" & familyType == "Control-only" ~ "Unrelated control",
    FamilyRelationship == "Other" & familyType != "Control-only" ~ "Other relative",
    TRUE ~ FamilyRelationship),
    .after = FamilyRelationship)

```


Check for non-D21 Families that have "Other" relatives,
Make sure Other Relative ID is filled in for them

Several Probands listed as Siblings or Other Relatives
e.g. F0876 - Proband HTP1123 is shown under "Other Relative"
F0061
Family Member IDs are listed for every member of family, not just Proband - is this ok?
```{r}
vbr_participants_harmonized %>% 
  filter(familyRelationship == "Other relative")

# other Probands listed as relatives?

probands <- vbr_participants_harmonized %>% 
  filter(familyRelationship == "Proband") %>% 
  pull(ParticipantExternalID)

vbr_participants_harmonized %>% 
  filter(OtherFamilyMemberID %in% probands)
```

## Add Global IDs

Check which participants need global IDs
```{r}
htp_dewrangle_patients <- read_csv(here::here("data", "global-descriptors-htp_2023-07-24.csv")) %>% 
  filter(fhir_resource_type == "Patient")

# check that the Global IDs in Matt's file are correct
vbr_participants %>% 
  select(ParticipantExternalID, Participant_global_id) %>% 
  left_join(htp_dewrangle_patients %>% select(global_id, descriptor),
            by = c("ParticipantExternalID" = "descriptor")) %>% 
  mutate(global_match = global_id == Participant_global_id) %>% 
  filter(global_match == FALSE)

# check overlap with existing dewrangle set
setdiff(vbr_participants$Participant_global_id, htp_dewrangle_patients$global_id) #NA
setdiff(htp_dewrangle_patients$global_id, vbr_participants$Participant_global_id) #24

# external IDs for 24 -> send to Matt
dewrangle_24_not_in_vbr_participants <- htp_dewrangle_patients %>% 
  filter(global_id %in% setdiff(htp_dewrangle_patients$global_id, 
                                vbr_participants$Participant_global_id)) %>% 
  pull(descriptor)

# are these patients in v3 portal? - Yes, all 24
htp_participants_v3 %>% 
  filter(`Participant External ID` %in% dewrangle_24_not_in_vbr_participants)

# Some of the 24 do have biospecimens
setdiff(dewrangle_24_not_in_vbr_participants, vbr_biospecimens$ParticipantExternalID)

```


Update vbr_participants with existing global IDs
```{r}
vbr_participants_harmonized_globalUpdated <- vbr_participants_harmonized %>% 
  left_join(htp_dewrangle_patients %>% select(global_id, descriptor),
            by = c("ParticipantExternalID" = "descriptor")) %>%
  rename(Participant_global_id_updated = global_id)

# check 355 needing globals against v3
vbr_participants_harmonized_globalUpdated %>% 
  filter(is.na(Participant_global_id_updated)) %>% 
  filter(ParticipantExternalID %in% htp_participants_v3$`Participant External ID`) #0
```

get new Global IDs from dewrangle
No changelog for participants because not changing any global IDs, just creating

```{r}
vbr_participant_dewrangle_input <- vbr_participants_harmonized_globalUpdated %>% 
  filter(is.na(Participant_global_id_updated)) %>% 
  transmute(descriptor = ParticipantExternalID, fhir_resource_type = "Patient")

write_csv(vbr_participant_dewrangle_input,
          here::here("output", "vbr_participant_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

htp_dewrangle_patients_UPDATED <- read_csv(here::here(
  "output", "global-descriptors-HTP_2023-09-07.csv")) %>% 
  filter(fhir_resource_type == "Patient")

vbr_participants_harmonized_globalUpdated <- vbr_participants_harmonized_globalUpdated %>% 
  left_join(htp_dewrangle_patients_UPDATED %>% select(global_id, descriptor),
            by = c("ParticipantExternalID" = "descriptor")) %>% 
  rename(Participant_global_id_updated_intermediate = Participant_global_id_updated) %>% 
  rename(Participant_global_id_final = global_id)

```

copy schema labels from https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit#gid=908117705
datapasta -> paste vertical vector -> toggle vector quotes

```{r}
vbr_participants_harmonized_final <- vbr_participants_harmonized_globalUpdated %>% 
  select(participantGlobalId = Participant_global_id_final,
         participantExternalId = ParticipantExternalID,
         familyId = FamilyID,
         familyType,
         fatherId = FatherID,
         motherId = MotherID,
         siblingId = SiblingIDs,
         otherFamilyMemberId = OtherFamilyMemberID,
         familyRelationship = FamilyRelationship,
         sex = Sex,
         race,
         ethnicity,
         downSyndromeStatus = DownSyndromeStatus,
         ageAtFirstPatientEngagement = AgeAtFirstPatientEngagement,
         outcomesVitalStatus = OutcomesVitalStatus,
         ageAtLastVitalStatus = AgeAtLastVitalStatusInDays) %>% 
  mutate(studyCode = "HTP",
         firstPatientEngagementEvent = "Enrollment") %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)

write_csv(vbr_participants_harmonized_final,
          here::here("output", "vbr_participants_harmonized_final.csv"))

```

## CONDITIONS

Get v3 conditions
```{r}
all_cohorts_condition <- read_csv(synGet("syn51216464")$path)
htp_conditions_v3 <- all_cohorts_condition %>% filter(`Study Code` == "HTP")

#355 participants don't have conditions yet - as expected
setdiff(vbr_participants_harmonized$ParticipantExternalID, htp_conditions_v3$`Participant External ID`)

#all participants in V3 Conditions are in vbr_participant
setdiff(htp_conditions_v3$`Participant External ID`, vbr_participants_harmonized$ParticipantExternalID)
```

Add DS MONDO code for new T21 participants and join to v3 conditions

```{r}
vbr_participants_conditions <- vbr_participants_harmonized_final %>% 
  filter(!`Participant External ID` %in% htp_conditions_v3$`Participant External ID`) %>% 
  filter(`Down Syndrome Status` == "T21") %>% 
  select(`Study Code`,
         `Participant External ID`) %>% 
  mutate(`Event ID` = NA,
         `Event Type` = NA,
         `Condition or Measure Source Text` = "Down syndrome",
         `Age At Condition or Measure Observation` = NA,
         `Condition Interpretation` = "Observed",
         `Condition Status` = "History Of", 
         `Condition Data Source` = "Clinical",
         `HPO Label` = NA,
         `HPO Code` = NA,
         `MONDO Label` = "Down syndrome",
         `MONDO Code` = "MONDO:0008608",
         `MAXO Label` = NA,
         `MAXO Code` = NA,
         `Other Label` = NA,
         `Other Code` = NA,
         `Measure Value` = NA,
         `Measure Unit` = NA) %>%
  select(all_of(condition_displaynames)) %>% 
  bind_rows(htp_conditions_v3)

write_csv(vbr_participants_conditions,
          here::here("output", "vbr_participants_conditions.csv"))
```

## QA CHECKS

Convert NAs in Race and Ethnicity back to Unknown
Convert all other NAs to blanks when writing to csv

```{r}
vbr_participants_harmonized_final %>% 
  mutate(Race = replace_na(Race, "Unknown"),
         Ethnicity = replace_na(Ethnicity, "Unknown")) %>% 
  write_csv(here::here("output", "vbr_participants_harmonized_final_na_2023-11-07.csv"),
            na = "")
```

