---
title: "HTP VBR biospecimen manifest for v4"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()
```


## CHECKS

Get VBR files

```{r}
vbr_biospecimens <- read_delim(synGet("syn52138216")$path, "\t")
vbr_participants <- read_delim(synGet("syn52138217")$path, "\t")

all_cohorts_participant <- read_csv(synGet("syn51216460")$path)
htp_participants_v3 <- all_cohorts_participant %>% filter(`Study Code` == "HTP")

portal_dump <- read_delim(synGet("syn52290446")$path, "\t")

```

Get list of dewrangle Global IDs
```{r}
htp_dewrangle_specimens <- read_csv(here::here("data", "global-descriptors-htp_2023-07-24.csv")) %>% 
  filter(fhir_resource_type == "Specimen")
```

check for dupes
```{r}
vbr_biospecimens %>% count(ContainerID) %>% arrange(desc(n))
# no dupes except for NA

vbr_biospecimens %>% select(-HTP_comment) %>% distinct() %>% count(ContainerID) %>% arrange(desc(n))

```


```{r}
# 13285 samples in vbr data not in dewrangle -> due to naming changes?
length(setdiff(unique(vbr_biospecimens$SampleID), htp_dewrangle$descriptor))

# 5430 samples in dewrangle not in vbr data -> due to naming changes?
length(setdiff(htp_dewrangle$descriptor, vbr_biospecimens$SampleID))
```

Compare biospecimen vs clinical datasets

```{r}
# all Biospecimens have Participant data
setdiff(vbr_biospecimens$ParticipantExternalID, vbr_participants$ParticipantExternalID)

# 153 Participants don't have biospecimens
setdiff(vbr_participants$ParticipantExternalID, vbr_biospecimens$ParticipantExternalID)
```


## GLOBAL IDs


Mapping of old -> new global IDs and compare to previous dewrangle output

various checks:

```{r}
htp_dewrangle_specimens # 6660

n_distinct(vbr_biospecimens$Participant_global_id) #677 - includes `NA`
n_distinct(portal_dump$`Participant ID`) #676
setdiff(vbr_biospecimens$Participant_global_id, portal_dump$`Participant ID`) #NA

n_distinct(portal_dump$`Sample ID`) #5405
n_distinct(vbr_biospecimens$Sample_global_id, na.rm = TRUE) #5335

unique_globals_in_portal <- unique(portal_dump$`Sample ID`)
unique_globals_in_vbr <- unique(vbr_biospecimens$Sample_global_id) #includes NA

globals_in_portal_not_in_vbr <- setdiff(globals_in_portal, global_in_vbr) #70 in portal, not in VBR
setdiff(global_in_vbr, globals_in_portal) #NA

portal_dump %>% filter(`Sample ID` %in% portal_not_in_vbr) %>%
  distinct() %>% arrange(`Sample Type`)
# 72 WBCs, 1 DNA, 1 CD4+, 1 CD8+
# 57 WBCs were deleted bc they don't have container IDs ("created" Parent Samples)
# the 3 others are known mismatches of Container IDs -> deleted from vbr
# not sure about the new WBCs since last update??

# make sure all sample IDs only have one sampleID_old
vbr_biospecimens %>% 
  select(SampleID, SampleID_old) %>%
  filter(!is.na(SampleID_old)) %>% 
  distinct() %>% 
  group_by(SampleID) %>% #or SampleID_old
  add_count() %>% 
  filter(n>1) %>% 
  arrange(SampleID)     #or SampleID_old

# get sample IDs without globals - 9226
# all of these samples also don't have SampleID_old
sample_without_globals <- vbr_biospecimens %>%
  select(SampleID, Sample_global_id, SampleID_old) %>%
  distinct() %>%
  filter(is.na(Sample_global_id)) %>% 
  pull(SampleID)

# check if they have matches in other rows
vbr_biospecimens %>% 
  filter(!is.na(Sample_global_id)) %>% 
  filter(SampleID %in% sample_without_globals) #0

# check if they have matches in dewrangle
sample_without_globals_but_in_dewrangle <- htp_dewrangle_specimens %>% 
  filter(descriptor %in% sample_without_globals) %>% 
  pull(descriptor) #27

# all have comment "NEW samples since portal v3 WITHOUT global IDs"
vbr_biospecimens %>% 
  filter(SampleID %in% sample_without_globals_but_in_dewrangle) %>% 
  pull(HTP_comment) %>% unique()

# none of the 27 (or any of the samples without globals) are in portal dump
portal_dump %>% 
  #filter(`Sample ID` %in% sample_without_globals_but_in_dewrangle)
  filter(`Sample ID` %in% sample_without_globals)

# check if old vs new sample types match for the 27
vbr_biospecimens %>% 
    filter(SampleID %in% sample_without_globals_but_in_dewrangle) %>% 
    select(SampleID, SampleID_old, SampleType, SampleType_old)


```

NOTE: in previous version (v8 on Synapse), found 46 samples with globals 
for both Saliva and DNA_sal -> deleted from v9 so we can move on; will deal with later

**UPDATE from Matt 9/28:**
Confirmed with biobank team that these are Saliva and not DNA_sal
I will ask Matt to delete DNA_sal from SampleID_old in biobank and associate correct
Global ID, but not send me that export until next update
```{r}
mismatched_46_sal_samples <- vbr_biospecimens_updated_participants %>% 
  left_join(htp_dewrangle_specimens %>% select(descriptor, global_id),
            by = c("SampleID" = "descriptor")) %>% 
  rename(dewrangle_globals_by_SampleID = global_id) %>% 
  filter(!is.na(dewrangle_globals_by_SampleID)) %>% 
  select(SampleID, SampleID_old, Sample_global_id, dewrangle_globals_by_SampleID) %>% 
  filter(Sample_global_id != dewrangle_globals_by_SampleID) %>% 
  distinct() %>% 
  left_join(htp_dewrangle_specimens %>% select(descriptor, global_id),
            by = c("dewrangle_globals_by_SampleID" = "global_id")) 

#%>% 
#  write_csv(here::here("output", "mismatched_46_sal_samples.csv"))

#46 samples have mismatch
# all DNA_sal

# Both HTP0087B_Saliva and HTP0087B_DNA_sal are in dewrangle and have different global IDs
# HTP0087B_Saliva	= bs-qrne3dqn9r
# HTP0087B_DNA_sal	bs-dkex39ktr6

#all in both cols are different
setdiff(mismatched_46_sal_samples$dewrangle_globals_by_SampleID, 
        mismatched_46_sal_samples$Sample_global_id) # & vice-versa

#are the Sample_global_id in dewrangle?
htp_dewrangle_specimens %>% 
  filter(global_id %in% mismatched_46_sal_samples$Sample_global_id)
  # the Sample_global_ids are associated with DNA_sal samples

htp_dewrangle_specimens %>% 
  filter(global_id %in% mismatched_46_sal_samples$dewrangle_globals_by_SampleID)
  # the dewrangle_globals are also all in dewrangle, associated with Saliva samples

# For reference, create table of dewrangle IDs
mismatched_46_sal_in_dewrangle <- htp_dewrangle_specimens %>% 
  filter(global_id %in% mismatched_46_sal_samples$dewrangle_globals_by_SampleID|
           global_id %in% mismatched_46_sal_samples$Sample_global_id) %>% 
  arrange(descriptor) %>% 
  select(-fhir_resource_type)

write_csv(mismatched_46_sal_in_dewrangle, here::here("output", "mismatched_46_sal_in_dewrangle.csv"))

# check which are in portal
setdiff(mismatched_46_sal_samples$Sample_global_id, portal_dump$`Sample ID`) #0
setdiff(mismatched_46_sal_samples$dewrangle_globals_by_SampleID, portal_dump$`Sample ID`) #46??

portal_dump %>% 
  filter(`Sample ID` %in% mismatched_46_sal_samples$dewrangle_globals_by_SampleID)
  # the Saliva samples in dewrangle are not in portal

portal_dump %>% 
  filter(`Sample ID` %in% mismatched_46_sal_samples$Sample_global_id)
  # the DNA_sal samples from dewrangle are in the portal


```


Start by updating Participant Global IDs from dewrangle
```{r}
vbr_biospecimens_updated_participants <- vbr_biospecimens %>% 
  left_join(htp_dewrangle_patients_UPDATED %>% select(descriptor, global_id),
            by = c("ParticipantExternalID" = "descriptor")) %>% 
  rename(Participant_global_id_updated = global_id)
  #filter(Participant_global_id != global_id)
```


Pull global IDs from dewrangle by SampleID and SampleID_old
```{r}
compare_vbr_dewrangle_globals <- vbr_biospecimens_updated_participants %>% 
  left_join(htp_dewrangle_specimens %>% select(descriptor, global_id),
            by = c("SampleID" = "descriptor")) %>% 
  rename(dewrangle_globals_by_SampleID = global_id) %>%
  left_join(htp_dewrangle_specimens %>% select(descriptor, global_id),
            by = c("SampleID_old" = "descriptor")) %>% 
  rename(dewrangle_globals_by_SampleIDold = global_id) %>% 
  select(SampleID, SampleID_old, Sample_global_id, 
         dewrangle_globals_by_SampleID, dewrangle_globals_by_SampleIDold) %>% 
  distinct() 

# filter(Sample_global_id != dewrangle_globals_by_SampleID) 
# filter(Sample_global_id != dewrangle_globals_by_SampleIDold)
# filter(dewrangle_globals_by_SampleID != dewrangle_globals_by_SampleIDold)
```

Checks:

```{r}
compare_vbr_dewrangle_globals %>% 
  filter(!is.na(Sample_global_id)) %>% 
  filter(is.na(dewrangle_globals_by_SampleID) & is.na(dewrangle_globals_by_SampleIDold)) %>% 
  left_join(htp_dewrangle_specimens %>% select(descriptor, global_id),
            by = c("Sample_global_id" = "global_id")) %>% 
  select(SampleID, SampleID_old, Sample_global_id, 
         dewrangle_descriptor_matching_Sample_global_id = descriptor) %>% 
  write_csv(here::here("output", "has_global_but_no_old_id_178.csv"))

# %>% pull(Sample_global_id) -> test
#portal_dump %>% filter(`Sample ID` %in% test)

# 178 samples have Sample_global_id but nothing pulled from dewrangle
# all of those globals exist in dewrangle -> just that SampleID_old was NA
# all are in Portal (under old sample ID)
# send to matt for review

compare_vbr_dewrangle_globals %>% 
  filter(is.na(Sample_global_id) & 
           (!is.na(dewrangle_globals_by_SampleID)|!is.na(dewrangle_globals_by_SampleIDold)))
# 27 Plasma samples also have only dewrangle_globals_by_SampleID, no SampleID_old


#do any samples have >1 row?
compare_vbr_dewrangle_globals %>% 
  filter(!is.na(SampleID_old)) %>% 
  janitor::get_dupes(SampleID) #0

```


Assuming the above 178 are ok - get a single final Global ID

```{r}
final_global_id_lookup <- compare_vbr_dewrangle_globals %>% 
  mutate(global_id_final = case_when(
    !is.na(Sample_global_id) ~ Sample_global_id,
    is.na(Sample_global_id) & !is.na(dewrangle_globals_by_SampleID) ~ dewrangle_globals_by_SampleID,
    is.na(Sample_global_id) & !is.na(dewrangle_globals_by_SampleIDold) ~ dewrangle_globals_by_SampleIDold
  )) %>% 
  filter(!is.na(global_id_final)) %>% 
  select(SampleID, global_id_final) %>% 
  distinct()



```


Update vbr biospecimens
```{r}
vbr_biospecimens_updated_samples <- vbr_biospecimens_updated_participants %>% 
  left_join(final_global_id_lookup, by = "SampleID")
```


CHECK: Of the samples that don't have any global IDs - they also don't have SampleID_old
-> are any of them erroneously missing SampleID_old? 
Not sure how to search for this
```{r}
no_global_ids_yet <- compare_vbr_dewrangle_globals %>% 
  mutate(global_id_final = case_when(
    !is.na(Sample_global_id) ~ Sample_global_id,
    is.na(Sample_global_id) & !is.na(dewrangle_globals_by_SampleID) ~ dewrangle_globals_by_SampleID,
    is.na(Sample_global_id) & !is.na(dewrangle_globals_by_SampleIDold) ~ dewrangle_globals_by_SampleIDold
  )) %>% 
  filter(is.na(global_id_final)) %>% 
  distinct() %>% 
  pull(SampleID)

# send filtered export to matt for review
vbr_biospecimens_updated_samples %>% 
  filter(SampleID %in% no_global_ids_yet) %>% 
  arrange(SampleID) %>% 
  write_csv(here::here("output", "for_review_no_global_ids_9199.csv"))

```


**NEW STUFF**: change of strategy - make sure all samples with files have Global IDs;
create new Globals for the rest of them

Copy/paste Christina's datafile manifests into htp_v1_v3_datafiles.xlsx
```{r}
htp_v1_v3_datafiles <- read_xlsx(here::here("data", "htp_v1_v3_datafiles.xlsx"))
```

Check v1/v3 datafile manifests from Christina
Do any of these samples not have global ID?

```{r}
htp_v1_v3_datafiles %>% 
  left_join(vbr_biospecimens_updated_samples %>% 
              select(SampleID, SampleID_old, global_id_final) %>% 
              distinct()) %>%
  filter(is.na(global_id_final))
# 1 sample, HTP0297A6_DNA has datafile but is not in vbr_biospecimens
# but is in dewrangle and v3 (bs-rbrtyrgb)

## Matt found that this should actually be HTP0297A_WhiteBloodCellsDNA 
## -> update global ID descriptor

```

<!-- Above 4 samples don't appear to be in vbr_biospecimens -->
<!-- But they are in dewrangle, and in portal dump with files associated: -->
<!-- - HTP0297A6_DNA = bs-rbrtyrgb -->
<!-- - HTP0298A_DNA_sal = bs-gexp663mcg -->
<!-- - HTP0303A_DNA_sal = bs-ffckciec6f -->
<!-- - HTP0306A_DNA_sal = bs-tti9rwdckw -->

<!-- HTP0297A6_DNA = bs-rbrtyrgb is correct in portal and in KFDS, need to figure out -->
<!-- why missing from Matt's VBR -->

<!-- Cristina will correct the other 3 DNA_sal samples in her source, regenerate manifest, -->
<!-- send back to me -->


Check Matt's 2021 X01 manifest
```{r}
vbr_x01_subset <- read_delim(synGet("syn52525144")$path, "\t")
```

2 samples (3 containers) don't have sampleID_old but do have global IDs:
HTP0471A_WhiteBloodCellsRNA
HTP0236A2_WhiteBloodCellsRNA
(Comment: NEW samples since portal v3 WITH global IDs)
-> don't worry about these, they just have 2 containers, 1 of which doesn't have 
sampleID_old but the other one does

the rest of the samples with no SampleID_old don't have global IDs 

```{r}
vbr_x01_subset %>% 
  #filter(str_detect(HTP_comment, "NEW")) %>%
  filter(str_detect(HTP_comment, "WITHOUT")) %>%
  select(SampleID, SampleID_old, Sample_global_id) %>% 
  distinct() %>% 
  left_join(vbr_biospecimens_updated_samples %>% 
              select(SampleID, SampleID_old, global_id_final) %>% 
              distinct(),
            by = "SampleID") %>% 
  filter(!is.na(global_id_final))

# I didn't find any other global IDs for those samples
```

Check the above X01 subset again with dewrangle and v3 portal
```{r}
vbr_x01_subset %>% 
  filter(str_detect(HTP_comment, "WITHOUT")) %>%
  select(SampleID, SampleID_old, Sample_global_id) %>% 
  distinct() %>% 
  left_join(htp_dewrangle_specimens %>% select(-fhir_resource_type),
            by = c("SampleID" = "descriptor")) %>% 
  filter(!is.na(global_id)) #0

vbr_x01_subset %>% 
  filter(str_detect(HTP_comment, "WITHOUT")) %>%
  select(SampleID, SampleID_old, Sample_global_id) %>% 
  distinct() %>% 
  left_join(portal_dump %>% select(`Sample ID`, `External Sample ID`),
            by = c("SampleID" = "External Sample ID")) %>% 
  filter(!is.na(`Sample ID`)) #0

```


check that globals in matt's export match mine

```{r}
vbr_x01_subset %>% 
  filter(!str_detect(HTP_comment, "WITHOUT")) %>%
  select(SampleID, SampleID_old, Sample_global_id) %>% 
  distinct() %>% 
  left_join(vbr_biospecimens_updated_samples %>% 
              select(SampleID, SampleID_old, global_id_final) %>% 
              distinct(),
            by = "SampleID") %>% 
  filter(Sample_global_id != global_id_final) #0
```

check if any vbr_biospecimen_updated_samples in x01 subset don't have global ID

```{r}
x01_with_global <- vbr_x01_subset %>% 
  select(SampleID, Sample_global_id) %>% 
  filter(!is.na(Sample_global_id)) %>% 
  distinct() %>% pull(SampleID)

vbr_biospecimens_updated_samples %>% 
  filter(SampleID %in% x01_with_global) %>% 
  filter(is.na(Sample_global_id)) #0

setdiff(vbr_x01_subset$Sample_global_id, vbr_biospecimens_updated_samples$global_id_final) #0

# check the 2 samples in Matt's x01 that are "new WITH global" - these match my global_id_final
vbr_biospecimens_updated_samples %>% 
  filter(SampleID %in% c("HTP0471A_WhiteBloodCellsRNA", "HTP0236A2_WhiteBloodCellsRNA")) %>% 
  select(SampleID, SampleID_old, Sample_global_id, global_id_final)
```


<!-- Final attempt to match more samples to dewrangle by putting spaces back in suffixes: -->
<!-- - get unique suffixes from vbr_biospecimens -->
<!-- - add spaces manually in excel, or match old dewrangle suffixes -->
<!-- - recreate "old" sample IDs -->
<!-- ```{r} -->
<!-- htp_dewrangle_specimens %>%  -->
<!--   separate(descriptor, into = c("LabID", "Suffix"), sep = "_", remove = FALSE) %>%  -->
<!--   select(Suffix) %>% distinct() -->

<!-- suffixes <- read_xlsx(here::here("data", "htp_vbr_suffixes.xlsx")) -->

<!-- vbr_biospecimens_updated_samples %>%  -->
<!--   filter(is.na(global_id_final)) %>% -->
<!--   select(SampleID) %>%  -->
<!--   distinct() %>%  -->
<!--   separate(SampleID, into = c("LabID", "Suffix"), sep = "_", remove = FALSE) %>%  -->
<!--   left_join(suffixes) %>%  -->
<!--   unite("SampleID_old_recreated", c(LabID, Suffix_old), sep = "_") %>%  -->
<!--   select(SampleID, SampleID_old_recreated) %>%  -->
<!--   left_join(htp_dewrangle_specimens %>% select(-fhir_resource_type), -->
<!--             by = c("SampleID_old_recreated" = "descriptor")) %>%  -->
<!--   filter(!is.na(global_id)) %>%  -->
<!--   write_csv(here::here("output", "htp_vbr_global_ids_by_semi_manual_matching.csv")) -->
<!-- ``` -->

## Dewrangle:

Separately update HTP0297A6_DNA to HTP0297A_WhiteBloodCellsDNA descriptor for bs-rbrtyrgb
Then redo the above steps to get vbr_biospecimens_updated_samples with correct global_id_final
for HTP0297A_WhiteBloodCellsDNA
```{r}
tibble(descriptor = "HTP0297A_WhiteBloodCellsDNA",
       fhir_resource_type = "Specimen",
       global_id = "bs-rbrtyrgb") %>% 
  write_csv(here::here("output", "dewrangle_update_HTP0297A.csv"))

# import back only NEW descriptors, including HTP0297A_WhiteBloodCellsDNA
htp_dewrangle_specimens <- read_csv(here::here("data", "global-descriptors-htp_2023-09-28.csv")) %>% 
  filter(fhir_resource_type == "Specimen")
```


dewrangle input: need new IDs for all samples in vbr_updated that don't have one
- all v1/v3 samples with datafiles (except HTP0297A6, which Christina needs to correct) are in vbr
- all x01 samples are in vbr
```{r}
needs_global <- vbr_biospecimens_updated_samples %>% 
  filter(is.na(global_id_final)) %>%
  #no global_id_final, Sample_global_id, or SampleID_old
  pull(SampleID) %>% 
  unique() #9198

htp_dewrangle_specimens %>% 
  filter(descriptor %in% needs_global) #0

vbr_x01_subset %>% 
  filter(SampleID %in% needs_global) #212
```

Check against Matt's numbers:
Distinct HTP samples = 14,561
Distinct HTP samples WITH global ids = 5,335
Distinct HTP samples WITHOUT global ids = 9,226
Existing Sample global ids = 6,660
Matched Sample global ids already in HTP Biospecimens table = 5,335
Unmatched Sample global ids TO BE DEPRECATED = 1,325 

```{r}
nrow(vbr_biospecimens_updated_samples) 
#41504

n_distinct(vbr_biospecimens_updated_samples$SampleID) 
#14515 [-46 saliva samples?]

n_distinct(vbr_biospecimens_updated_samples$global_id_final, na.rm = TRUE) 
#5317

vbr_biospecimens_updated_samples %>% 
  select(SampleID, global_id_final) %>% 
  filter(is.na(global_id_final)) %>% 
  unique() 
#9198

n_distinct(htp_dewrangle_specimens$global_id) 
#6660

htp_dewrangle_specimens %>% 
  filter(!global_id %in% vbr_biospecimens_updated_samples$global_id_final) 
#1343

```

Compare my list of globals to be deprecated with Matt's
```{r}
matt_deprecate_list <- read_delim(synGet("syn52570426")$path, "\t")

pierrette_deprecate_list <- htp_dewrangle_specimens %>% 
  filter(!global_id %in% vbr_biospecimens_updated_samples$global_id_final)

setdiff(pierrette_deprecate_list$descriptor, matt_deprecate_list$descriptor_orig)
# the 46 saliva samples!
```

dewrangle input:


```{r}
vbr_biospecimens_updated_samples %>% 
  select(SampleID, global_id_final) %>% 
  filter(is.na(global_id_final)) %>% 
  distinct() %>% 
  select(descriptor = SampleID) %>% 
  mutate(fhir_resource_type = "Specimen") %>% 
  write_csv(here::here("output", "vbr_dewrangle_input_2023-09-28.csv"))

# submit input to dewrangle; download only new descriptors

htp_dewrangle_specimens_updated <- read_csv(here::here("data", "global-descriptors-htp-new_2023-09-28.csv")) %>% 
  filter(fhir_resource_type == "Specimen")

```

add new globals to vbr_biospecimen
```{r}
vbr_biospecimens_updated_samples_20230928 <- vbr_biospecimens_updated_samples %>% 
  left_join(htp_dewrangle_specimens_updated %>% select(descriptor, global_id),
            by = c("SampleID" = "descriptor")) %>% 
  rename(global_id_new_20230928 = global_id) %>% 
  #filter(is.na(global_id_new_20230928) & is.na(global_id_final)) #0
  #filter(!is.na(global_id_final) & !is.na(global_id_new_20230928))  %>%
    #filter(global_id_final != global_id_new_20230928) #0
  mutate(global_id_final_20230928 = case_when(
    is.na(global_id_final) ~ global_id_new_20230928,
    TRUE ~ global_id_final
  )) 
  #filter(is.na(global_id_final_20230928))
  #filter(!is.na(global_id_final)) %>% filter(global_id_final != global_id_final_20230928)
  #filter(!is.na(global_id_new_20230928)) %>% filter(global_id_new_20230928 != global_id_final_20230928)
  #filter(!is.na(Sample_global_id)) %>% filter(Sample_global_id != global_id_final_20230928)
```

Get Container Global IDs
(Matt had already replaced -999999 ContainerIDs with NA)
```{r}
vbr_biospecimens_updated_samples_20230928 %>% 
  filter(ContainerID > 0) %>% 
  select(ContainerID) %>% 
  distinct() %>% 
  select(descriptor = ContainerID) %>% 
  mutate(fhir_resource_type = "Specimen") %>% 
  write_csv(here::here("output", "vbr_dewrangle_input_containers_2023-09-28.csv"))

# submit to dewrangle, download only new descriptors

htp_dewrangle_specimens_updated_containers <- read_csv(here::here("data", "global-descriptors-htp-new_containers_2023-09-28.csv")) %>% 
  filter(fhir_resource_type == "Specimen")

```

Add Parent, Collection & Container Globals
```{r}
vbr_biospecimens_updated_final <- vbr_biospecimens_updated_samples_20230928 %>%
  mutate(ContainerID_char = as.character(ContainerID)) %>% 
  left_join(htp_dewrangle_specimens_updated_containers %>% 
              select(descriptor, global_id),
            by = c("ContainerID_char" = "descriptor")) %>% 
  rename(containerGlobalId = global_id) %>%
  #filter(ContainerID>0) %>% filter(is.na(containerGlobalId)) #0
  left_join(htp_dewrangle_specimens_updated_containers %>% 
              select(descriptor, global_id),
            by = c("ParentSampleID" = "descriptor")) %>% 
  rename(parentSampleGlobalId = global_id) %>%
  #filter(is.na(parentSampleGlobalId)) %>% filter(ParentSampleID != "NULL")
  #HTP0748A_PAXgeneWholeBloodDNATube, HTP0755A_PAXgeneWholeBloodDNATube, HTP0760A_PAXgeneWholeBloodDNATube
  left_join(htp_dewrangle_specimens_updated_containers %>% 
              select(descriptor, global_id),
            by = c("CollectionID" = "descriptor")) %>% 
  rename(collectionGlobalId = global_id)
  #filter(is.na(collectionGlobalId)) #0
   

```

Parent Samples HTP0748A_PAXgeneWholeBloodDNATube, HTP0755A_PAXgeneWholeBloodDNATube, HTP0760A_PAXgeneWholeBloodDNATube didn't match parent sample global IDs in dewrangle
They all have Sample_global_id in Matt's original export
bs-afybb9mkp8
bs-tq8i7p8nbn
bs-aekbdanfp3
Sample_old is DNA_paxgene
These globals match DNA_paxgene samples in dewrangle

In X01 manifest:
HTP0748A_PAXgeneWholeBloodDNA
HTP0755A_PAXgeneWholeBloodDNA
HTP0760A_PAXgeneWholeBloodDNA

In dewrangle:
both DNA_paxgene and PAXgeneWholeBloodDNA for each
global IDs for DNA_paxgene match the PAXgeneWholeBloodDNATube

The PAXgeneWholeBloodDNA globals in dewrangle
bs-iyh48xa22e
bs-b2ejdfk7jh
bs-k94dfqy39z


##### HARMONIZATION

CLean up NULL vs N/A

```{r}
 vbr_biospecimens_harmonized <- vbr_biospecimens_updated_final %>% 
  select(studyCode = StudyCode,
         participantExternalId = ParticipantExternalID,
         participantGlobalId = Participant_global_id_updated,
         sampleExternalId = SampleID,
         sampleGlobalId = global_id_final_20230928,
         sampleType = SampleType,
         ageAtBiospecimenCollection = AgeInDaysAtBiospecimenCollection,
         parentSampleExternalId = ParentSampleID,
         parentSampleGlobalId,
         parentSampleType = ParentSampleType,
         collectionExternalId = CollectionID,
         collectionGlobalId,
         collectionSampleType = CollectionSampleType,
         containerExternalId = ContainerID,
         containerGlobalId,
         volume = Volume,
         volumeUnit = VolumeUnit,
         laboratoryProcedure = LaboratoryProcedure,
         biospecimenStorage = BiospecimenStorage,
         sampleAvailability = SampleAvailability,
         containerAvailability = containerAvailability) %>%
  #select_if(function(x) any(is.na(x)))
  #parentSampleGlobalId, containerExternalId, containerGlobalId
  mutate(across(where(is.character), ~na_if(., "NULL"))) %>% 
  mutate(across(where(is.character), ~na_if(., "N/A")))

vbr_biospecimens_harmonized %>% 
  write_csv(here::here("output", "vbr_biospecimens_harmonized_2023-09-28.csv"))

```

Check against Matt's numbers again:
Distinct HTP samples = 14,561
Distinct HTP samples WITH global ids = 5,335
Distinct HTP samples WITHOUT global ids = 9,226
Existing Sample global ids = 6,660
Matched Sample global ids already in HTP Biospecimens table = 5,335
Unmatched Sample global ids TO BE DEPRECATED = 1,325 

```{r}
nrow(vbr_biospecimens_harmonized) 
#41504

n_distinct(vbr_biospecimens_harmonized$sampleExternalId) 
#14515 [-46 saliva samples?]

n_distinct(vbr_biospecimens_harmonized$sampleExternalId, na.rm = TRUE) 
#14515

vbr_biospecimens_harmonized %>% 
  select(sampleExternalId, sampleGlobalId) %>% 
  filter(is.na(sampleGlobalId)) %>% 
  unique() 
#0

n_distinct(htp_dewrangle_specimens_updated$global_id) 
#15858 -> 53027 with containers

htp_dewrangle_specimens_updated %>% 
  filter(!global_id %in% vbr_biospecimens_harmonized$sampleGlobalId) 
#1343

```

## MANUAL FIXES TO LAST 3 SAMPLES

HTP0748A_PAXgeneWholeBloodDNA
HTP0748A_PAXgeneWholeBloodDNATube
HTP0755A_PAXgeneWholeBloodDNA
HTP0755A_PAXgeneWholeBloodDNATube
HTP0760A_PAXgeneWholeBloodDNA
HTP0760A_PAXgeneWholeBloodDNATube

I manually edited the VBR export and saved it as a new version on Synapse.
- swapped HTP_comment, SampleType_old, SampleID_old for DNA/DNATube samples
- Added Participant_global_id to all samples
- Filled in new Global IDs for DNATube samples created in dewrangle 
- Manually added HTP0297A6_DNA as SampleID_old for HTP0297A_WhiteBloodCellsDNA, 
and added Sample_global_id bs-rbrtyrgb


Saved HTP_DCC_VBR_BIOSPECIMENS_PL_corrected_3_DNATube_2023-10-04.csv
on Synapse as HTP_VBR_biospecimens v11 (syn52138216)
===
Marcel modified the dewrangle file - newest:
HTP0748A_PAXgeneWholeBloodDNA	        bs-iyh48xa22e
HTP0748A_PAXgeneWholeBloodDNATube	    bs-afybb9mkp8
HTP0755A_PAXgeneWholeBloodDNA	        bs-b2ejdfk7jh
HTP0755A_PAXgeneWholeBloodDNATube	    bs-tq8i7p8nbn
HTP0760A_PAXgeneWholeBloodDNA	        bs-k94dfqy39z
HTP0760A_PAXgeneWholeBloodDNATube	    bs-aekbdanfp3

DNA_paxgene is now the older descriptor for PAXgeneWholeBloodDNA. Unfortunately it
is different from the v3 portal where e.g. HTP0748A_DNA_paxgene was bs-afybb9mkp8,
but let's just move on with this

Saved global-descriptors-htp-newest_2023-10-05.csv on Synapse 
as global_ids_from_dewrangle v3 (syn52293082)
===
Manual modification of vbr_biospecimens_harmonized for corrected Global IDs:
just added Global IDs to Parent Sample Global ID col for DNATube samples

Check again:
```{r}
vbr_biospecimens_harmonized_20231005 <- read_csv(
  here::here("output", "vbr_biospecimens_harmonized_interim_2023-10-05.csv"))

htp_dewrangle_specimens_20231005 <- read_csv(here::here("data", "global-descriptors-htp-newest_2023-10-05.csv")) %>% 
  filter(fhir_resource_type == "Specimen")


nrow(vbr_biospecimens_harmonized_20231005) 
#41504

n_distinct(vbr_biospecimens_harmonized_20231005$sampleExternalId) 
#14515 [-46 saliva samples?]

n_distinct(vbr_biospecimens_harmonized_20231005$sampleExternalId, na.rm = TRUE) 
#14515

vbr_biospecimens_harmonized_20231005 %>% 
  select(sampleExternalId, sampleGlobalId) %>% 
  filter(is.na(sampleGlobalId)) %>% 
  unique() 
#0

n_distinct(htp_dewrangle_specimens_20231005$global_id) 
#15858 -> 53027 with containers

htp_dewrangle_specimens_20231005 %>% 
  filter(str_detect(descriptor, "HTP")) %>% #only samples, not containers
  filter(!global_id %in% vbr_biospecimens_harmonized_20231005$sampleGlobalId) %>% 
  distinct()
#1343 to be deprecated

```

Finalize colnames & write out
```{r}
vbr_biospecimens_harmonized_20231005 %>%
  mutate(concentration = NA,
         concentrationUnit = NA) %>% 
  select(all_of(biospecimen_schemalabels)) %>% 
  setNames(biospecimen_displaynames) %>%
  write_csv(here::here("output", "vbr_biospecimens_harmonized_2023-10-05.csv"))
```


Upload vbr_biospecimens_harmonized_2023-10-05.csv to Synapse
as htp_vbr_biospecimens_harmonized (syn52599895)


==========================

## DATAFILE MANIFESTS


Join harmonized data to v1/v3 datafile manifest to make sure sample IDs are up to date
```{r}
# corrected vbr export
vbr_export_20231004 <- read_delim(synGet("syn52138216")$path, "\t")
# v3 files 
v3_datafiles <- read_csv(synGet("syn52431530")$path)
# v1 files (need to check global IDs)
v1_datafiles <- read_csv(synGet("syn52431533")$path)
```

Compare v1 globals to vbr export
```{r}
v1_datafiles %>% 
  left_join(vbr_export_20231004 %>% 
              select(SampleID, SampleID_old, Sample_global_id) %>% 
              distinct(),
            by = c("sample_id" = "SampleID_old")) %>% 
  filter(global_id != Sample_global_id)
  #filter(sample_id != SampleID) #0
```

All globals & sample IDs in v1 manifest match vbr export 
Now check against latest dewrangle

```{r}
v1_datafiles %>%
  select(sample_id, global_id) %>% 
  left_join(htp_dewrangle_specimens_20231005 %>% select(descriptor, global_id),
            by = c("sample_id" = "descriptor")) %>% 
  filter(global_id.x != global_id.y) #0

v1_datafiles %>%
  select(sample_id, global_id) %>% 
  left_join(htp_dewrangle_specimens_20231005 %>% select(descriptor, global_id),
            by = "global_id") %>% 
  filter(sample_id != descriptor) #0

# all good!
```


Now check v3 files
```{r}
v3_datafiles %>% 
  select(`Sample External ID`) %>% 
  distinct() %>% 
  left_join(vbr_export_20231004 %>% 
              select(SampleID, SampleID_old, Sample_global_id) %>% 
              distinct(),
            by = c("Sample External ID" = "SampleID_old")) %>% 
  #filter(is.na(Sample_global_id))
  filter(`Sample External ID` != SampleID)

# all good after manual correction of 297A_DNA
```

Bring in New Sample IDs and Global IDs
```{r}
v3_datafiles %>% 
  rename(SampleID_old = `Sample External ID`) %>%
  left_join(vbr_export_20231004 %>% 
              select(SampleID, SampleID_old, Sample_global_id, Participant_global_id) %>% 
              distinct(),
            by = "SampleID_old") %>%
  relocate(Sample_global_id, SampleID, .before = SampleID_old) %>%
  relocate(Participant_global_id, .before = `Participant External ID`) %>% 
  #filter(is.na(Participant_global_id))
  select(`Study Code`, 
         `Participant Global ID` = Participant_global_id,
         `Participant External ID`,
         `Sample Global ID` = Sample_global_id,
         `Sample External ID` = SampleID,
         `File Global ID`,
         `File External ID`,
         `File Name`,
         `File Hash`,
         `Data Access`,
         `Data Category`,
         `Data Type`,
         `Experimental Strategy`,
         `Experimental Platform` = Platform,
         `File Format`,
         `File Size` = Size,
         `File Size Unit`,
         `Access URL`,
         `Publication DOI`) %>% 
  write_csv(here::here("output",
                       "FV_PJ79R4DY_FV_9Q2N4BTN_FV_7MVH1YCE_htp_datafile_manifest_w_crnic_files_v5.csv"))
  
```


Create File Global IDs for v1
Per Christina: use convention `include-study-us-east-1-prd-sd-7ydc1w4h/<file_id>`

```{r}
v1_datafiles %>% 
  select(descriptor = file_id) %>% 
  distinct() %>% 
  mutate(descriptor = paste0("include-study-us-east-1-prd-sd-7ydc1w4h/", descriptor),
         fhir_resource_type = "DocumentReference") %>% 
  write_csv(here::here("output",
                       "htp_dewrangle_input_v1_datafiles.csv"))

## Input to dewrangle

## Get output
htp_dewrangle_files_v1 <- read_csv(here::here("data", "global-descriptors-htp-newest_2023-10-10.csv")) %>% 
  filter(str_detect(descriptor, "include-study-us-east-1-prd-sd-7ydc1w4h/")) %>% 
  select(descriptor, file_global_id = global_id) %>% 
  mutate(descriptor = str_replace(descriptor, "include-study-us-east-1-prd-sd-7ydc1w4h/", ""))
```


Reformat v1 to match v3
Add Participant Global IDs
Add File Global IDs
```{r}
v1_datafiles %>%
  left_join(htp_dewrangle_files_v1,
            by = c("file_id" = "descriptor")) %>% 
 #filter(is.na(file_global_id))
  left_join(vbr_export_20231004 %>% 
              select(SampleID, Participant_global_id) %>% 
              distinct(),
            by = c("sample_id" = "SampleID")) %>%
  mutate(studyCode = "HTP",
         fileHash = NA,
         fileSizeUnit = "bytes",
         publicationDoi = NA) %>% 
  select(studyCode,
         participantGlobalId = Participant_global_id,
         participantExternalId = subject_id,
         sampleGlobalId = global_id,
         sampleExternalId = sample_id,
         fileGlobalId = file_global_id,
         fileExternalId = file_id,
         fileName = filename,
         fileHash,
         dataAccess = access_type,
         dataCategory = data_category,
         dataType = data_type,
         experimentalStrategy = experiment_strategy,
         experimentalPlatform = platform,
         fileFormat = file_format,
         fileSize = size_bytes,
         fileSizeUnit,
         accessUrl = drs_uri,
         publicationDoi) %>% 
  select(all_of(datafile_schemalabels)) %>% 
  setNames(datafile_displaynames) %>% 
  write_csv(here::here("output",
                       "htp_v1_omics_datafiles_2023-10-10.csv"))

 


```





































## OLD WORK

<!-- ## Get Unique Sample Types for enums -->

<!-- ```{r} -->
<!-- unique(c(vbr_biospecimens$SampleType,  -->
<!--          vbr_biospecimens$ParentSampleType, -->
<!--          vbr_biospecimens$CollectionSampleType)) %>% sort() -->
<!-- ``` -->

<!-- ## FIGURING OUT OLD VS NEW SAMPLE IDs -->

<!-- Get sample ID suffixes from dewrangle -->

<!-- ```{r} -->
<!-- htp_dewrangle %>%  -->
<!--   pull(descriptor) %>%  -->
<!--   str_replace("HTP.{5,7}_", "") %>%  -->
<!--   unique() -->
<!-- ``` -->

<!-- Get sample ID suffixes from VBR export -->

<!-- ```{r} -->
<!-- vbr_biospecimens %>%  -->
<!--   pull(SampleID) %>%  -->
<!--   str_replace("HTP.{5,7}_", "") %>%  -->
<!--   unique() -->
<!-- ``` -->

<!-- Read in Matt's lists of missing/unmatched samples -->

<!-- PART 1: 72 samples with --, 6 with NA or "Damaged BC"  -->
<!-- ```{r} -->
<!-- missing_containers <- read_delim(synGet("syn52293384")$path, "\t") %>%  -->
<!--   left_join(htp_dewrangle %>% select(global_id, descriptor), -->
<!--             by = c("Sample_global_id" = "global_id")) -->
<!-- ``` -->

<!-- look in aliquot.csv -->

<!-- ```{r} -->
<!-- aliquotcsv <- read_csv(synGet("syn52293019")$path) -->

<!-- aliquotcsv %>% filter(`Sample ID` %in% missing_containers$descriptor) -->
<!-- # these are all the Plasmas with NA or Damaged, so not helpful -->
<!-- ``` -->

<!-- compare v1 (htp_sample.csv) to v3 (FV_S71B1672_htp_samples_AND_COLLECTIONS_with_KF_global_ids_updated_2023-06-09.csv) -->

<!-- ```{r} -->
<!-- v1_sample_manifest <- read_csv(synGet("syn52293022")$path) #5494 -->
<!-- v3_sample_manifest <- read_csv(synGet("syn51514604")$path) #6430 -->
<!-- ``` -->
<!-- 161 samples in v1 aliquots.csv were not in v3  -->
<!-- -> Eric said 891 rows were dropped from aliquots.csv because of this -->
<!-- Per Matt, these should probably be deleted anyway, but he will check -->
<!-- ```{r} -->
<!-- #setdiff(v3_sample_manifest$`Sample External ID`, v1_sample_manifest$`Sample ID`) #1097 -->

<!-- v1_v3_setdiff <- setdiff(v1_sample_manifest$`Sample ID`, v3_sample_manifest$`Sample External ID`) #161 -->

<!-- htp_sample_manifest_robert %>%  -->
<!--   filter(`Sample ID` %in% v1_v3_setdiff) %>%  -->
<!--   write_csv(here::here("output", "v1_v3_setdiff.csv")) # sent to Matt -->
<!-- ``` -->


<!-- Look in v3 sample manifest -->
<!-- ```{r} -->
<!-- v3_sample_manifest %>% filter(`Sample External ID` %in% missing_containers$descriptor) -->

<!-- ``` -->

<!-- Look in portal dump -->
<!-- ```{r} -->
<!-- portal_dump <- read_tsv(synGet("syn52290446")$path) -->

<!-- missing_containers_that_have_files <- portal_dump %>%  -->
<!--   filter(`External Sample ID` %in% missing_containers$descriptor) %>%  -->
<!--   filter(Files != "--") %>%  -->
<!--   pull(`External Sample ID`) %>% unique() %>% sort() -->

<!-- #21 samples that don't have containers but do have files -->
<!-- ``` -->

<!-- Look in the HTP sample manifest file (Robert/Adam) -->
<!-- ```{r} -->
<!-- htp_sample_manifest_robert <- read_csv(synGet("syn48250502")$path) -->

<!-- htp_sample_manifest_robert %>%  -->
<!--   filter(`Sample ID` %in% missing_containers$descriptor) %>%  -->
<!--   pull(`Sample ID`) %>% unique() %>% sort() -->
<!-- # only the 6 Plasmas match Sample ID in Robert's sheet, and they don't have container IDs here either -->

<!-- # now look for those samples in Parent Sample column -->
<!-- htp_sample_manifest_robert %>%  -->
<!--   filter(`Parent Sample ID` %in% missing_containers$descriptor) %>% -->
<!--   pull(`Parent Sample ID`) %>% unique() -->
<!-- # 57, correct number -->
<!-- # these only appear as Parents and not Samples --> "created"? -->

<!-- ``` -->
<!-- From VBR Agenda 2023-05-30 -->

<!-- MISSING SAMPLES -->
<!-- 15 HTP samples with RNAseq in Data Service but not in portal  -->
<!-- These might have been in a small pilot - are there FASTQ files? Also public on GEO * check with Christina -->
<!-- HTP0084A2_RNA_paxgene -->
<!-- HTP0085B2_RNA_paxgene -->
<!-- HTP0204A_RNA_paxgene -->
<!-- HTP0206B_RNA_paxgene -->
<!-- HTP0288B_RNA_paxgene -->
<!-- HTP0302B_RNA_paxgene -->
<!-- HTP0310A_RNA_paxgene -->
<!-- HTP0322B_RNA_paxgene -->
<!-- HTP0326A_RNA_paxgene -->
<!-- HTP0328A_RNA_paxgene -->
<!-- HTP0332A_RNA_paxgene -->
<!-- HTP0338B_RNA_paxgene -->
<!-- HTP0348A_RNA_paxgene -->
<!-- HTP0351A_RNA_paxgene -->
<!-- HTP0442A2_RNA_paxgene -->


<!-- ```{r} -->
<!-- missing_15_samples <- c("HTP0084A2_RNA_paxgene", "HTP0085B2_RNA_paxgene", "HTP0204A_RNA_paxgene", "HTP0206B_RNA_paxgene", "HTP0288B_RNA_paxgene", "HTP0302B_RNA_paxgene", "HTP0310A_RNA_paxgene", "HTP0322B_RNA_paxgene", "HTP0326A_RNA_paxgene", "HTP0328A_RNA_paxgene", "HTP0332A_RNA_paxgene", "HTP0338B_RNA_paxgene", "HTP0348A_RNA_paxgene", "HTP0351A_RNA_paxgene", "HTP0442A2_RNA_paxgene") -->

<!-- setdiff(missing_containers_that_have_files, missing_15_samples) -->
<!-- ``` -->

<!-- PART 2: MISMATCHED CONTAINER IDs -->

<!-- ```{r} -->
<!-- unmatched_containers <- read_delim(synGet("syn52293382")$path, "\t") %>%  -->
<!--   left_join(htp_dewrangle %>% select(global_id, descriptor), -->
<!--             by = c("Sample_global_id" = "global_id")) -->
<!-- ``` -->
























