---
title: "Harmonize DS-Sleep clinical data to LinkML v2 model - March 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()
```

Read in data:

```{r}
dssleep_clinical_source <- read_xlsx(synGet("syn42114527")$path)
```

Replace headers with working_headers
See Google sheet `INCLUDE clinical data dictionaries` -> `DS-Sleep_demographics`
```{r}
gs4_deauth()

dssleep_working_headers <- read_sheet(
  "https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw/edit#gid=1550573864",
  sheet = "DS-Sleep_demographic")

dssleep_clinical_source <- setNames(dssleep_clinical_source, dssleep_working_headers$working_header)
```


Get ETL plan
```{r}
# source("harmonization_setup_2023-03.R")

dssleep_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "DS-Sleep")
```


##### PARTICIPANT

Remove Condition & Measure cols
```{r}
dssleep_participant <- dssleep_clinical_source %>% 
  select(-study_name, -ds_diagnosis,
         -age_at_first_recorded_study_event) %>% 
  rename(participantExternalId = participant_external_id)
```


Sex, Race (Ethnicity = Unknown):
```{r}
dssleep_race_lookup_values <- dssleep_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

dssleep_ethnicity_lookup_values <- dssleep_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

dssleep_familyrel_lookup_values <- dssleep_etl_plan %>% 
  filter(model_slot == "familyRelationship") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

dssleep_familytype_lookup_values <- dssleep_etl_plan %>% 
  filter(model_slot == "familyType") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

dssleep_ds_lookup_values <- dssleep_etl_plan %>% 
  filter(model_slot == "downSyndromeStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>% 
  mutate(original_value = na_if(original_value, "NA"))

dssleep_participant <- dssleep_participant %>%
  left_join(dssleep_race_lookup_values, by = c("race_original" = "original_value")) %>% 
  rename(race = model_enum_label) %>% 
  left_join(dssleep_ethnicity_lookup_values, by = c("ethnicity_original" = "original_value")) %>% 
  rename(ethnicity = model_enum_label) %>% 
  left_join(dssleep_familyrel_lookup_values, by = c("family_relationship" = "original_value")) %>% 
  rename(familyRelationship = model_enum_label) %>% 
  left_join(dssleep_familytype_lookup_values, by = c("family_type" = "original_value")) %>%
  rename(familyType = model_enum_label) %>% 
  left_join(dssleep_ds_lookup_values, by = c("ds_status" = "original_value")) %>%
  rename(downSyndromeStatus = model_enum_label)

#%>% select(participantExternalId, ethnicity_original, ethnicity, family_relationship, familyRelationship, family_type, familyType, ds_status, downSyndromeStatus)
```

Family:
- change "NA" and "N/A" in father_id and mother_id to `NA`
- family 113 has twin Probands -> add Sibling IDs manually

```{r}
dssleep_participant <- dssleep_participant %>% 
  mutate(father_id = na_if(father_id, "NA"),
         father_id = na_if(father_id, "N/A"),
         mother_id = na_if(mother_id, "NA"),
         mother_id = na_if(mother_id, "N/A"),
         sibling_id = case_when(participantExternalId == "RDoC_113_P1" ~ "RDoC_113_P2",
                                participantExternalId == "RDoC_113_P2" ~ "RDoC_113_P1"))
```


Get Global IDs:

```{r}
dssleep_dewrangle_input <- dssleep_participant %>% 
  distinct(participantExternalId) %>% 
  rename(descriptor = participantExternalId) %>% 
  mutate(fhir_resource_type = "patient")

write_csv(dssleep_dewrangle_input, 
          here::here("output", "dssleep_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

dssleep_global_ids <- read_csv(here::here("output", "global-descriptors-DS-Sleep.csv"))

dssleep_participant <- dssleep_participant %>% 
  left_join(dssleep_global_ids %>% select(descriptor, global_id),
            by = c("participantExternalId" = "descriptor")) 

#%>% select(participantExternalId, global_id) %>% filter(is.na(global_id))
```


Create final Participant table:

- Fill in info that is same for every participant (mutate)
- Rename columns that need renamed with schema labels (also mutate)
- select all Participant columns
- rename with display names

```{r}
dssleep_participant <- dssleep_participant %>% 
  mutate(studyCode = "DS-Sleep",
         participantGlobalId = global_id,
         familyId = family_id,
         fatherId = father_id,
         motherId = mother_id,
         sex = sex_original,
         siblingId = sibling_id,
         otherFamilyMemberId = NA,
         ageAtFirstPatientEngagement = age_at_registration_days,
         firstPatientEngagementEvent = "Enrollment",
         outcomesVitalStatus = NA,
         ageAtLastVitalStatus = NA) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)

##TODO: check against data
```

```{r}
write_csv(dssleep_participant, here::here("output", "dssleep_participant.csv"))
```


##### CONDITION

Get condition codes
```{r}
condition_lookup_dssleep <- condition_codes_v3 %>% filter(study == "DS-Sleep") %>% 
  unnest(source_column_value)
```


```{r}
dssleep_condition <- dssleep_clinical_source %>% 
  select(participantExternalId = participant_external_id, ds_diagnosis) %>%
  left_join(condition_lookup_dssleep, by = c("ds_diagnosis" = "source_column_value")) %>%
  filter(!is.na(conditionMeasureSourceText)) %>% 
  mutate(studyCode = study,
         eventId = NA,
         eventType = NA,
         ageAtConditionMeasureObservation = NA,
         conditionInterpretation = "Observed",
         conditionStatus = "History Of",
         conditionDataSource = "Clinical",
         measureValue = NA,
         measureUnit = NA) %>%
  select(all_of(condition_schemalabels)) %>% 
  setNames(condition_displaynames)
```


```{r}
write_csv(dssleep_condition, here::here("output", "dssleep_condition.csv"))
```
