---
title: "Harmonize Hakon to LinkML v2 model - March 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```


Read in data/Initial cleanup:

```{r}
hakon_clin_dna <- read_xlsx(synGet("syn44291080")$path)
hakon_clin_rna <- read_xlsx(synGet("syn44291064")$path)

#delete the last few cols since they are for solid tumors
hakon_clinical_source <- hakon_clin_dna %>% 
  rbind(hakon_clin_rna) %>% 
  select(-(`Days to First Event`:`Tumor resection extent`))
```

Replace headers with working_headers
See Google sheet `INCLUDE clinical data dictionaries` -> `X01-Hakon`
```{r}
gs4_deauth()

hakon_working_headers <- read_sheet(
  "https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw/edit#gid=0",
  sheet = "X01-Hakon")

hakon_clinical_source <- setNames(hakon_clinical_source, hakon_working_headers$working_header)
```

Get ETL plan
```{r}
# source("harmonization_setup_2023-03.R")

hakon_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "X01-Hakon")
```

Participant External ID
```{r}
hakon_clinical_harmonized <- hakon_clinical_source %>% 
  rename(participantExternalId = participant_id) 
```

## PARTICIPANT

Remove Condition/Sample cols and take distinct (many participants have RNA and DNA)
```{r}
hakon_participant <- hakon_clinical_harmonized %>% 
  select(participantExternalId, starts_with("family"), affected_status,
         age_at_sample_procurement_days, sex_original, race_original,
         ethnicity_original, vital_status, age_at_last_known_vital_status) %>% 
  distinct()
```

Cols that need transformed:
- Sex
- Race
- Ethnicity
- DS status

```{r}
hakon_sex_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "sex") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_race_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_eth_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_ds_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "downSyndromeStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_vital_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "outcomesVitalStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>% 
  mutate(original_value = na_if(original_value, "NA"))

hakon_participant <- hakon_participant %>% 
  left_join(hakon_sex_lookup_values, by = c("sex_original" = "original_value")) %>% 
  rename("sex" = model_enum_label) %>% 
  left_join(hakon_race_lookup_values, by = c("race_original" = "original_value")) %>% 
  rename("race" = model_enum_label) %>%  
  left_join(hakon_eth_lookup_values, by = c("ethnicity_original" = "original_value")) %>% 
  rename("ethnicity" = model_enum_label)%>%  
  left_join(hakon_ds_lookup_values, by = c("affected_status" = "original_value")) %>% 
  rename("downSyndromeStatus" = model_enum_label) %>% 
  left_join(hakon_vital_lookup_values, by = c("vital_status" = "original_value")) %>% 
  rename("outcomesVitalStatus" = model_enum_label) 

#%>% select(participantExternalId, sex_original, sex, race_original, race, ethnicity_original, ethnicity, affected_status, downSyndromeStatus, vital_status, outcomesVitalStatus)
```

Family relationships:
**NOTE: family KFDS1418766486 has 2 Probands - not indicated as sibs in source dataset**

```{r}
hakon_family_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "familyRelationship") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_participant <- hakon_participant %>% 
  left_join(hakon_family_lookup_values, by = c("family_relationship" = "original_value")) %>% 
  rename("familyRelationship" = model_enum_label) 

# %>% select(participantExternalId, family_relationship, familyRelationship)
```

Family member IDs - match familyRelationship with column for target ID
Do each relative type separately b/c chaining multiple unnests acts weird - drops rows 
```{r}
hakon_mother_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Mother") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  rename(motherId = Mother)

hakon_father_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Father") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  rename(fatherId = Father)  

hakon_sibling_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Sibling") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  unnest_wider(Sibling, names_sep = "") %>% 
  unite(col = "siblingId", starts_with("Sibling"), sep = ", ", na.rm = TRUE) %>% 
  # MANUALLY ADD family KFDS1418766486 Probands as sibs of each other
  add_row(proband_id = "KFDS8439296231", siblingId = "KFDS9616047740") %>% 
  add_row(proband_id = "KFDS9616047740", siblingId = "KFDS8439296231")

hakon_other_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Other relative") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  unnest_wider(`Other relative`, names_sep = "") %>% 
  unite(col = "otherFamilyMemberId", starts_with("Other"), sep = ", ", na.rm = TRUE)

hakon_participant <- hakon_participant %>% 
  arrange(family_id) %>% 
  left_join(hakon_mother_ids, by = c("participantExternalId" = "proband_id")) %>% 
  left_join(hakon_father_ids, by = c("participantExternalId" = "proband_id")) %>% 
  left_join(hakon_sibling_ids, by = c("participantExternalId" = "proband_id")) %>% 
  left_join(hakon_other_ids, by = c("participantExternalId" = "proband_id"))
```


Family Type:
- Proband-only: familyRel = Proband AND other ids = NA
- Duo: familyRel = Proband AND has motherId OR fatherId AND sib & other = NA
- Trio: familyRel = Proband AND has motherId AND fatherId ANd sib/other = NA
- Trio+: familyRel = Proband AND has motherId AND fatherId AND sib or other

find rows with 1 non-NA value: https://stackoverflow.com/a/73359219

```{r}
hakon_family_types <- hakon_participant %>% 
  filter(familyRelationship == "Proband") %>% 
  select(participantExternalId, motherId, fatherId, siblingId, otherFamilyMemberId) %>%
  mutate(family_count = rowSums(!is.na(across(c(motherId, fatherId, siblingId, otherFamilyMemberId))))) %>% 
  mutate(familyType = case_when(
    family_count == 0 ~ "Proband-only",
    family_count == 1 & (!is.na(motherId) | !is.na(fatherId)) ~ "Duo",
    family_count == 2 & (!is.na(motherId) & !is.na(fatherId)) ~ "Trio",
    family_count > 2 & (!is.na(motherId) & !is.na(fatherId)) ~ "Trio+",
    TRUE ~ "Other"
  ))

hakon_participant <- hakon_participant %>%
  left_join(hakon_family_types) %>% 
  arrange(family_id) %>% 
  group_by(family_id) %>% #fill familyType for non-Probands
  fill(familyType, .direction = "downup") %>% 
  ungroup()

#%>% select(participantExternalId, family_id, familyRelationship, familyType)

```

Age at first patient engagement:
Not collected for non-DS relatives
**consider making non-required in future data model**

```{r}
hakon_participant <- hakon_participant %>% 
  mutate(ageAtFirstPatientEngagement = case_when(
    downSyndromeStatus == "T21" ~ age_at_sample_procurement_days,
    downSyndromeStatus == "D21" ~ "Not collected"
  )) 

#%>% select(participantExternalId, downSyndromeStatus, age_at_sample_procurement_days, ageAtFirstPatientEngagement)
```



Get Global IDs:

```{r}
hakon_dewrangle_input <- hakon_participant %>% 
  distinct(participantExternalId) %>% 
  rename(descriptor = participantExternalId) %>% 
  mutate(fhir_resource_type = "patient")

write_csv(hakon_dewrangle_input, 
          here::here("output", "hakon_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

hakon_participant_global_ids <- read_csv(here::here("output", "global-descriptors-X01-Hakon.csv")) %>% 
  filter(fhir_resource_type == "Patient")

hakon_participant <- hakon_participant %>% 
  left_join(hakon_global_ids %>% select(descriptor, global_id),
            by = c("participantExternalId" = "descriptor")) 

#%>% select(participantExternalId, global_id) %>% filter(is.na(global_id))
```


Create final Participant table:

- Fill in info that is same for every participant (mutate)
- Rename columns that need renamed with schema labels (also mutate)
- select all Participant columns
- rename with display names

```{r}
hakon_participant <- hakon_participant %>% 
  mutate(studyCode = "X01-Hakon",
         participantGlobalId = global_id,
         familyId = family_id,
         firstPatientEngagementEvent = "Biospecimen collection",
         ageAtLastVitalStatus = as.numeric(age_at_last_known_vital_status)) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)

```

```{r}
write_csv(hakon_participant, here::here("output", "hakon_participant.csv"))
```


##### CONDITION

Get condition codes
```{r}
condition_lookup_hakon <- condition_codes_v3 %>% 
  filter(study == "X01-Hakon") %>% 
  mutate(source_column_value = as.character(source_column_value))
```


Pull conditions from clinical data
SPECIAL CASE: Replace "Respiratory failure; insufficiency; arrest" with "Respiratory failure, insufficiency, arrest"
```{r}
hakon_conditions <- hakon_clinical_harmonized %>% 
  select(participantExternalId, phenotypes_text) %>% 
  distinct() %>% #bc pts can have dna & rna
  mutate(phenotypes_text = str_replace(phenotypes_text, "Respiratory failure; insufficiency; arrest", "Respiratory failure, insufficiency, arrest")) %>% 
  separate_rows(phenotypes_text, sep = ";") %>% 
  mutate(phenotypes_text = str_trim(phenotypes_text)) %>%
  left_join(condition_lookup_hakon, by = c("phenotypes_text" = "source_column_value")) %>% 
  select(-phenotypes_text, -study, -source_column) %>%
  distinct() #remove rows with different phenotypes_text but same conditionSourceText

# %>% mutate(ontology_count = rowSums(!is.na(across(c(hpoCode, mondoCode, maxoCode, otherCode))))) 
# %>% arrange(ontology_count)

```

Add DS Diagnosis
```{r}
hakon_ds <- hakon_clinical_harmonized %>% 
  select(participantExternalId, diagnoses_text, age_at_diagnosis_days) %>%
  distinct() %>% 
  filter(!is.na(diagnoses_text)) %>% 
  left_join(condition_lookup_hakon, by = c("diagnoses_text" = "source_column_value")) %>% 
  select(-diagnoses_text, -study, -source_column)

hakon_conditions <- hakon_conditions %>% 
  bind_rows(hakon_ds) %>% arrange(participantExternalId)
```

Add age at condition observation (from Hakon, added directly to Condition table)
```{r}
hakon_condition_ages <- read_xlsx(synGet("syn51181708")$path) %>% 
  select(participant_external_id, condition_or_measure_source_text, age_at_condition_or_measure_observation) %>% 
  distinct()

hakon_conditions <- hakon_conditions %>% 
  left_join(hakon_condition_ages,
            by = c("participantExternalId" = "participant_external_id",
                   "conditionMeasureSourceText" = "condition_or_measure_source_text"))
```


Create empty/all same value columns
Select columns in order of DCA template

```{r}
hakon_condition <- hakon_conditions %>% 
  arrange(participantExternalId) %>% 
  mutate(studyCode = "X01-Hakon",
         eventId = NA,
         eventType = NA,
         ageAtConditionMeasureObservation = as.numeric(age_at_condition_or_measure_observation),
         conditionInterpretation = "Observed",
         conditionStatus = "History Of",
         conditionDataSource = "Clinical",
         measureValue = NA,
         measureUnit = NA) %>%
  select(all_of(condition_schemalabels)) %>% 
  setNames(condition_displaynames)
```

```{r}
write_csv(hakon_condition, here::here("output", "hakon_condition.csv"))
```


##### dbGaP SUBMISSION FILES

First, check overlap between Broad manifest & clinical data
```{r}
hakon_broad_wgs <- read_tsv("C:/Users/pierrettelo/Documents/INCLUDE/2021 X01s/FV_249MK9XS_hakon-include-wgs.tsv")

hakon_broad_rna <- read_tsv("C:/Users/pierrettelo/Documents/INCLUDE/2021 X01s/FV_2QD0Y620_hakon-include-rna.tsv")

n_distinct(hakon_broad_wgs$collaborator_participant_id) #1134
n_distinct(hakon_broad_rna$sample_alias) #379

hakon_wgs_participants <- hakon_broad_wgs %>% select(participant_id = collaborator_participant_id) %>%
  pull(participant_id)

hakon_rna_participants <- hakon_broad_rna %>% select(sample_alias) %>% 
  mutate(participant_id = str_extract(sample_alias, "^.*(?=_R1)")) %>% 
  pull(participant_id)

participants_with_any_seq <- unique(c(hakon_wgs_participants, hakon_rna_participants)) #1147

hakon_dna_clinical_participants <- hakon_clinical_source %>% 
  filter(analyte_type == "DNA") %>% 
  pull(participant_id) #1140

hakon_rna_clinical_participants <- hakon_clinical_source %>% 
  filter(analyte_type == "RNA") %>% 
  pull(participant_id) #384

participants_with_clin <- (c(hakon_dna_clinical_participants, hakon_rna_clinical_participants)) #1152

clin_but_no_seq <- setdiff(participants_with_clin, participants_with_any_seq) #5
# these patients have NO seq (DNA or RNA) -> delete from Subject-Sample DS, leave in Subject-Consent DS

```

Another check:
```{r}
# setdiff(hakon_dna_clinical_participants, hakon_wgs_participants) #6
# setdiff(hakon_wgs_participants, hakon_dna_clinical_participants) #0
# 
# setdiff(hakon_rna_clinical_participants, hakon_rna_participants) #5
# setdiff(hakon_rna_participants, hakon_rna_clinical_participants) #0

#see below - only 5 patients have NO seq (DNA or RNA) -> delete from dbGaP file
unique(c(setdiff(hakon_dna_clinical_participants, hakon_wgs_participants), setdiff(hakon_rna_clinical_participants, hakon_rna_participants)))
```
Of the above 11:

no DNA or RNA
"KFDS1508310361" 
"KFDS4451493159"
"KFDS0051582897"
"KFDS5396312025"
"KFDS0620027956"

RNA only
"KFDS6970433974" 

DNA only
"KFDS9949593004"
"KFDS7377609306"
"KFDS8324321410"
"KFDS1094210396" 
"KFDS0225323816"

=======================

1. Subject-Consent Mapping File - INCLUDE the 5 participants with no seq
SUBJECT_ID, CONSENT, SEX

2. Dictionary for the Subject-Consent Mapping File

```{r}
hakon_subject_consent_DS <- hakon_clinical_source %>%
  mutate(SUBJECT_ID = participant_id,
         CONSENT = "1",
         SEX = case_when(
           sex_original == "Male" ~ "1",
           sex_original == "Female" ~ "2",
           sex_original == "Not Reported" ~ "UNK"
         )) %>% 
  select(SUBJECT_ID, CONSENT, SEX) %>%
  distinct() %>% 
  #filter(!SUBJECT_ID %in% clin_but_no_seq) %>% 
  arrange(SUBJECT_ID) 

#%>% filter(SUBJECT_ID %in% clin_but_no_seq)
```

```{r}
hakon_subject_consent_DS %>%  
  write_csv(here::here("output", "hakon_subject_consent_DS.csv"))
```

3. Subject-Sample Mapping File
SUBJECT_ID, SAMPLE_ID

4. Dictionary for the Subject-Sample Mapping File
```{r}
hakon_subject_sample_DS <- hakon_clinical_source %>% 
  mutate(SUBJECT_ID = participant_id,
         SAMPLE_ID = as.character(aliquot_id)) %>% 
  select(SUBJECT_ID, SAMPLE_ID) %>% 
  distinct() %>%
  filter(!SUBJECT_ID %in% clin_but_no_seq) %>% 
  arrange(SUBJECT_ID)
```


```{r}
hakon_subject_sample_DS %>%  
  write_csv(here::here("output", "hakon_subject_sample_DS.csv"))
```


##### BIOSPECIMEN

"whole blood DNA from EDTA tube"

from Christina on Slack:
When we receive either sample_id or aliquot_id from the investigator (but not both),
we ingest the given specimen ID column into both external_sample_id and
external_aliquot_id . This is so we do not leave these fields as NULL in the dataservice
and to keep the consistency of “always pull from the aliquot ID column for dbgap files”
Again, this ID column will match what is in the Broad’s manifest.

##For now, leave Sample ID blank, use Aliquot IDs to create Global IDs

Blood = DNA
PBMC = RNA    [Do PBMCs come from the same draw as the Blood??]
Saliva = DNA

Leave in the participants with no seq in case we want the record for the VBR

Transform sample types to match what is in Portal currently

```{r}
hakon_biospecimen <- hakon_clinical_source %>% 
  select(participant_id, aliquot_id, sample_method_of_procurement, 
         age_at_sample_procurement_days, sample_composition, analyte_type) %>% 
  mutate(studyCode = "X01-Hakon",
         participantExternalId = participant_id,
         sampleGlobalId = NA,
         sampleExternalId = NA,
         sampleType = analyte_type,
         ageAtBiospecimenCollection = age_at_sample_procurement_days,
         parentSampleGlobalId = NA,
         parentSampleExternalId = NA,
         parentSampleType = sample_composition,
         collectionGlobalId = NA,
         collectionExternalId = NA,
         collectionSampleType = sample_method_of_procurement,
         containerGlobalId = NA,
         containerExternalId = aliquot_id,
         volume = NA,
         volumeUnit = NA,
         concentration = NA,
         concentrationUnit = NA,
         laboratoryProcedure = NA,
         biospecimenStorage = NA,
         sampleAvailability = "TBD") %>%
  mutate(parentSampleType = case_when(
           parentSampleType == "PBMC" ~ "PBMCs",
           parentSampleType == "Blood" ~ "Peripheral Whole Blood",
           parentSampleType == "Saliva" ~ "Saliva"),
         collectionSampleType = case_when(
           collectionSampleType == "Blood Draw" ~ "Peripheral Whole Blood",
           collectionSampleType == "Saliva Kit" ~ "Saliva"
         )) %>% 
  select(all_of(biospecimen_schemalabels)) %>% 
  setNames(biospecimen_displaynames)

```

Get Global IDs for Aliquot IDs:

```{r}
hakon_aliquot_dewrangle_input <- hakon_clinical_source %>% 
  select(descriptor = aliquot_id) %>% 
  mutate(fhir_resource_type = "specimen") %>% 
  distinct()

write_csv(hakon_aliquot_dewrangle_input, 
          here::here("output", "hakon_aliquot_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

hakon_aliquot_global_ids <- read_csv(here::here("output", "global-descriptors-X01-Hakon.csv")) %>% 
  filter(fhir_resource_type == "Specimen" & !str_detect(descriptor, "aliquot"))

hakon_biospecimen <- hakon_biospecimen %>%
  mutate(`Container External ID` = as.character(`Container External ID`)) %>% 
  left_join(hakon_aliquot_global_ids %>% select(descriptor, global_id),
            by = c("Container External ID" = "descriptor")) %>% 
  mutate(`Container Global ID` = global_id) %>% 
  select(-global_id)
```


#### Make up Sample External IDs for seq pipeline
Christina can't use Container IDs, so I need to add placeholder Sample IDs
Get Global IDs
Later - get actual Sample External IDs from Mateusz

```{r}
hakon_biospecimen <- hakon_biospecimen %>% 
  mutate(`Sample External ID` = paste0("aliquot_", `Container External ID`, "_parent"))

hakon_sample_dewrangle_input <- hakon_biospecimen %>% 
  select(descriptor = `Sample External ID`) %>% 
  mutate(fhir_resource_type = "specimen")

write_csv(hakon_sample_dewrangle_input, here::here("output", "hakon_sample_dewrangle_input.csv"))

hakon_sample_global_ids <- read_csv(here::here("output", "global-descriptors-X01-Hakon.csv")) %>% 
  filter(str_detect(descriptor, "parent")) %>% 
  select(descriptor, global_id)

hakon_biospecimen <- hakon_biospecimen %>% 
  left_join(hakon_sample_global_ids, by = c(`Sample External ID` = "descriptor")) %>% 
  mutate(`Sample Global ID` = global_id) %>% 
  select(-global_id)

write_csv(hakon_biospecimen, here::here("output", "hakon_biospecimen_DRAFT.csv"))

```


#### Modify Participant manifest for Christina
Seq pipeline needs specific family relationships -> replace "Other relative" with
original relationship from Hakon

Christina filtered all_cohorts_participant and uploaded to Data Tracker,
then I downloaded for this
```{r}
hakon_participant_for_seq <- read_csv(here::here("data", "FV_4D2JPH5S_hakon_participant_manifest.csv"))
```

```{r}
hakon_participant_for_seq %>% 
  left_join(hakon_clinical_source %>% 
              select(participant_id, family_relationship) %>% 
              distinct(),
            by = c(`Participant External ID` = "participant_id")) %>% 
  rename(original_family_relationship = family_relationship) %>% 
  relocate(original_family_relationship, .after = `Family Relationship`) %>% 
  write_csv(here::here("output", "hakon_participant_for_seq.csv"))
```






