
```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()

gs4_deauth()
```


**Copy this file for each new dataset**

Get dataset

```{r}
clinical_data <- read_xlsx(synGet("syn53058587")$path)
```


In Excel, identify columns that should be numeric but aren't - list below:

- Height (inches)
- Weight (lbs)
- BMI

For the above cols, replace NA value: "missing"
And convert to numeric

Then select all numeric columns and get ranges (min-max)

```{r}
clinical_data_converted <- clinical_data %>% 
  mutate(`Height (inches)` = as.numeric(na_if(`Height (inches)`, "missing")), 
         `Weight (lbs)` = as.numeric(na_if(`Weight (lbs)`, "missing")), 
         BMI = as.numeric(na_if(BMI, "missing")))

num_entries <- clinical_data_converted %>% 
  select(where(is.numeric))  %>% 
  broom::tidy(summary()) %>% 
  select(original_header = column, min, max) %>% 
  unite(min, max, col = "values", sep = "-")

```

Get unique values for character columns

```{r}
clin_entries <- clinical_data_converted %>% 
  select(where(is.character)) %>% 
  pivot_longer(cols = everything(),
               names_to = "original_header",
               values_to = "all_values") %>% 
  group_by(original_header) %>% 
  summarize(values = paste(unique(all_values), collapse = "; "))
```


Assemble dictionary

```{r}
all_entries <- bind_rows(num_entries, clin_entries)

dict <- tibble("original_header" = colnames(clinical_data),
       "working_header" = NA) %>% 
  left_join(all_entries)
```

Write out and paste into Google sheet 
(https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw)

In Google sheet, add original NA value to numeric ranges where NA was replaced 
(e.g. "36-69, missing")

```{r}
write_csv(dict, here::here("output", "dict.csv"))
```


Get "long" version for ETL sheet

```{r}
conditions_long <- clinical_data_converted %>% 
  select(where(is.character)) %>% 
  pivot_longer(cols = everything(),
               names_to = "original_header",
               values_to = "all_values") %>% 
  group_by(original_header) %>% 
  summarize(unique_values = unique(all_values)) %>% 
  arrange(unique_values, .by_group = TRUE)
```






