---
title: "Hakonarson VBR v4"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()
```


## CHECKS

Get VBR files

```{r}
hakon_vbr <- read_csv(synGet("syn52504777")$path)

hakon_clin_dna <- read_xlsx(synGet("syn44291080")$path)
hakon_clin_rna <- read_xlsx(synGet("syn44291064")$path)

#delete the last few cols since they are for solid tumors
hakon_clinical_all <- hakon_clin_dna %>% 
  rbind(hakon_clin_rna) %>% 
  select(-(`Days to First Event`:`Tumor resection extent`))

```


Confirm all samples have clinical data
```{r}
nrow(hakon_vbr) #old version v5 = 18794 new version v7 = 17137
n_distinct(hakon_vbr$`Participant External ID orig`) #1152
n_distinct(hakon_clinical_all$`Participant ID`) #1152

setdiff(hakon_vbr$`Participant External ID orig`, hakon_clinical_all$`Participant ID`)
setdiff(hakon_clinical_all$`Participant ID`, hakon_vbr$`Participant External ID orig`)

hakon_vbr %>% 
  filter(`Participant External ID orig` %in% hakon_clinical_all$`Participant ID`)
```

Excel
- make sure any blank Volume Unit or Concentration Unit doesn't also have a number
- if Sample Availability = No then Container Availability should also = No

Check that every Collection and Parent Sample is also a Sample
Check that all Container IDs are unique
```{r}
n_distinct(hakon_vbr$`Sample ID (Calc)`) #v5=4151 v7=4151
n_distinct(hakon_vbr$`Container ID`) #10855
n_distinct(hakon_vbr$`Parent Sample ID (Calc)`) #v5=4473 v7=3659
n_distinct(hakon_vbr$`Parent Sample Container ID`) #4473 #1588
n_distinct(hakon_vbr$`Collection ID`) #1152
n_distinct(hakon_vbr$`Collection Container ID`) #1588

```

Why is there >1 Container of a Collection?
```{r}
hakon_vbr %>% 
  select(`Collection ID`, `Collection Container ID`) %>%
  distinct() %>% 
 count(`Collection ID`) %>% 
  arrange(desc(n))
```

**2023-10-12:** Replace old v3 Sample IDs (e.g. aliquot_105366888_parent) 
with new Sample IDs from their VBR manifest (e.g. KFDS0451188158_Proband_DNA)
These will eventually be replaced with newer Sample IDs without spaces

```{r}
hakon_datafiles_v3 <- read_csv(synGet("syn52660168")$path)

hakon_dewrangle_specimens <- read_csv(here::here("data", "global-descriptors-new-x01-hakon_2023-10-12.csv")) %>% 
  filter(fhir_resource_type == "Specimen")
```

hakon biospecimens draft - get container IDs
```{r}
hakon_biospecimen_draft <- read_csv(here::here("output", "hakon_biospecimen_DRAFT.csv"))
```

```{r}
hakon_datafiles_v4 <- hakon_datafiles_v3 %>% 
  left_join(hakon_biospecimen_draft %>% 
              select(`Sample External ID`, `Sample Global ID`, 
                     `Container External ID`, `Container Global ID`) %>% 
              mutate(`Container External ID` = paste0("KFDS", `Container External ID`, sep = "")),
            by = "Sample External ID")

hakon_datafiles_v4 %>%
  left_join(hakon_vbr %>% select(`Sample ID (Calc)`, `Container ID`),
            by = c("Container External ID" = "Container ID")) %>% view()

```


TODO: Aliquot IDs != Container IDs!! :(






