---
title: "Harmonize HTP clinical data to LinkML v2 model - March 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()
```

Read in data:

NOTE: BMI, Weight, Height have "NA" strings
OfficialDSDiagnosis has blanks (`NA`) 

```{r}
htp_clinical_source <- read_xlsx(synGet("syn25574963")$path)
```

Replace headers with working_headers
See Google sheet `INCLUDE clinical data dictionaries` -> `HTP`
```{r}
gs4_deauth()

htp_working_headers <- read_sheet(
  "https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw/edit#gid=0",
  sheet = "HTP")

htp_clinical_source <- setNames(htp_clinical_source, htp_working_headers$working_header)
```

Get ETL plan
```{r}
# source("harmonization_setup_2023-03.R")

htp_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "HTP")
```

Add Participant IDs: 
```{r}
htp_clinical_harmonized <- htp_clinical_source %>% 
  mutate(participantExternalId = 
           str_extract(lab_id, "HTP[:digit:]{4}"),
         .before = everything())
```


##### PARTICIPANT

Remove Condition cols & anything that changes by visit
Remove duplicate rows caused by multiple visits
```{r}
htp_participant <- htp_clinical_harmonized %>% 
  select(participantExternalId, sex_original, age_at_enrollment_days, family_id, 
         karyotype, relationship, ds_diagnosis, race_original, ethnicity_original, 
         official_ds_diagnosis) %>% 
  distinct()
```


Sex: No changes needed

Race, Ethnicity, DS Status:
```{r}
htp_race_lookup_values <- htp_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

htp_eth_lookup_values <- htp_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

htp_ds_lookup_values <- htp_etl_plan %>% 
  filter(model_slot == "downSyndromeStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

htp_participant <- htp_participant %>% 
  left_join(htp_race_lookup_values, by = c("race_original" = "original_value")) %>% 
  rename(race = model_enum_label) %>% 
  left_join(htp_eth_lookup_values, by = c("ethnicity_original" = "original_value")) %>% 
  rename(ethnicity = model_enum_label) %>% 
  left_join(htp_ds_lookup_values, by = c("karyotype" = "original_value")) %>% 
  rename(downSyndromeStatus = model_enum_label) 

# check:
# %>% select(participantExternalId, race_original, race, ethnicity_original, ethnicity, karyotype, downSyndromeStatus)
```


Family ID: No changes needed

Family Relationship:
(Note: "of control" means "of D21" - NOT unrelated
Currently only classifying relationships to Proband, not within Unrelated Control families)

**Special cases (per Angela):**
- F0001 is a relative of a Proband who is no longer in the study
- 0052 and 0053 are Probands + adopted siblings so they have different Family IDs (F0034 
and F0031) -> Treat as unrelated; leave as Probands, delete "Other relation of Proband"
- 0603 is also adopted sib to another proband -> leave as Proband and delete "Other 
relation of Proband"
- 0566 and 0561 are twins - delete "Other relation of Proband" as that is just to trigger a
free text box in redcap
- HTP0077 has only one relationship = Other Relation of Control -> rename as Control manually

- e.g. F0072-F0078 has parents of Proband but not Proband - there are several like this
These won't have family member IDs because there's no Proband

```{r}
htp_relationships <- htp_participant %>%
  select(participantExternalId, sex_original, family_id, relationship, karyotype) %>% 
  #fix the above special cases
  mutate(relationship = case_when(
    participantExternalId %in% c("HTP0052", "HTP0053", "HTP0603") ~ "Proband",
    participantExternalId %in% c("HTP0566", "HTP0561") ~ "Proband,Sibling of Proband",
    participantExternalId == "HTP0077" ~ "Control - No relation to proband",
    TRUE ~ relationship
  )) %>% 
  separate_rows(relationship, sep = ",") %>% 
  mutate(relationship = str_trim(relationship)) %>%
  filter(!str_detect(relationship, "of Control")) %>% 
  mutate(familyRelationship = case_when(
    relationship == "Control - No relation to proband" ~ "Unrelated control",
    relationship == "Proband" ~ "Proband",
    relationship == "Sibling of Proband" ~ "Sibling",
    relationship == "Other Relation of Proband" ~ "Other relative", #HTP0001
    relationship == "Other relation of Proband" ~ "Other relative", #HTP0602
    relationship == "Parent of Proband" & sex_original == "Male" ~ "Father",
    relationship == "Parent of Proband" & sex_original == "Female" ~ "Mother"
  )) %>% 
  arrange(family_id)

# htp_relationships %>% select(participantExternalId, family_id, sex_original, relationship, familyRelationship) 

#filter(participantExternalId %in% c("HTP0052", "HTP0053", "HTP0603"))
```

<!-- Check participants who have more than 1 relationship listed -->
<!-- Make note of those for Family Type and Family Member IDs -->

<!-- ```{r} -->

<!-- htp_relationships %>%  -->
<!--   group_by(participantExternalId) %>%  -->
<!--   mutate(n = n()) %>%  -->
<!--   arrange(desc(n)) %>%  -->
<!--   filter(n >= 2) -->

<!-- ``` -->

Family Member IDs:
```{r}
# mother & father IDs
parent_ids <- htp_relationships %>%
  arrange(family_id) %>% 
  select(participantExternalId, family_id, familyRelationship) %>%
  filter(str_detect(familyRelationship, "Proband|Mother|Father")) %>% 
  pivot_wider(id_cols = family_id,
              names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  unnest() %>% 
  rename(participantExternalId = Proband,
         motherId = Mother,
         fatherId = Father)


sibling_ids <- htp_relationships %>%
  arrange(family_id) %>% 
  select(participantExternalId, family_id, familyRelationship) %>%
  filter(str_detect(familyRelationship, "Proband|Sibling")) %>% 
  pivot_wider(id_cols = family_id,
              names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  unnest_wider(Sibling, names_sep = "") %>% 
  unnest(Proband) %>% 
  unite(col = "siblingId", starts_with("Sibling"), sep = ", ", na.rm = TRUE) %>%
  filter(siblingId != "") %>% 
  rename(participantExternalId = Proband) %>% 
  #remove probands that are listed as their own sib (eg. 0295)
  mutate(siblingId = str_replace(siblingId, paste0(.$participantExternalId, ", "), "")) %>% 
  mutate(siblingId = str_replace(siblingId, paste0(", ", .$participantExternalId), ""))

# other relative IDs - looks like there are no families with Other Relatives and Probands
# (F0001 and F0058 have Other Relatives that aren't twins/etc but those families don;t have
# Probands)
other_rel_ids <- htp_relationships %>%
  arrange(family_id) %>% 
  select(participantExternalId, family_id, familyRelationship) %>%
  filter(str_detect(familyRelationship, "Proband|Other")) %>% 
  pivot_wider(id_cols = family_id,
              names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  unnest_wider(`Other relative`, names_sep = "") %>% 
  unnest(Proband) %>% 
  unite(col = "otherFamilyMemberId", starts_with("Other"), sep = ", ", na.rm = TRUE) %>% 
  filter(otherFamilyMemberId != "") %>% 
  rename(participantExternalId = Proband) %>% 
  #remove probands that are listed as their own other relative (e.g. HTP0602) 
  mutate(otherFamilyMemberId = na_if(otherFamilyMemberId, participantExternalId))

htp_relationships <- htp_relationships %>% 
  left_join(parent_ids) %>% 
  left_join(sibling_ids) %>% 
  left_join(other_rel_ids) %>% 
  group_by(participantExternalId) %>%  
  #combine multiple relationships into single row (e.g. Proband, Sibling)
  mutate(familyRelationship = paste(familyRelationship, collapse = ", ")) %>%
  #for now, if multiple relationships, only take Proband (due to data model enums)
  mutate(familyRelationship = if_else(
    str_detect(familyRelationship, "Proband, "), "Proband", familyRelationship)) %>% 
  ungroup() %>% 
  #remove duplicates caused by splitting out relationships into separate rows
  select(-relationship) %>% 
  distinct()

# select(participantExternalId, family_id, familyRelationship, motherId, fatherId, siblingId, otherFamilyMemberId) %>% arrange(family_id) %>%
#mutate(n = n()) %>% arrange(desc(n)) %>%
```




Family Type:
**DO WE NEED DUO+ OR PROBAND+SIBLING?**
SPECIAL CASES: 
F0001 has a Proband that is not in this dataset -> should be listed as familyType Other
F0058, F0072-78 have family members but no Proband
F0162, F0084 have 2 probands

Use FamilyID instead of Proband, otherwise of families with no Proband will be left out

**define OTHER as also including either control families OR Proband withdrawn due to consent)**
```{r}
htp_family_types <- htp_relationships %>% 
  select(-participantExternalId, -sex_original) %>%
  group_by(family_id) %>%
  arrange(familyRelationship, .by_group = TRUE) %>% #alphabetical order for consistency 
  mutate(family_list = paste(familyRelationship, collapse = "|"), .after = family_id) %>% 
  mutate(family_count = max(rowSums(!is.na(across(c(motherId, fatherId, siblingId,
                                                    otherFamilyMemberId))))), .after = family_list) %>% 
  select(family_id, family_list, family_count) %>%
  distinct() %>% 
  ungroup() %>% #make sure to ungroup or this will run SLOW
  mutate(familyType = case_when(
    family_list == "Proband" ~ "Proband-only",
    family_list == "Mother|Proband" ~ "Duo",
    family_list == "Father|Proband" ~ "Duo",
    family_list == "Father|Mother|Proband" ~ "Trio",
    family_count > 2 & str_detect(family_list, "Mother") & str_detect(family_list, "Father") ~ "Trio+",
    TRUE ~ "Other")
  ) %>% 
  select(family_id, familyType)
```


Join family stuff

```{r}
htp_participant <- htp_participant %>% 
  arrange(family_id) %>% 
  left_join(htp_family_types, by = "family_id") %>% 
  left_join(htp_relationships %>% select(-sex_original, -family_id), 
            by = "participantExternalId")
```

Get Global IDs:
**NOTE:** some of my global IDs have been replaced with KF IDs - be sure to download the 
latest Global IDs from Dewrangle next time
```{r}
htp_dewrangle_input <- htp_participant %>% 
  distinct(participantExternalId) %>% 
  rename(descriptor = participantExternalId) %>% 
  mutate(fhir_resource_type = "patient")

write_csv(htp_dewrangle_input, 
          here::here("output", "htp_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

htp_global_ids <- read_csv(here::here("output", "global-descriptors-HTP.csv"))

htp_participant <- htp_participant %>% 
  left_join(htp_global_ids %>% select(descriptor, global_id),
            by = c("participantExternalId" = "descriptor")) 

#%>% select(participantExternalId, global_id) %>% filter(is.na(global_id))
```


Create final Participant table:

- Fill in info that is same for every participant (mutate)
- Rename columns that need renamed with schema labels (also mutate)
- select all Participant columns
- rename with display names

```{r}
htp_participant <- htp_participant %>% 
  mutate(studyCode = "HTP",
         participantGlobalId = global_id,
         familyId = family_id,
         sex = sex_original,
         ageAtFirstPatientEngagement = age_at_enrollment_days,
         firstPatientEngagementEvent = "Enrollment",
         outcomesVitalStatus = NA,
         ageAtLastVitalStatus = NA) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)

##TODO: check against data
```

```{r}
write_csv(htp_participant, here::here("output", "htp_participant.csv"))
```


##### CONDITION

Get HTP condition & measure codes
```{r}
condition_lookup_htp <- condition_codes_v3 %>% filter(study == "HTP")
measure_lookup_htp <- measure_codes_v3 %>% filter(study == "HTP")
```

<!-- Data Check: confirm that every set of visits for a participant has the same phenotype values (i.e. 1) -->
<!-- ```{r} -->
<!-- htp_clinical_harmonized %>%  -->
<!--   select(-(lab_id:mr_abstraction_status)) %>%  -->
<!--   group_by(participantExternalId) %>%  -->
<!--   distinct() %>%  -->
<!--   count() %>%  -->
<!--   arrange(desc(n)) -->

<!-- test <- htp_clinical_harmonized %>%  -->
<!--   filter(participantExternalId == "HTP0332") %>%  -->
<!--   select(-(participantExternalId:mr_abstraction_status)) -->

<!-- test[1,] == test[2,] -->
<!-- ``` -->

Pull conditions from clinical data
** SPECIAL CASE: Delete column "Anatomical Anomaly" and all "Unsure" cols, per Angela
```{r}
htp_conditions_subset <- htp_clinical_harmonized %>% 
  select(-(lab_id:mr_abstraction_status), -`Anatomical anomaly`, -starts_with("Unsure")) %>%
  group_by(participantExternalId) %>% 
  distinct() %>% 
  ungroup() %>% 
  pivot_longer(cols = c(-participantExternalId), 
               names_to = "source_text",
               values_to = "status") %>%
  filter(status == TRUE) %>% 
  left_join(condition_lookup_htp, by = c("source_text" = "source_column")) %>% 
  select(-study, -source_column_value)
```

Add DS Diagnosis (T21 participants only)
**participants w/ T21 but OfficialDSDiagnosis is blank -> change to Unsure**
(currently this only applies to HTP0888)

```{r}
htp_ds_diagnoses <- htp_clinical_harmonized %>% 
  select(participantExternalId, karyotype,
         source_text = official_ds_diagnosis) %>%
  mutate(source_text_edited = case_when(
    karyotype == "Trisomy 21" & is.na(source_text) ~ "Unsure",
    TRUE ~ source_text
  )) %>% 
  filter(!is.na(source_text_edited)) %>% 
  distinct() %>%
  mutate(status = "TRUE") %>% 
  left_join(condition_lookup_htp %>% 
              filter(source_column == "official_ds_diagnosis") %>% 
              unnest(),
            by = c("source_text_edited" = "source_column_value")) %>% 
  select(-study,-karyotype, -source_column, -source_text) %>% 
  rename(source_text = source_text_edited)
```

Combine conditions + ds_diagnoses
Create empty/all same value columns
Select columns in order of DCA template
```{r}
htp_conditions <- htp_ds_diagnoses %>% 
  rbind(htp_conditions_subset) %>% 
  arrange(participantExternalId) %>% 
  mutate(studyCode = "HTP",
         eventId = NA,
         eventType = NA,
         ageAtConditionMeasureObservation = NA,
         conditionInterpretation = "Observed",
         conditionStatus = "History Of",
         conditionDataSource = "Clinical",
         measureValue = NA,
         measureUnit = NA) %>%
  select(all_of(condition_schemalabels))

```

Measures:
**NOTE: decimal number vary when displayed in R and Excel**
**Actual numbers match up in output csv and Excel when you click on the cell**

```{r}
htp_measures <- htp_clinical_harmonized %>% 
  select(participantExternalId, event_name, bmi, weight_kg, height_cm, age_at_visit_days) %>% 
  # #round to number of decimals specified in measure_codes_v3
  # mutate(bmi = round(as.numeric(bmi), 1),
  #        weight_kg = round(as.numeric(weight_kg), 1),
  #        height_cm = round(as.numeric(height_cm), 1)) %>% 
  pivot_longer(cols = c(bmi, weight_kg, height_cm),
               names_to = "source_text",
               values_to = "measureValue") %>% 
  left_join(measure_lookup_htp,
            by = c("source_text" = "source_column")) %>% 
  mutate(studyCode = "HTP",
         eventId = event_name,
         eventType = "Visit",
         ageAtConditionMeasureObservation = age_at_visit_days,
         conditionInterpretation = NA,
         conditionStatus = NA, conditionDataSource = NA,
         hpoLabel = NA,
         hpoCode = NA,
         mondoLabel = NA,
         mondoCode = NA,
         maxoLabel = NA,
         maxoCode = NA,
         measureValue = as.numeric(measureValue)
  ) %>% 
  select(all_of(condition_schemalabels))
```


Join conditions & measures, rename with display names

```{r}
htp_condition <- htp_conditions %>% 
  rbind(htp_measures) %>% 
  arrange(participantExternalId, eventId) %>% 
  setNames(condition_displaynames)
```


```{r}
write_csv(htp_condition, here::here("output", "htp_condition.csv"))
```


##### Add Global IDs to Eric's v2 sample manifest and rename cols in sample manifests

```{r}
htp_sample <- read_csv("C:/Users/pierrettelo/Documents/INCLUDE/Biospecimen-FileManifests-pre-v3/Biospecimen-FileManifests-pre-v3/HTP/htp_sample.csv") %>% 
  rename(`Participant External ID` = `Subject ID`,
         `Sample External ID` = `Sample ID`,
         `Age at Biospecimen Collection` = `Age at Collection`,
         `Parent Sample External ID` = `Parent Sample ID`
  )

htp_sample_dewrangle_input <- htp_sample %>%
  select(descriptor = `Sample External ID`) %>% 
  mutate(fhir_resource_type = "specimen")

write_csv(htp_sample_dewrangle_input, here::here("output", "htp_sample_dewrangle_input.csv"))

htp_sample_global_ids <- read_csv(here::here("output", "global-descriptors-HTP.csv")) %>% 
  filter(str_detect(fhir_resource_type, "Specimen")) %>% 
  select(descriptor, global_id)

htp_sample <- htp_sample %>% 
  left_join(htp_sample_global_ids, by = c(`Sample External ID` = "descriptor")) %>% 
  rename(`Sample Global ID` = global_id) %>% 
  relocate(`Sample Global ID`, .before = `Sample External ID`)

write_csv(htp_sample, here::here("output", "htp_sample_renamed_with_global_ids.csv"))

```


```{r}
htp_aliquot <- read_csv("C:/Users/pierrettelo/Documents/INCLUDE/Biospecimen-FileManifests-pre-v3/Biospecimen-FileManifests-pre-v3/HTP/aliquot.csv") %>% 
  rename(`Sample External ID` = `Sample ID`,
         `Container External ID` = Barcode,
         Volume = `Vial Volume`)


updated_htp_sample_collection_supplement <- read_csv("C:/Users/pierrettelo/Documents/INCLUDE/Biospecimen-FileManifests-pre-v3/Biospecimen-FileManifests-pre-v3/HTP/updated-htp_sample_collection_supplement.csv") %>% 
  rename(`Participant External ID` = `Subject ID`,
         `Sample External ID` = `Sample ID`,
         `Age at Biospecimen Collection` = `Age at Collection`,
         `Parent Sample External ID` = `Parent Sample ID`
  )

write_csv(htp_aliquot, here::here("output", "htp_aliquot_renamed.csv"))

write_csv(updated_htp_sample_collection_supplement, here::here("output", "updated_htp_sample_collection_supplement_renamed.csv"))

```


#### MATT'S VBR EXTRACT - Experimentation

```{r}
matt_vbr <- read_tsv(synGet("syn51384359")$path)

```

Get unique hierarchies
```{r}
matt_vbr %>% 
  select(CollectionHierarchy) %>% 
  mutate(hierarchy_types = str_remove_all(CollectionHierarchy, "HTP.*?(?=_)")) %>%
  mutate(hierarchy_types = str_remove_all(hierarchy_types, "_")) %>%
  select(hierarchy_types) %>% 
  distinct() %>%
  arrange(desc(hierarchy_types)) %>% 
  view()
```


Which lab procedure goes with which sample type
```{r}
matt_vbr %>% 
  select(CollectionSampleType, ParentSampleType, SampleType, LaboratoryProcedure) %>% 
  distinct() %>% 
  arrange(desc(CollectionSampleType), desc(ParentSampleType)) %>% 
  view()
```


#### Jamed list of IDs with no data

```{r}
no_data <- c("HTP0153", "HTP0157", "HTP0164", "HTP0167", "HTP0168", "HTP0169", "HTP0170", "HTP0171", "HTP0173", "HTP0174", "HTP0177", "HTP0178", "HTP0188", "HTP0189", "HTP0190")

htp_clinical_harmonized %>% 
  filter(participantExternalId %in% no_data)

```


#### Check my original Global IDs against Christina's

```{r}
# global IDs i created in dewrangle from Eric's v2 manifest
old_global_ids <- read_csv(here::here("output", "global-descriptors-HTP.csv"))

# global IDs that Christina and Marcel curated with original KF IDs
new_global_ids <- read_csv("C:/Users/pierrettelo/Documents/INCLUDE/htp_global_kfids_for_pierrette.csv")
```

```{r}
new_global_ids %>% filter(str_detect(descriptor, "84A2_RNA"))
```


```{r}
old_global_ids %>% 
  left_join(new_global_ids %>% select(-fhir_resource_type), by = "descriptor") %>% 
  #filter(!is.na(global_id.y)) %>% 
  view()
```
























