---
title: "Harmonize DSC clinical data, March 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```

Read in data (Acknowledgments row already deleted in Synapse)
```{r}
dsc_demographic <- read_csv(synGet("syn26445516")$path)
dsc_ihq <- read_csv(synGet("syn26445517")$path)
```


<!-- ##### DATA CHECKS -->

<!-- ```{r} -->
<!-- n_distinct(dsc_demographic$patient_id) #3484 -->
<!-- n_distinct(dsc_ihq$patient_id) #3484 -->

<!-- # all pts with ihq have demographics -->
<!-- setdiff(dsc_ihq$patient_id, dsc_demographic$patient_id) -->

<!-- # parent_ids that don't exist as patient_ids - only 5455 -->
<!-- setdiff(dsc_ihq$parent_id, dsc_ihq$patient_id) -->
<!-- ``` -->


##### PARTICIPANT

Get ETL plan
```{r}
dsc_etl_plan <- etl_to_linkml_plan %>% filter(study == "DSC")
```


Bring in parent_id from ihq
Family ID = parent_id if exists, otherwise = NA

```{r}
dsc_demo_harmonized <- dsc_demographic %>% 
  left_join(dsc_ihq %>% select(patient_id, parent_id) %>% distinct()) %>% 
  relocate(parent_id, .after = patient_id) %>% 
  mutate(familyId = case_when(
    patient_id %in% dsc_ihq$parent_id ~ patient_id,
    !is.na(parent_id) ~ parent_id),
    .after = parent_id) 

#%>% arrange(familyId)
```

Family Type:
"Proband-only" if familyId = NA
"Other" if not NA
**may want to add "Siblings" to Family Type in schema**

Family relationships - all are Probands but list siblings of each other

```{r}
dsc_demo_harmonized <- dsc_demo_harmonized %>%
  mutate(familyType = if_else(is.na(familyId), "Proband-only", "Other"),
         familyRelationship = "Proband") 

#%>% select(patient_id, parent_id, familyId, familyType, familyRelationship) %>% arrange(familyId)
```


Family IDs:
The parent_id is always the min(patient_id) of the family group

```{r}
dsc_sibling_ids <- dsc_demo_harmonized %>%
  filter(!is.na(familyId)) %>% 
  group_by(familyId) %>% 
  mutate(siblingId = case_when(
    patient_id != familyId ~ parent_id,
    patient_id == min(patient_id) ~ max(patient_id)
  )) %>%
  ungroup() %>% 
  select(patient_id, siblingId)

dsc_demo_harmonized <- dsc_demo_harmonized %>% 
  left_join(dsc_sibling_ids) 

#%>% select(patient_id, familyId, siblingId)
```

Race:
- A couple of participants (4068 & 6181) chose all the options including Unknown -> assume they are "more than one race"
- Not specifying the different races (if >1) for now - maybe in a future release
- "Prefer not to answer" if all race columns are blank (don't see any of those in this
dataset)

Create race_original column combining "1" answers from all race cols
```{r}
dsc_races <- dsc_demo_harmonized %>% 
  select(patient_id, contains("race")) %>% 
  pivot_longer(cols = -patient_id,
               names_to = "race_cat",
               values_to = "status") %>% 
  filter(status == 1) %>% 
  group_by(patient_id) %>% 
  summarize(race_combined = paste(race_cat, collapse = "|")) %>% 
  mutate(race_original = case_when(
    str_detect(race_combined, "\\|") ~ "More than one race",
    TRUE ~ race_combined
  )) %>% 
  select(-race_combined) 

```


```{r}
dsc_race_lookup_values <- dsc_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>%
  mutate(original_value = na_if(original_value, "NA"))

dsc_ethnicity_lookup_values <- dsc_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>%
  mutate(original_value = na_if(original_value, "NA"))

dsc_vital_lookup_values <- dsc_etl_plan %>% 
  filter(model_slot == "outcomesVitalStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>%
  mutate(original_value = na_if(original_value, "NA"))

dsc_demo_harmonized <- dsc_demo_harmonized %>% 
  left_join(dsc_races) %>% #get race_original col
  left_join(dsc_ihq %>% select(patient_id, vital_status)) %>%  #get vital status
  rename(ethnicity_original = ethnicity) %>% #rename col to avoid confusion
  left_join(dsc_race_lookup_values, by = c("race_original" = "original_value")) %>% 
  rename("race" = model_enum_label) %>%  
  left_join(dsc_ethnicity_lookup_values, by = c("ethnicity_original" = "original_value")) %>%
  rename("ethnicity" = model_enum_label)  %>%  
  left_join(dsc_vital_lookup_values, by = c("vital_status" = "original_value")) %>%
  rename("outcomesVitalStatus" = model_enum_label) 

#%>% select(patient_id, race_original, race, ethnicity_original, ethnicity, vital_status, outcomesVitalStatus)  

```

<!-- Age at Vital Status - leave out for now due to below inconsistencies -->
<!-- ```{r} -->
<!-- # 4 patients have vital_status = Yes (Alive) but deceased_year is not NA -->
<!-- # 6725 = survey_time 2015 -->
<!-- # 6841 = 2016 -->
<!-- # 11358 = 2019 -->
<!-- # 11823 = 2020 -->
<!-- dsc_demo_transformed %>%  -->
<!--   left_join(dsc_ihq %>% select(patient_id, vital_status, survey_time)) %>%  -->
<!--   select(patient_id, deceased_year, vital_status, survey_time) %>%  -->
<!--   filter(vital_status == "Yes" & !is.na(deceased_year)) -->

<!-- # 5 patients are dead but no deceased_year -->
<!-- dsc_demo_transformed %>%  -->
<!--   left_join(dsc_ihq %>% select(patient_id, vital_status)) %>%  -->
<!--   select(patient_id, deceased_year, vital_status) %>%  -->
<!--   filter(vital_status == "No") -->

<!-- # 1 pt (3981) has vital_status NA but has deceased_year -->
<!-- dsc_demo_transformed %>%  -->
<!--   left_join(dsc_ihq %>% select(patient_id, vital_status, survey_time)) %>%  -->
<!--   select(patient_id, deceased_year, vital_status, survey_time) %>%  -->
<!--   filter(!is.na(deceased_year) | vital_status == "No") %>%  -->
<!--   arrange(vital_status)  -->

<!-- ``` -->


Get Global IDs:

```{r}
dsc_dewrangle_input <- dsc_demo_harmonized %>% 
  distinct(patient_id) %>% 
  rename(descriptor = patient_id) %>% 
  mutate(fhir_resource_type = "patient")

write_csv(dsc_dewrangle_input, 
          here::here("output", "dsc_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

dsc_global_ids <- read_csv(here::here("output", "global-descriptors-DS-Connect.csv"))

dsc_demo_harmonized <- dsc_demo_harmonized %>% 
  left_join(dsc_global_ids %>% select(descriptor, global_id),
            by = c("patient_id" = "descriptor"))

#%>% select(patient_id, global_id) %>% filter(is.na(global_id))
```


Create final Participant table

```{r}
dsc_participant <- dsc_demo_harmonized %>%
  mutate(studyCode = "DSC",
         participantGlobalId = global_id,
         participantExternalId = patient_id,
         fatherId = NA,
         motherId = NA,
         otherFamilyMemberId = NA,
         sex = gender,
         downSyndromeStatus = "T21",
         ageAtFirstPatientEngagement = round(age_at_registration*365.25, 0),
         firstPatientEngagementEvent = "Enrollment",
         ageAtLastVitalStatus = NA
  ) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)
```

```{r}
write_csv(dsc_participant, here::here("output", "dsc_participant.csv"))
```


##### CONDITION

Special import of condition_codes_v3 with col type specification
(having mixed numeric & char values causes error when unnesting)
```{r}
gs4_deauth()

condition_lookup_dsc <- read_sheet("https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit?pli=1#gid=384137116",
                                   sheet = "condition_codes_v3",
                                   col_types = "c") %>% 
  filter(study == "DSC")
```

Special cols (ie. NOT 1):
  leukemia_ever_diagnosis
  dementia_ad_cognitive_ever_diagnosis
  dementia_ad_cognitive_ever_diagnosis
  dementia_ad_cognitive_ever_diagnosis
  celiac_ever_diagnosis -> gi_condition_celiac
  chd_ever_diagnosis -> chd_condition_
  skeletal_ever_diagnosis -> skeletal_condition_
  thyroid_ever_diagnosis -> thyroid_condition_
  seizures_epilepsy_ever_diagnosis
  gi_ever_diagnosis -> gi_condition_
  sleep_ever_diagnosis -> sleep_condition_
  covid19_ever_diagnosis
  diabetes_ever_diagnosis
  eye_exam_ever
  hearing_testing_ever
  ds_specific_diagnosis


Data checks - CHD:
```{r}
# patients where chd_ever = Yes but doesn't have other chd_conditions: 17
# https://stackoverflow.com/a/62162600
dsc_ihq %>%
  select(patient_id, chd_ever_diagnosis, starts_with("chd_condition_"), -chd_condition_none) %>%
  rowwise() %>%
  filter(chd_ever_diagnosis == "Yes" &
           all(is.na(c_across(starts_with("chd_condition")))))

# patients where chd_ever = No but has other chd_conditions?
dsc_ihq %>%
  select(patient_id, chd_ever_diagnosis, starts_with("chd_condition_"), -chd_condition_none) %>%
  rowwise() %>%
  filter(chd_ever_diagnosis == "No" &
           any(!is.na(c_across(starts_with("chd_condition")))))

# or patients where chd_ever = Yes but chd_condition_none = also Yes
dsc_ihq %>%
  filter(chd_ever_diagnosis == "Yes" & chd_condition_none == 1) %>%
  pull(patient_id)

```

<!-- Check GI cols -->
<!-- ```{r} -->
<!-- # are there any patients where gi_ever = Yes but doesn't have other gi_conditions? 37 -->
<!-- dsc_ihq %>%  -->
<!--   select(patient_id, gi_ever_diagnosis, starts_with("gi_condition_"), -gi_condition_none) %>%  -->
<!--   rowwise() %>%  -->
<!--   filter(gi_ever_diagnosis == "Yes" &  -->
<!--            all(is.na(c_across(starts_with("gi_condition"))))) -->

<!-- # or patients where gi_ever = No but has other gi_conditions?  146 -->
<!-- dsc_ihq %>%  -->
<!--   select(patient_id, gi_ever_diagnosis, starts_with("gi_condition_"), -gi_condition_none) %>%  -->
<!--   rowwise() %>%  -->
<!--   filter(gi_ever_diagnosis == "No" &  -->
<!--            any(!is.na(c_across(starts_with("gi_condition"))))) -->

<!-- # or patients where gi_ever = Yes but gi_condition_none = also Yes -->
<!-- dsc_ihq %>%  -->
<!--   filter(gi_ever_diagnosis == "Yes" & gi_condition_none == 1) %>%  -->
<!--   pull(patient_id) -->

<!-- ``` -->


CHD cols - just choose any positive assertion:
- chd_ever_diagnosis = Yes and no other selections, or chd_condition_unsure -> CHD
- chd_ever_diagnosis = Yes and chd_condition_none = 1 -> still CHD (maybe they don't have
CHD now but they did at some point?)
- Otherwise go with the other selections (chd_)
- leave out chd_condition_unsure

GI cols
- gi_ever_diagnosis = Yes and no other selections, or gi_condition_unsure -> GI
- gi_ever_diagnosis = Yes and gi_condition_none = 1 -> still GI
- Otherwise go with the other selections (gi_condition_)
- leave out gi_condition_unsure
- celiac_ever_diagnosis = Yes OR gi_condition_celiac = 1 -> choose the positive assertion, ignore negative (will need to distinct() if both)

Sleep cols
- sleep_ever_diagnosis = Yes and no other selections, or sleep_condition_unsure -> sleep
- sleep_ever_diagnosis = Yes and sleep_condition_none = 1 -> still sleep
- Otherwise go with the other selections (sleep_)
- leave out sleep_condition_unsure

skeletal cols
- skeletal_ever_diagnosis = Yes and no other selections, or skeletal_condition_unsure -> skeletal
- skeletal_ever_diagnosis = Yes and skeletal_condition_none = 1 -> still skeletal
- Otherwise go with the other selections (skeletal_)
- leave out skeletal_condition_unsure

thyroid cols
- thyroid_ever_diagnosis = Yes and no other selections, or thyroid_condition_unsure -> thyroid
- thyroid_ever_diagnosis = Yes and thyroid_condition_none = 1 -> still thyroid
- Otherwise go with the other selections (thyroid_)
- leave out thyroid_condition_unsure


Pull conditions from ihq
```{r}
dsc_conditions_subset <- dsc_ihq %>% 
  select(patient_id, ds_specific_diagnosis,
         starts_with(c("chd", "cardiac", "gi", "skin", "seizures",
                       "neurologic", "sleep", "skeletal", "leukemia",
                       "cancer", "covid19", "thyroid", "diabetes", "celiac",
                       "metabolic", "hearing", "eye", "mental", "learning", "dementia"))) %>% 
  mutate(across(everything(), as.character)) %>% 
  pivot_longer(cols = -patient_id,
               names_to = "source_column",
               values_to = "status") %>% 
  filter(!is.na(status) & status != "No") %>% # delete blanks and _ever = No
  left_join(condition_lookup_dsc, by = c("source_column", "status" = "source_column_value")) %>% 
  filter(!is.na(conditionMeasureSourceText))  # remove conditions like _none and _unsure that don't have annotations
```

Deal with _ever columns:

If a patient has ever_diagnosis AND _condition starting with the same prefix -> 
delete ever, keep the more specific condition
```{r}
keep_specific_conditions <- dsc_conditions_subset %>% 
  filter(str_detect(source_column, "^chd|gi_|sleep|skeletal|thyroid") & !str_detect(source_column, "surgery")) %>% 
  mutate(prefix = str_extract(source_column, "^[^_]+(?=_)"), .before = source_column) %>% 
  group_by(patient_id, prefix) %>% # group by patient and condition category
  mutate(n = n(), .before = prefix) %>% 
  arrange(n) %>% 
  filter(n > 1) %>% # filter out groups that only have 1 condition per category (could be _ever or something else) 
  filter(any(status == "Yes")) %>% # get only groups that include _ever
  ungroup() %>% 
  filter(status == 1) # remove evers, keep specific condition

#get back the groups with multiple conditions but no ever
keep_multiple_without_ever <- dsc_conditions_subset %>% 
  filter(str_detect(source_column, "^chd|gi_|sleep|skeletal|thyroid") & !str_detect(source_column, "surgery")) %>% 
  mutate(prefix = str_extract(source_column, "^[^_]+(?=_)"), .before = source_column) %>% 
  group_by(patient_id, prefix) %>% # group by patient and condition category
  mutate(n = n(), .before = prefix) %>% 
  arrange(n) %>% 
  filter(n > 1) %>% # filter groups that only have 1 condition per category (could be _ever or something else) 
  filter(all(status != "Yes")) %>% # get only groups that don't have _ever
  ungroup()

#get back the groups with 1 condition per category (incl)
keep_single_conditions <- dsc_conditions_subset %>% 
  filter(str_detect(source_column, "^chd|gi_|sleep|skeletal|thyroid") & !str_detect(source_column, "surgery")) %>% 
  mutate(prefix = str_extract(source_column, "^[^_]+(?=_)"), .before = source_column) %>% 
  group_by(patient_id, prefix) %>% # group by patient and condition category
  mutate(n = n(), .before = prefix) %>% 
  arrange(n) %>% 
  filter(n == 1) %>% 
  ungroup()

filtered_ever_conditions <- rbind(keep_specific_conditions, keep_multiple_without_ever, keep_single_conditions) %>% 
  select(-n, -prefix)

# check for duplicate rows
dsc_conditions_subset %>% 
  filter(!str_detect(source_column, "^chd|gi_ever|gi_condition|sleep|skeletal|thyroid")) %>% 
  rbind(filtered_ever_conditions) %>% 
  janitor::get_dupes() 

# add these back to the rest of the conditions
dsc_conditions_final <- dsc_conditions_subset %>% 
  filter(!str_detect(source_column, "^chd|gi_ever|gi_condition|sleep|skeletal|thyroid")) %>% 
  rbind(filtered_ever_conditions)

#dsc_conditions_final %>% arrange(patient_id)
```

Final table:
Create empty/all same value columns
Select columns in order of DCA template
```{r}
dsc_condition <- dsc_conditions_final %>%
  left_join(dsc_ihq %>% 
              mutate(patient_id = as.character(patient_id)) %>% 
              select(patient_id, session_id)) %>% 
  mutate(studyCode = "DSC",
         participantExternalId = patient_id,
         eventId = session_id,
         eventType = "Survey",
         ageAtConditionMeasureObservation = NA,
         conditionInterpretation = "Observed",
         conditionDataSource = "Self-reported",
         conditionStatus = "History Of",
         measureValue = NA,
         measureUnit = NA) %>%
  select(all_of(condition_schemalabels)) %>% 
  setNames(condition_displaynames)

```


```{r}
write_csv(dsc_condition, here::here("output", "dsc_condition.csv"))
```
