---
title: "Harmonize Hakon to LinkML v2 model - March 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```


Read in data/Initial cleanup:

```{r}
hakon_clin_dna <- read_xlsx(synGet("syn44291080")$path)
hakon_clin_rna <- read_xlsx(synGet("syn44291064")$path)

#delete the last few cols since they are for solid tumors
hakon_clinical_source <- hakon_clin_dna %>% 
  rbind(hakon_clin_rna) %>% 
  select(-(`Days to First Event`:`Tumor resection extent`))
```

Replace headers with working_headers
See Google sheet `INCLUDE clinical data dictionaries` -> `X01-Hakon`
```{r}
gs4_deauth()

hakon_working_headers <- read_sheet(
  "https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw/edit#gid=0",
  sheet = "X01-Hakon")

hakon_clinical_source <- setNames(hakon_clinical_source, hakon_working_headers$working_header)
```

Get ETL plan
```{r}
# source("harmonization_setup_2023-03.R")

hakon_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "X01-Hakon")
```

Participant External ID
```{r}
hakon_clinical_harmonized <- hakon_clinical_source %>% 
  rename(participantExternalId = participant_id) 
```

## PARTICIPANT

Remove Condition/Sample cols and take distinct (many participants have RNA and DNA)
```{r}
hakon_participant <- hakon_clinical_harmonized %>% 
  select(participantExternalId, starts_with("family"), affected_status,
         age_at_sample_procurement_days, sex_original, race_original,
         ethnicity_original, vital_status, age_at_last_known_vital_status) %>% 
  distinct()
```

Cols that need transformed:
- Sex
- Race
- Ethnicity
- DS status

```{r}
hakon_sex_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "sex") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_race_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_eth_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_ds_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "downSyndromeStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_vital_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "outcomesVitalStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>% 
  na_if("NA")

hakon_participant <- hakon_participant %>% 
  left_join(hakon_sex_lookup_values, by = c("sex_original" = "original_value")) %>% 
  rename("sex" = model_enum_label) %>% 
  left_join(hakon_race_lookup_values, by = c("race_original" = "original_value")) %>% 
  rename("race" = model_enum_label) %>%  
  left_join(hakon_eth_lookup_values, by = c("ethnicity_original" = "original_value")) %>% 
  rename("ethnicity" = model_enum_label)%>%  
  left_join(hakon_ds_lookup_values, by = c("affected_status" = "original_value")) %>% 
  rename("downSyndromeStatus" = model_enum_label) %>% 
  left_join(hakon_vital_lookup_values, by = c("vital_status" = "original_value")) %>% 
  rename("outcomesVitalStatus" = model_enum_label) 

#%>% select(participantExternalId, sex_original, sex, race_original, race, ethnicity_original, ethnicity, affected_status, downSyndromeStatus, vital_status, outcomesVitalStatus)
```


Family relationships:

```{r}
hakon_family_lookup_values <- hakon_etl_plan %>% 
  filter(model_slot == "familyRelationship") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

hakon_participant <- hakon_participant %>% 
  left_join(hakon_family_lookup_values, by = c("family_relationship" = "original_value")) %>% 
  rename("familyRelationship" = model_enum_label) 

# %>% select(participantExternalId, family_relationship, familyRelationship)
```

Family member IDs - match familyRelationship with column for target ID
Do each relative type separately b/c chaining multiple unnests acts weird - drops rows 
```{r}
hakon_mother_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Mother") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  rename(motherId = Mother)

hakon_father_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Father") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  rename(fatherId = Father)  

hakon_sibling_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Sibling") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  unnest_wider(Sibling, names_sep = "") %>% 
  unite(col = "siblingId", starts_with("Sibling"), sep = ", ", na.rm = TRUE)

hakon_other_ids <- hakon_participant %>%
  arrange(family_id) %>% 
  select(proband_id = family_relationship_target_participant_id, familyRelationship, 
         participantExternalId) %>%
  filter(familyRelationship == "Other relative") %>% 
  pivot_wider(names_from = familyRelationship,
              values_from = participantExternalId) %>% 
  unnest_wider(`Other relative`, names_sep = "") %>% 
  unite(col = "otherFamilyMemberId", starts_with("Other"), sep = ", ", na.rm = TRUE)

hakon_participant <- hakon_participant %>% 
  arrange(family_id) %>% 
  left_join(hakon_mother_ids, by = c("participantExternalId" = "proband_id")) %>% 
  left_join(hakon_father_ids, by = c("participantExternalId" = "proband_id")) %>% 
  left_join(hakon_sibling_ids, by = c("participantExternalId" = "proband_id")) %>% 
  left_join(hakon_other_ids, by = c("participantExternalId" = "proband_id"))
```


Family Type:
- Proband-only: familyRel = Proband AND other ids = NA
- Duo: familyRel = Proband AND has motherId OR fatherId AND sib & other = NA
- Trio: familyRel = Proband AND has motherId AND fatherId ANd sib/other = NA
- Trio+: familyRel = Proband AND has motherId AND fatherId AND sib or other

find rows with 1 non-NA value: https://stackoverflow.com/a/73359219

```{r}
hakon_family_types <- hakon_participant %>% 
  filter(familyRelationship == "Proband") %>% 
  select(participantExternalId, motherId, fatherId, siblingId, otherFamilyMemberId) %>%
  mutate(family_count = rowSums(!is.na(across(c(motherId, fatherId, siblingId, otherFamilyMemberId))))) %>% 
  mutate(familyType = case_when(
    family_count == 0 ~ "Proband-only",
    family_count == 1 & (!is.na(motherId) | !is.na(fatherId)) ~ "Duo",
    family_count == 2 & (!is.na(motherId) & !is.na(fatherId)) ~ "Trio",
    family_count > 2 & (!is.na(motherId) & !is.na(fatherId)) ~ "Trio+",
    TRUE ~ "Other"
  ))

hakon_participant <- hakon_participant %>%
  left_join(hakon_family_types) %>% 
  arrange(family_id) %>% 
  group_by(family_id) %>% #fill familyType for non-Probands
  fill(familyType, .direction = "downup") %>% 
  ungroup()

#%>% select(participantExternalId, family_id, familyRelationship, familyType)

```

Create final Participant table:

- Fill in info that is same for every participant (mutate)
- Rename columns that need renamed with schema labels (also mutate)
- select all Participant columns
- rename with display names

```{r}
hakon_participant <- hakon_participant %>% 
  mutate(studyCode = "X01-Hakon",
         participantGlobalId = NA,
         familyId = family_id,
         ageAtFirstPatientEngagement = as.numeric(age_at_sample_procurement_days),
         firstPatientEngagementEvent = "Biospecimen collection",
         ageAtLastVitalStatus = as.numeric(age_at_last_known_vital_status)) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)
```

```{r}
write_csv(hakon_participant, here::here("output", "hakon_participant.csv"))
```


##### CONDITION

Get condition codes
```{r}
condition_lookup_hakon <- condition_codes_v3 %>% 
  filter(study == "X01-Hakon") %>% 
  mutate(source_column_value = as.character(source_column_value))
```


Pull conditions from clinical data
SPECIAL CASE: Replace "Respiratory failure; insufficiency; arrest" with "Respiratory failure, insufficiency, arrest"
```{r}
hakon_conditions <- hakon_clinical_harmonized %>% 
  select(participantExternalId, phenotypes_text) %>% 
  distinct() %>% #bc pts can have dna & rna
  mutate(phenotypes_text = str_replace(phenotypes_text, "Respiratory failure; insufficiency; arrest", "Respiratory failure, insufficiency, arrest")) %>% 
  separate_rows(phenotypes_text, sep = ";") %>% 
  mutate(phenotypes_text = str_trim(phenotypes_text)) %>%
  left_join(condition_lookup_hakon, by = c("phenotypes_text" = "source_column_value")) %>% 
  select(-phenotypes_text, -study, -source_column)

# %>% mutate(ontology_count = rowSums(!is.na(across(c(hpoCode, mondoCode, maxoCode, otherCode))))) 
# %>% arrange(ontology_count)

```

Add DS Diagnosis
```{r}
hakon_ds <- hakon_clinical_harmonized %>% 
  select(participantExternalId, diagnoses_text, age_at_diagnosis_days) %>%
  distinct() %>% 
  filter(!is.na(diagnoses_text)) %>% 
  left_join(condition_lookup_hakon, by = c("diagnoses_text" = "source_column_value")) %>% 
  select(-diagnoses_text, -study, -source_column)

hakon_conditions <- hakon_conditions %>% 
  bind_rows(hakon_ds)
```

Add age at condition observation (from Hakon, added directly to Condition table)
```{r}
hakon_condition_ages <- read_xlsx(synGet("syn51181708")$path) %>% 
  select(participant_external_id, condition_or_measure_source_text, age_at_condition_or_measure_observation)

hakon_conditions <- hakon_conditions %>% 
  left_join(hakon_condition_ages,
            by = c("participantExternalId" = "participant_external_id",
                   "conditionMeasureSourceText" = "condition_or_measure_source_text"))

```


Combine conditions + ds_diagnoses
Create empty/all same value columns
Select columns in order of DCA template

```{r}
hakon_condition <- hakon_conditions %>% 
  arrange(participantExternalId) %>% 
  mutate(studyCode = "X01-Hakon",
         eventId = NA,
         eventType = NA,
         ageAtConditionMeasureObservation = as.numeric(age_at_condition_or_measure_observation),
         conditionInterpretation = "Observed",
         conditionStatus = "History Of",
         conditionDataSource = "Clinical",
         measureValue = NA,
         measureUnit = NA) %>%
  select(all_of(condition_schemalabels)) %>% 
  setNames(condition_displaynames)
```

```{r}
write_csv(hakon_condition, here::here("output", "hakon_condition.csv"))
```

















