---
title: "Harmonize ABCDS clinical data & assemble file manifest, March 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```

Read in data
```{r}
# control & DS demographics
abcds_demo_ctl <- read_csv(synGet("syn33401283")$path)
abcds_demo_ds <- read_csv(synGet("syn33401291")$path)

# control & DS health histories
abcds_health_ctl <- read_csv(synGet("syn33401148")$path)
abcds_health_ds <- read_csv(synGet("syn33401163")$path)

# DS karyotypes
abcds_karyo_ds <- read_csv(synGet("syn33401175")$path)

# link between controls & DS siblings
abcds_control_linkage <- read_xlsx(synGet("syn34654889")$path)

# age at event
abcds_age_ctl <- read_csv(synGet("syn33401687")$path)
abcds_age_ds <- read_csv(synGet("syn33401689")$path) 
```


Combine DS and control subjects
```{r}
abcds_health_all <- rbind(abcds_health_ds, abcds_health_ctl)
abcds_demo_all <- rbind(abcds_demo_ds, abcds_demo_ctl)
abcds_age_all <- rbind(abcds_age_ctl, abcds_age_ds)
```


<!-- Data checks [from Oct 2022] -->
<!-- ```{r} -->
<!-- #417 unique subjects, no setdiff between health & demo -->
<!-- n_distinct(health_all$subject_label) -->
<!-- n_distinct(demo_all$subject_label) -->
<!-- setdiff(health_all$subject_label, demo_all$subject_label) -->

<!-- # most patients have both Baseline & Cycle 2 health & demo -->
<!-- health_all %>% count(subject_label) %>% arrange(n) -->
<!-- demo_all %>% count(subject_label) %>% arrange(desc(n)) -->

<!-- # the demo cols we're interested in are the same at both timepoints -->
<!-- demo_all %>% select(subject_label, de_gender, de_race, de_ethnicity, ds_vs_control_flag) %>%  -->
<!--   distinct() %>% count(subject_label) %>% arrange(desc(n)) -->

<!-- # there are changes in hh between the 2 timepoints - asked ABCDS team what to do (FOR NOW - use baseline data only) -->
<!-- health_all %>% select(subject_label, starts_with("hh")) %>%  -->
<!--   distinct() %>% count(subject_label) %>% arrange(desc(n)) -->

<!-- # controls - how many have DS sibs -->
<!-- n_distinct(demo_ctl$subject_label) -->
<!-- n_distinct(control_linkage$ABCDS_ID_SC) -->

<!-- # 10233 10224 10237 10235 are sibs in the linkage sheet who don't have any data -->
<!-- setdiff(control_linkage$ABCDS_ID_SC, demo_ctl$subject_label) -->
<!-- ``` -->


FOR NOW: limit to Baseline data only (still for March 2023)
```{r}
abcds_health_BL <- abcds_health_all %>% filter(event_code == "bl")
abcds_demo_BL <- abcds_demo_all %>% filter(event_code == "bl")
abcds_age_BL <- abcds_age_all %>% filter(event_code == "bl")

# setdiff(demo_BL$subject_label, health_BL$subject_label)
# 2 subjects, 23001 and 23002, don't have baseline health histories
```

##### PARTICIPANT


Get ETL plan
Special import of condition_codes_v3 with col type specification
  (having mixed numeric & char values causes error when unnesting)
```{r}
gs4_deauth()

abcds_etl_plan <- read_sheet("https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit?pli=1#gid=963585824",
           sheet = "etl_to_linkml_plan",
           col_types = "c") %>% 
  filter(study == "ABC-DS")
```


Bring in ABCDS_ID_DS from linkage table
Family ID = ABCDS_ID_DS if exists, otherwise = `NA`
```{r}
# filter out sibs 10233 10224 10237 10235 who don't have data
# DS patients with these sibs will show as Proband-only, no Family ID

#TODO: 10233 and 10235 DIDN'T get filtered out -> FIX NEXT TIME

sibs_with_data <- abcds_control_linkage %>% 
  filter(ABCDS_ID_SC %in% abcds_demo_ctl$subject_label) %>% 
  select(ABCDS_ID_SC, ABCDS_ID_DS)

abcds_demo_BL_harmonized <- abcds_demo_BL %>%  
  left_join(sibs_with_data,
            by = c("subject_label" = "ABCDS_ID_SC"), keep = TRUE) %>% 
  relocate(ABCDS_ID_DS, .after = subject_label) %>% 
  mutate(familyId = case_when(
    !is.na(ABCDS_ID_DS) ~ ABCDS_ID_DS,
    subject_label %in% sibs_with_data$ABCDS_ID_DS ~ subject_label
  ), .after = subject_label) %>% 
  arrange(subject_label)
```

Family Type:
"Proband-only" if family_id = NA
"Other" if not NA
**may want to add "Siblings" to Family Type in schema**

```{r}
abcds_demo_BL_harmonized <- abcds_demo_BL_harmonized %>% 
  mutate(familyType = if_else(is.na(familyId), "Proband-only", "Other")) 

#%>% select(subject_label, family_id, familyType, ds_vs_control_flag)
```

Family relationships:

```{r}
abcds_demo_BL_harmonized <- abcds_demo_BL_harmonized %>% 
  mutate(familyRelationship = case_when(
    ds_vs_control_flag == "DS" ~ "Proband",
    ds_vs_control_flag == "Control" ~ "Sibling"
  )) 

#%>% select(subject_label, family_id, ABCDS_ID_DS, ABCDS_ID_SC, familyType, ds_vs_control_flag, familyRelationship)
```

Family IDs

```{r}
sibling_ids <- abcds_control_linkage %>% 
  rename(subject_label = ABCDS_ID_DS,
         siblingId = ABCDS_ID_SC) %>% 
  select(subject_label, siblingId)

abcds_demo_BL_harmonized <- abcds_demo_BL_harmonized %>% 
  left_join(sibling_ids) 

#%>% select(subject_label, family_id, siblingId)
```

Sex, Race, Ethnicity, DS status

```{r}
abcds_sex_lookup_values <- abcds_etl_plan %>% 
  filter(model_slot == "sex") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>%
  mutate(original_value = as.double(original_value))

abcds_race_lookup_values <- abcds_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

abcds_ethnicity_lookup_values <- abcds_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>%
  mutate(original_value = as.double(original_value))

abcds_ds_lookup_values <- abcds_etl_plan %>% 
  filter(model_slot == "downSyndromeStatus") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

abcds_demo_BL_harmonized <- abcds_demo_BL_harmonized %>%
  # remove "|" from de_race
  #mutate(de_race = as.double(str_replace_all(de_race, "[^[:digit:]]", ""))) %>% 
  left_join(abcds_sex_lookup_values, by = c("de_gender" = "original_value")) %>% 
  rename("sex" = model_enum_label) %>% 
  left_join(abcds_race_lookup_values, by = c("de_race" = "original_value")) %>% 
  rename("race" = model_enum_label) %>%  
  left_join(abcds_ethnicity_lookup_values, by = c("de_ethnicity" = "original_value")) %>%   rename("ethnicity" = model_enum_label)%>%  
  left_join(abcds_ds_lookup_values, by = c("ds_vs_control_flag" = "original_value")) %>% 
  rename("downSyndromeStatus" = model_enum_label) 

#%>% select(subject_label, de_gender, sex, de_race, race, de_ethnicity, ethnicity, ds_vs_control_flag, downSyndromeStatus)

```

Age at baseline visit

```{r}
abcds_demo_BL_harmonized <- abcds_demo_BL_harmonized %>% 
  left_join(abcds_age_BL %>% select(subject_label, age_at_visit))
```

Get Global IDs:

```{r}
abcds_dewrangle_input <- abcds_demo_BL_harmonized %>% 
  distinct(subject_label) %>% 
  rename(descriptor = subject_label) %>% 
  mutate(fhir_resource_type = "patient")

write_csv(abcds_dewrangle_input, 
          here::here("output", "abcds_dewrangle_input.csv"))

# submit manifest at dewrangle.com; save result to "output"

abcds_global_ids <- read_csv(here::here("output", "global-descriptors-ABC-DS.csv"))

abcds_demo_BL_harmonized <- abcds_demo_BL_harmonized %>% 
  left_join(abcds_global_ids %>% select(descriptor, global_id),
            by = c("subject_label" = "descriptor")) 

#%>% select(subject_label, global_id) %>% filter(is.na(global_id))
```


Create final Participant table - Rename/create empty cols to match DCA template, has to be in order - For empty columns, replace NA with "Not provided"

```{r}
abcds_participant <- abcds_demo_BL_harmonized %>% 
  mutate(studyCode = "ABC-DS",
         participantGlobalId = global_id,
         participantExternalId = subject_label,
         fatherId = NA,
         motherId = NA,
         otherFamilyMemberId = NA,
         ageAtFirstPatientEngagement = round(age_at_visit*365.25, 0),
         firstPatientEngagementEvent = "Baseline visit",
         outcomesVitalStatus = NA,
         ageAtLastVitalStatus = NA) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)
```

```{r}
write_csv(abcds_participant, here::here("output", "abcds_participant.csv"))
```


##### CONDITION

Get condition codes (no Measures allowed for ABC-DS)
Special import of condition_codes_v3 with col type specification
  (having mixed numeric & char values causes error when unnesting)
```{r}
gs4_deauth()

condition_lookup_abcds <- read_sheet("https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit?pli=1#gid=384137116",
           sheet = "condition_codes_v3",
           col_types = "c") %>% 
  filter(study == "ABC-DS") %>% 
  mutate(source_column_value = na_if(source_column_value, "NA"))
```

- Remove columns that won't be harmonized (either for Sept, or ever)
ie. highlighted in grey or blue in `abcds_dic_harmonizing.xlsx`
- Pull in karyotype

**FIX LATER: 20 pts from Karyotype table do not have karyotype. `NA` is already in condition_codes_v3 as general DS -> find out why it doesn't get pulled in**
**23001 and 23002 not in karyotype table b/c they don't have BL data -> add manually**

```{r}
abcds_health_BL_harmonized <- abcds_health_BL %>% 
  select(-c(site_id, event_code, event_label, hh_cancer, hh_diabetes, hh_done, hh_fracture_5yr, hh_fracture_loc, ends_with("status"), ends_with("treated"), hh_hep_b_carrier, hh_hep_b_immune, hh_hep_b_vaccine, starts_with("hh_hrt_"), hh_menopausal, hh_menopausal_onset, hh_menopausal_status, hh_menstruated, hh_ndreason, update_stamp)) %>% 
  left_join(abcds_karyo_ds %>% select(subject_label, karyotype))
```

Get conditions:

Cols where NA matters (highlighted yellow in condition_codes_v3):
- if hh_anemia = 1 AND hh_anemia_type = NA -> "blank AND hh_anemia = 1"
- if hh_chd = 1 AND hh_chd_type = NA ~ hh_chd_type -> "blank AND hh_chd = 1"
    10065
    20441
    21306
    21381
    22768
Make all cols char so I can pivot longer
Pivot longer
split cells with | using separate_rows

```{r}
# first fix the special cases for hh_anemia and hh_chd
abcds_BL_conditions <- abcds_health_BL_harmonized %>%
  mutate(hh_chd_type = ifelse(hh_chd == 1 & is.na(hh_chd_type), "blank AND hh_chd = 1", hh_chd_type)) %>% 
  mutate(hh_anemia_type = ifelse(hh_anemia == 1 & is.na(hh_anemia_type), "blank AND hh_anemia = 1", hh_anemia_type)) %>% 
  select(-hh_chd, -hh_anemia) %>% 
  # then pivot longer
  mutate(across(everything(), as.character)) %>% 
  pivot_longer(cols = -subject_label,
               names_to = "condition",
               values_to = "status") %>% 
  # keep positive assertions
  filter(status != "0") %>% 
  # separate out multi-responses
  separate_rows(status, sep = "\\|") %>% 
  # look up annotations
  left_join(condition_lookup_abcds, by = c("condition" = "source_column", "status" = "source_column_value"))
```


Create empty/all same value columns
Select columns in order of DCA template
```{r}
abcds_condition <- abcds_BL_conditions %>% 
  mutate(studyCode = "ABC-DS",
         participantExternalId = subject_label,
         eventId = "Baseline",
         eventType = "Visit",
         ageAtConditionMeasureObservation = NA,
         conditionInterpretation = "Observed",
         conditionDataSource = "Clinical",
         conditionStatus = "History Of",
         measureValue = NA,
         measureUnit = NA) %>%
  select(all_of(condition_schemalabels)) %>% 
  setNames(condition_displaynames)
  
```

```{r}
write_csv(abcds_condition, here::here("output", "abcds_condition.csv"))
```



##### FILE MANIFEST

Read in subject_label columns from all datasets:

```{r}
#TODO: rewrite to get names from Synapse
filenames <- list.files(path = "C:/Users/pierrettelo/Documents/INCLUDE/data uploaded to Synapse/ABC-DS data from LONI/All Study Data", pattern = ".csv$", full.names = TRUE)

#41 files, subtract data dictionary = 40 files.. don't forget 1 has ABCDS instead of abcds

subject_labels <- map_df(filenames, read_csv, col_select = matches("subject_label", ignore.case = TRUE), id = "filename")

abcds_datafile_manifest <- subject_labels %>% 
  mutate(filename = str_extract(filename, "(abcds|ABCDS).*(?=.csv)")) %>% 
  pivot_longer(cols = -filename, names_to = "colname", values_to = "subject_label") %>% 
  filter(!is.na(subject_label)) %>%
  select(subject_label, filename) %>% 
  arrange(subject_label) %>% 
  distinct() #many files have >1 row per subject

#check = 417 participants
#pull(abcds_datafile_manifest, subject_label) %>% unique()
```

Read in Expansion sheet

```{r}
gs4_deauth()

abcds_file_metadata <- read_sheet(
  "https://docs.google.com/spreadsheets/d/1wNtmceBQwZu71gmQ1Z48yqNdV0zNu9WfZJ5IAhoRnMw/edit#gid=500276219&fvid=2074659656",
  sheet = "ABCDS_file_metadata")

abcds_datafile <- abcds_datafile_manifest %>% 
  left_join(abcds_file_metadata, by = c("filename" = "file_name")) %>% 
  rename(participantExternalId = subject_label,
         fileExternalId = filename,
         fileName = file_title,
         dataCategory = data_category,
         dataType = data_type,
         fileFormat = format) %>% 
  mutate(studyCode = "ABC-DS",
         sampleExternalId = "Not Available",
         fileGlobalId = NA,
         fileHash = NA,
         dataAccess = "Controlled",
         experimentalStrategy = NA,
         experimentalPlatform = NA,
         fileSize = NA,
         fileSizeUnit = NA,
         accessUrl = "https://www.nia.nih.gov/research/abc-ds#data",
         publicationDoi = NA) %>% 
  select(all_of(datafile_schemalabels)) %>% 
  setNames(datafile_displaynames)
```


```{r}
abcds_datafile %>% write_csv(here::here("output", "abcds_datafile.csv"))
```
































