---
title: "Harmonize Khor to LinkML v2 model - March 2023"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```


Read in data:
```{r}
khor_clinical_source <- read_xlsx(synGet("syn32937504")$path)
```

Replace headers with working_headers
See Google sheet `INCLUDE clinical data dictionaries` -> `BRI-DSR`
```{r}
gs4_deauth()
khor_working_headers <- read_sheet(
  "https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw/edit#gid=0",
  sheet = "BRI-DSR")

khor_clinical_source <- setNames(khor_clinical_source, khor_working_headers$working_header)
```

Get ETL plan
```{r}
# source("harmonization_setup_2023-03.R")

khor_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "BRI-DSR")
```

Participant External ID
```{r}
khor_clinical_harmonized <- khor_clinical_source %>% 
  rename(participantExternalId = bri_include_id)
```


## PARTICIPANT

Columns that need to be transformed:

Sex = needs initial cap
Race
- White, Caucasian = White
- Black, African American = Black or African American
- American Indian, Alaska Native = American Indian or Alaska Native
- Do Not Know = Unknown
- NA = Unknown
- Other Race = Other
- everything with comma = More than one race
Ethnicity

```{r}
khor_sex_lookup_values <- khor_etl_plan %>% 
  filter(model_slot == "sex") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value)

khor_race_lookup_values <- khor_etl_plan %>% 
  filter(model_slot == "race") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>% 
  na_if("NA")

khor_eth_lookup_values <- khor_etl_plan %>% 
  filter(model_slot == "ethnicity") %>% 
  select(model_enum_label, original_value) %>% 
  unnest(original_value) %>% 
  na_if("NA")

khor_participant <- khor_clinical_harmonized %>% 
  left_join(khor_sex_lookup_values, by = c("gender" = "original_value")) %>% 
  rename("sex" = model_enum_label) %>% 
  left_join(khor_race_lookup_values, by = c("race_original" = "original_value")) %>% 
  rename("race" = model_enum_label) %>%  
  left_join(khor_eth_lookup_values, by = c("ethnicity_original" = "original_value")) %>% 
  rename("ethnicity" = model_enum_label) 

# %>% select(participantExternalId, gender, sex, race_original, race, ethnicity_original, ethnicity)
```

Create final Participant table:

- Fill in info that is same for every participant (mutate)
- Rename columns that need renamed with schema labels (also mutate)
- select all Participant columns
- rename with display names

```{r}
bridsr_participant <- khor_participant %>% 
  mutate(studyCode = "BRI-DSR",
         participantGlobalId = NA,
         familyId = NA,
         familyType = "Proband-only",
         fatherId = NA,
         motherId = NA,
         siblingId = NA,
         otherFamilyMemberId = NA,
         familyRelationship = NA,
         downSyndromeStatus = "T21",
         ageAtFirstPatientEngagement = round((age_at_enrollment_years * 365.25), 0),
         firstPatientEngagementEvent = "Enrollment",
         outcomesVitalStatus = NA,
         ageAtLastVitalStatus = NA) %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames)

```


```{r}
write_csv(bridsr_participant, here::here("output", "bridsr_participant.csv"))
```


##### CONDITION

Get Khor conditions & measures
```{r}
condition_lookup_khor <- condition_codes_v3 %>% filter(study == "BRI-DSR")
```

Add column "DS_Diagnosis" = Down syndrome
Pull conditions from clinical data
Look up annotations
```{r}
khor_condition <- khor_clinical_harmonized %>% 
  mutate(ds_diagnosis = "Down syndrome") %>% 
  select(-c(patient_type:ethnicity_original)) %>% 
  pivot_longer(cols = c(-participantExternalId), 
               names_to = "source_text",
               values_to = "status") %>% 
  filter(!is.na(status)) %>% 
  left_join(condition_lookup_khor, by = c("source_text" = "source_column")) %>% 
  select(-study, -source_column_value)
```

Create empty/all same value columns
Select columns in order of DCA template
```{r}
bridsr_condition <- khor_condition %>% 
  mutate(eventId = NA,
         eventType = NA,
         ageAtConditionMeasureObservation = NA,
         conditionInterpretation = "Observed",
         conditionStatus = "History Of",
         conditionDataSource = "Clinical",         
         measureValue = NA,
         measureUnit = NA) %>% 
  select(all_of(condition_schemalabels)) %>% 
  setNames(condition_displaynames)
```

```{r}
write_csv(bridsr_condition, here::here("output", "bridsr_condition.csv"))
```






