---
title: "v3 fixes to all_cohorts files"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```

##### HTP: Swap out my global IDs for KF IDs

```{r}
# get all_cohorts_participant
all_cohorts_participant <- read_csv(synGet("syn51216460")$path)


# download latest Global IDs from Dewrangle
# curated with original KFIDs by Christina & Marcel
new_global_ids <- read_csv(here::here("output", "global-descriptors-htp-WITH-KFIDS.csv"))

new_global_ids %>% janitor::get_dupes()
```


Participant IDs Changelog:
```{r}
all_cohorts_participant %>% 
  filter(`Study Code` == "HTP") %>% 
  select(`Study Code`, `Participant External ID`, `Participant Global ID`) %>% 
  left_join(new_global_ids %>% select(descriptor, new_global_id = global_id),
            by = c("Participant External ID" = "descriptor")) %>% 
  mutate(id_changed = (`Participant Global ID` != new_global_id)) %>% 
  write_csv(here::here("output", "htp_participants_id_changelog.csv"))

# 535 changed, 165 not changed
```


Change in all_cohorts_participant
#### ALSO change X01-Hakon to X01-Hakonarson
```{r}
all_cohorts_participant %>% 
  left_join(new_global_ids %>% select(descriptor, new_global_id = global_id),
            by = c("Participant External ID" = "descriptor")) %>% 
  relocate(new_global_id, .after = `Participant Global ID`) %>%
  mutate(updated_global_id = ifelse(!is.na(new_global_id), new_global_id, `Participant Global ID`), .after = new_global_id) %>% 
  select(-`Participant Global ID`, -new_global_id) %>% 
  rename(`Participant Global ID` = updated_global_id) %>% 
  mutate(`Study Code` = str_replace(`Study Code`, "X01-Hakon", "X01-Hakonarson")) %>% 
  write_csv(here::here("output", "all_cohorts_participant_2023-05-17.csv"))

```


#### Change X01-Hakon in Conditions & Study
```{r}
all_cohorts_condition <- read_csv(synGet("syn51216464")$path)
all_cohorts_study <- read_csv(synGet("syn51216461")$path)

```


```{r}
all_cohorts_study %>% 
  mutate(`Study Code` = str_replace(`Study Code`, "X01-Hakon", "X01-Hakonarson")) %>% 
  write_csv(here::here("output", "all_cohorts_study_2023-05-17.csv"))

all_cohorts_condition %>% 
  mutate(`Study Code` = str_replace(`Study Code`, "X01-Hakon", "X01-Hakonarson")) %>% 
  write_csv(here::here("output", "all_cohorts_condition_2023-05-17.csv"))
```


Specimen IDs changelog:
```{r}
# get Christina's manifest with samples and collection samples
FV_NGY8HVR3_htp_combined_samples_with_globals <- read_csv("C:/Users/pierrettelo/Documents/INCLUDE/FV_NGY8HVR3_htp-combined-samples-with-globals.csv")

FV_NGY8HVR3_htp_combined_samples_with_globals %>% 
  select(`Participant External ID`, `Sample External ID`, `Sample Global ID`) %>% 
  left_join(new_global_ids %>% select(descriptor, new_global_id = global_id),
            by = c("Sample External ID" = "descriptor")) %>% 
  mutate(id_changed = (`Sample Global ID` != new_global_id)) %>% 
  write_csv(here::here("output", "htp_samples_AND_COLLECTIONS_id_changelog.csv"))
```


Change the sample manifest:
```{r}
FV_NGY8HVR3_htp_combined_samples_with_globals %>% 
  left_join(new_global_ids %>% select(descriptor, new_global_id = global_id),
            by = c("Sample External ID" = "descriptor")) %>% 
  relocate(new_global_id, .after = `Sample Global ID`) %>%
  select(-`Sample Global ID`) %>% 
  rename(`Sample Global ID` = new_global_id) %>% 
  write_csv(here::here("output", "htp_samples_AND_COLLECTIONS_with_KF_global_ids_2023-05-18.csv"))
```


#### DSC - add MONDO codes for 6 participants who left ds_specific_diagnosis blank

```{r}
# get all_cohorts_condition
all_cohorts_condition <- read_csv(synGet("syn51216464")$path)

# get dsc source data
dsc_ihq <- read_csv(synGet("syn26445517")$path)

```

```{r}
six_without_ds_diagnosis <- c("11296", "12481", "12633", "12657", "12721", "13056")

dsc_ihq %>% 
  filter(patient_id %in% six_without_ds_diagnosis) %>%
  mutate(`Study Code` = "DSC",
         `Participant External ID` = as.character(patient_id),
         `Event ID` = as.character(session_id),
         `Event Type` = "Survey",
         `Condition or Measure Source Text` = "Down syndrome - specific diagnosis not provided",
         `Age At Condition or Measure Observation` = NA,
         `Condition Interpretation` = "Observed",
         `Condition Status` = "History Of",
         `Condition Data Source` = "Self-reported",
         `HPO Label` = NA,
         `HPO Code` = NA,
         `MONDO Label` = "Down syndrome",
         `MONDO Code` = "MONDO:0008608",
         `MAXO Label` = NA,
         `MAXO Code` = NA,
         `Other Label` = NA,
         `Other Code` = NA,
         `Measure Value` = NA,
         `Measure Unit` = NA) %>% 
  select(-contains("_")) %>% 
  bind_rows(all_cohorts_condition) %>%
  write_csv(here::here("output", "all_cohorts_condition_2023-05-25.csv"))

```


### ABCDS - add MONDO codes for 22 participants who didn't have karyotype data

```{r}
all_cohorts_participant <- read_csv(synGet("syn51216460")$path)
all_cohorts_condition <- read_csv(synGet("syn51216464")$path)


ABCDS_D21_QA <- c("pt-23rgqn89ft", "pt-2fmw236n3z", "pt-2q9hncqgad", "pt-4d2b3khehd", "pt-4xjepzbmqn", "pt-6ewrk9xr2r", "pt-6mexz2gge2", "pt-6mkrmp3c8h", "pt-73g2kx64x7", "pt-7ek927e2b7", "pt-7j96cxf8fg", "pt-8e6dfm2tf7", "pt-8jgyene3ke", "pt-8mfmexxgjd", "pt-8qdrjm4kyq", "pt-98hcfxgrtf", "pt-9txqfbejqe", "pt-ahipt2e798", "pt-ax7xfdej6e", "pt-baf7hp4jcg", "pt-bd4wfkjrec", "pt-c2b4pq3jxq", "pt-cmwmexc87r", "pt-cx6wxe3zzr", "pt-dbmkfxpmnk", "pt-dddjgwn4tz", "pt-dg7zyh89kj", "pt-dgk3qj4zgf", "pt-dqri3qt3h6", "pt-dt7wptjkaf", "pt-e2pi3i6yi8", "pt-ecqe829gdt", "pt-epn8b2piek", "pt-eqxi2d98t6", "pt-ewhemibzbi", "pt-f2y3mhtdmz", "pt-f9jg62jb9j", "pt-gaq7fphiyc", "pt-gc8dnddzif", "pt-h6cdhxe4zf", "pt-hc8edp8a8c", "pt-ibeajp8dfg", "pt-jiqnpnmmdk", "pt-jjqizpzc9p", "pt-mhz7whk24i", "pt-miwfxq4qze", "pt-mmhyzc3xyh", "pt-mwc8n9k2jq", "pt-nf62qqn3wr", "pt-p6iypxr84z", "pt-pjj83a6ce9", "pt-r3wj62az92", "pt-rdnbkptfck", "pt-rwj6mxzgqa", "pt-rytnyybn4z", "pt-t2kfwdd7qb", "pt-t98yjex2tw", "pt-tgifcz78ty", "pt-tig7e9kbbx", "pt-trbdq8bidw", "pt-tyknt83e28", "pt-xpt274y7ht", "pt-zhj73xfhyt")

ABCDS_wrong_external <- all_cohorts_participant %>% 
  filter(`Participant Global ID` %in% ABCDS_D21_QA) %>% 
  filter(`Down Syndrome Status` == "T21") %>% 
  pull(`Participant External ID`)

all_cohorts_condition %>% 
  filter(`Participant External ID` %in% ABCDS_wrong_external) %>% view()

# these are 20 participants without karyotype and 2 (23001, 23002) who don't have any BL data (and therefore also don't have karyotype)
```


Add DS diagnosis to all_cohorts_condition:

```{r}
all_cohorts_condition %>% 
  add_row(`Participant External ID` = ABCDS_wrong_external, .before = 1) %>% 
  filter(is.na(`Study Code`)) %>% 
  mutate(`Study Code` = "ABC-DS",
         `Participant External ID` = `Participant External ID`,
         `Event ID` = NA,
         `Event Type` = NA,
         `Condition or Measure Source Text` = "Down syndrome - karyotype not provided",
         `Age At Condition or Measure Observation` = NA,
         `Condition Interpretation` = "Observed",
         `Condition Status` = "History Of",
         `Condition Data Source` = "Clinical",
         `HPO Label` = NA,
         `HPO Code` = NA,
         `MONDO Label` = "Down syndrome",
         `MONDO Code` = "MONDO:0008608",
         `MAXO Label` = NA,
         `MAXO Code` = NA,
         `Other Label` = NA,
         `Other Code` = NA,
         `Measure Value` = NA,
         `Measure Unit` = NA) %>% 
  bind_rows(all_cohorts_condition) %>% 
  write_csv(here::here("output", "all_cohorts_condition_2023-05-26.csv"))

```


##### HTP: Checking if I have clinical data for the 15 missing P4C samples

```{r}
P4C_15_samples <- c("HTP0084A2_RNA_paxgene", "HTP0085B2_RNA_paxgene", "HTP0204A_RNA_paxgene", "HTP0206B_RNA_paxgene", "HTP0288B_RNA_paxgene", "HTP0302B_RNA_paxgene", "HTP0310A_RNA_paxgene", "HTP0322B_RNA_paxgene", "HTP0326A_RNA_paxgene", "HTP0328A_RNA_paxgene", "HTP0332A_RNA_paxgene", "HTP0338B_RNA_paxgene", "HTP0348A_RNA_paxgene", "HTP0351A_RNA_paxgene", "HTP0442A2_RNA_paxgene") %>% str_extract("^.*[:digit:]{4}")

all_cohorts_participant %>% 
  filter(`Participant External ID` %in% P4C_15_samples)

# 15 rows -> all clinical data present
```

#### Updating CDE numbers

```{r}
all_cohorts_participant %>% 
  filter(`Study Code` == "HTP" & Sex == "Male")
```

```{r}
all_cohorts_condition %>% 
  filter(`Study Code` == "DSC" & str_detect(`Condition or Measure Source Text`, "alopecia"))
```

##### check for uniqueness of Global IDs, then lowercase all

```{r}
all_cohorts_participant %>% 
  pull(`Participant Global ID`) %>%
  #tolower() %>% 
  n_distinct()

all_cohorts_participant %>% 
  filter(str_detect(tolower(`Participant Global ID`), tolower("VVECSZ13")))
```

```{r}
all_cohorts_participant %>% 
  mutate(`Participant Global ID` = tolower(`Participant Global ID`)) %>% 
  write_csv(here::here("output", "all_cohorts_participant_2023-06-08.csv"))

```

#### Add Global IDs to HTP omics metadata

```{r}
htp_omics_metadata <- read_csv("C:/Users/pierrettelo/Documents/INCLUDE/Biospecimen-FileManifests-pre-v3/Biospecimen-FileManifests-pre-v3/HTP/updated-kf-omics.csv")
```


```{r}
htp_global_ids <- read_csv(here::here("output", "global-descriptors-htp_2023-06-20.csv")) %>% 
  filter(fhir_resource_type == "Specimen") %>% 
  select(descriptor, global_id) %>% 
  arrange(descriptor)
```


```{r}
htp_omics_metadata %>% 
  left_join(htp_global_ids, by = c("sample_id" = "descriptor")) %>% 
  relocate(global_id, .after = sample_id) %>% 
  arrange(sample_id) %>% 
  write_csv(here::here("output", "htp_omics_metadata_with_global_ids_2023-06-20.csv"))
```

