---
title: "DS-Connect prep unharmonized data"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)
library(janitor)

synLogin()

gs4_deauth()
```


### 1. Participant IDs & GUIDs

- quick check in Excel - any weird looking or missing IDs?
- check for dupes
- compare participants in both datasets (demographic vs IHQ) - report if differences
- report total N unique across dataset
- compare with v3 - report added/deleted
- add dewrangle IDs/get IDs for new participants
- assemble mapping & guids file: DSC ID + de-ID ID + global ID + guid

Any issues, record in Cohort-specific QA sheet (https://docs.google.com/spreadsheets/d/1uxFOl-SDpLRPaeIkfpqnksKMyeoK9pNqBDC33OEoBIs/edit#gid=190853806) 
and follow up with contributor

===============================================================================

Get data
```{r}
dsc_demographic <- read_csv(synGet("syn26445516")$path)
dsc_ihq <- read_csv(synGet("syn26445517")$path)

#v3 harmonized data
v3_dsc_harmonized <- read_csv(synGet("syn51216460")$path) %>% 
  filter(`Study Code` == "DSC")
```

##**IMPORTANT**

- deid IDs from Invitae overlap with original patient_ids
- Fix by adding "DSC" in front of all deid IDs (also parent_id_deid in IHQ)
- Filter out Acknowledgments line; remove GUIDs for DMC version
- Then re-upload to Synapse as new version of source data
(versions with GUIDs should be a separate file with restricted access)
- Also add to data dictionary (with Excel) and reupload to Synapse
- Be sure to send these new deid IDs (and the global IDs) to Invitae replacement

```{r}
dsc_demographic <- dsc_demographic %>%
  filter(!str_detect(patient_id, "ACK")) %>% 
  mutate(patient_id_deid = paste0("DSC", patient_id_deid))

dsc_ihq <- dsc_ihq %>% 
  filter(!str_detect(patient_id, "ACK")) %>% 
  mutate(patient_id_deid = paste0("DSC", patient_id_deid)) %>%
  mutate(parent_id_deid = case_when(!is.na(parent_id_deid) ~ paste0("DSC", parent_id_deid),
                                    TRUE ~ NA))

# write out without guids; leave in for rest of this notebook
dsc_demographic %>%
  select(-guid) %>% 
  write_csv(here::here("output",                     "dsconnect_demographics_to_dcc_20240226_PLdeid_noguid.csv"))

dsc_ihq %>%
  select(-guid) %>% 
  write_csv(here::here("output",
                       "dsconnect_ihq_survey_20240226_PLdeid_noguid.csv"))
```


Isolate IDs
```{r}
dsc_demo_ids <- dsc_demographic %>% select(patient_id, patient_id_deid, guid)
dsc_ihq_ids <- dsc_ihq %>% select(patient_id, patient_id_deid, guid)

v3_dsc_ids <- v3_dsc_harmonized %>% 
  select(participant_global_id = `Participant Global ID`, 
         participant_external_id = `Participant External ID`)
```

Check for dupes
- Patient 5458 has 2 rows in demo & ihq that appear to be identical
- There are 12 duplicate GUIDs (not including 5458)
- 181 rows where GUID is blank or #ERROR#
```{r}
get_dupes(dsc_demo_ids, patient_id) #5458
get_dupes(dsc_ihq_ids, patient_id)

get_dupes(dsc_demo_ids, patient_id_deid)
get_dupes(dsc_ihq_ids, patient_id_deid)

dsc_demo_ids %>% filter(str_detect(guid, "NDAR")) %>% get_dupes(guid)
dsc_ihq_ids %>% filter(str_detect(guid, "NDAR")) %>% get_dupes(guid)
# 7 but one is pt 5458, so 6

dsc_demo_ids %>% filter(!str_detect(guid, "NDAR") | is.na(guid)) #181

get_dupes(v3_dsc_ids, participant_global_id) #0
```

Compare IDs between demo/ihq and v3
- no differences in demo/ihq for original or de-id IDs
```{r}
setdiff(dsc_demo_ids$patient_id, dsc_ihq_ids$patient_id)
setdiff(dsc_ihq_ids$patient_id, dsc_demo_ids$patient_id)

setdiff(dsc_demo_ids$patient_id_deid, dsc_ihq_ids$patient_id_deid)
setdiff(dsc_ihq_ids$patient_id_deid, dsc_demo_ids$patient_id_deid)
```


Set up table of unique IDs 
(filters out duplicate #5458 rows for now - delete in data later)
```{r}
dsc_id_guid_map <- dsc_demo_ids %>% unique()
```

Compare with v3
- 9 patients deleted from v3
- 159 added
- N unique in v5: 3634

```{r}
#deleted from v3 (use mapping file with unique IDs)
setdiff(v3_dsc_ids$participant_external_id, dsc_id_guid_map$patient_id) #9

#added to latest export
length(setdiff(dsc_id_guid_map$patient_id, v3_dsc_ids$participant_external_id))

n_distinct(dsc_id_guid_map$patient_id) #3634
```


Add existing Global IDs
```{r}
dsc_id_guid_map <- dsc_id_guid_map %>% 
  left_join(v3_dsc_ids, by = c("patient_id" = "participant_external_id"))
```

Check v3 against global IDs from dewrangle, just in case
Download "newest descriptor" from dewrangle.com to "data" folder
```{r}
dsc_dewrangle_output <- read_csv(here::here("data", "global-descriptors-ds-connect.csv"))

setdiff(v3_dsc_ids$participant_global_id, dsc_dewrangle_output$globalId) #0
setdiff(dsc_dewrangle_output$globalId, v3_dsc_ids$participant_global_id)
#just the Study ID sd-ickkyna4w9
```

Prepare dewrangle input
1. upload ORIGINAL patient_id for 159 new participants
(so every participant has original ID associated with global ID)
2. upload global_id + patient_id_deid for all participants
(so every participant has original ID as old descriptor and patient_id_deid
as new descriptor) -> use new descriptors going forward
```{r}
dsc_id_guid_map %>% 
  filter(is.na(participant_global_id)) %>% 
  select(descriptor = patient_id) %>% 
  mutate(fhirResourceType = "Patient") %>% 
  write_csv(here::here("output",
                       "dsc_dewrangle_input_original.csv"))

```


Add new Global IDs
generate new dewrangle input with patient_id_deid as descriptor
```{r}
dsc_dewrangle_output_v2 <- read_csv(here::here("data", "global-descriptors-ds-connect.csv"))

dsc_id_guid_map %>%
  left_join(dsc_dewrangle_output_v2 %>% select(descriptor, globalId),
            by = c("patient_id" = "descriptor")) %>% 
  mutate(fhirResourceType = "Patient") %>% 
  select(descriptor = patient_id_deid, fhirResourceType, globalId) %>% 
  write_csv(here::here("output",
                       "dsc_dewrangle_input_deid.csv"))
  
```

Check new vs old descriptors:
```{r}
dsc_dewrangle_output_all <- read_csv(here::here("data", "global-descriptors-ds-connect-all.csv")) %>% 
  filter(fhirResourceType == "Patient")

old_descriptors <- dsc_dewrangle_output_all %>% 
  filter(!str_detect(descriptor, "DSC")) %>% 
  select(descriptor, globalId)
#3484+159=3643

new_descriptors <- dsc_dewrangle_output_all %>% 
  filter(str_detect(descriptor, "DSC")) %>% 
  select(descriptor, globalId) #3634

setdiff(old_descriptors$globalId, new_descriptors$globalId) #9
setdiff(new_descriptors$globalId, old_descriptors$globalId) #0
```


Finalize DSC ID map:
```{r}
dsc_id_guid_map_complete <- dsc_id_guid_map %>%
  left_join(dsc_dewrangle_output_v2 %>% select(descriptor, globalId),
            by = c("patient_id" = "descriptor")) %>% 
  select(-participant_global_id) %>% 
  select(participant_external_id = patient_id_deid,
         guid, participant_global_id = globalId)
```


Write out & upload to Synapse
```{r}
dsc_id_guid_map_complete %>% 
  write_csv(here::here("output", "dsc_guids_2024-03-04.csv"))
```

#### UNHARMONIZED DATA

Convo w/ Sujata, 2024-04-11:
- Leave in birth year (even though redundant)
- Remove survey timestamp (full dates not allowed per DCC policy)
- Delete zip code (unsure if Invitae did any cleaning of areas with <20k people;
still useful with just state & country)
- delete GUID from demo & IHQ

ID usage, per email thread 2024-04-24:

- Data Hub UI: De-identified Patient IDs
- Unharmonized data download: Original and de-identified Patient IDs
- GUID mapping file: De-identified IDs (to match portal) and GUIDs

**add these decisions to file wiki when unharmonized file is on Synapse**

```{r}
dsc_demo_unharmonized <- dsc_demographic %>% 
  select(-address_postal_code)

dsc_demo_unharmonized %>% 
  write_csv(here::here("output",
                       "dsc_demographic_unharmonized_2024-02-26.csv"))
```

```{r}
dsc_ihq_unharmonized <- dsc_ihq %>%
  select(-survey_time)

dsc_ihq_unharmonized %>% 
  write_csv(here::here("output",
                       "dsc_ihq_unharmonized_2024-02-26.csv"))
```

**indicate above fields were REDACTED in dictionary; combine dicts into 1 file**

Check fields in data vs dictionary:

```{r}
dsc_demo_dict_fields <- read_xlsx(synGet("syn58614987")$path, sheet = 1, skip = 2) %>% 
  filter(!str_detect(Field, "REDACTED")) %>% 
  pull(Field)

dsc_ihq_dict_fields <- read_xlsx(synGet("syn58614987")$path, sheet = 2, skip = 2) %>% 
  filter(!str_detect(Field, "REDACTED")) %>% 
  pull(Field)
```

```{r}
setdiff(colnames(dsc_demo_unharmonized), dsc_demo_dict_fields)
# corrected dictionary from register_year to registration_year
setdiff(dsc_demo_dict_fields, colnames(dsc_demo_unharmonized))

setdiff(colnames(dsc_ihq_unharmonized), dsc_ihq_dict_fields)
setdiff(dsc_ihq_dict_fields, colnames(dsc_ihq_unharmonized))
```









