---
title: "DS-Connect prep unharmonized data"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)
library(janitor)

synLogin()

gs4_deauth()
```


### 1. Participant IDs

- quick check in Excel - any weird looking or missing IDs?
- check for dupes
- compare participants in both datasets (demographic vs IHQ) - report if differences
- report total N unique across dataset
- compare with v3 - report added/deleted
- add dewrangle IDs/get IDs for new participants
- assemble mapping & guids file: DSC ID + de-ID ID + global ID + guid

Any issues, record in Cohort-specific QA sheet (https://docs.google.com/spreadsheets/d/1uxFOl-SDpLRPaeIkfpqnksKMyeoK9pNqBDC33OEoBIs/edit#gid=190853806) 
and follow up with contributor

===============================================================================

Get data
- Filter out Acknowledgments line for now (will delete in the dataset for unharmonized data)
```{r}
dsc_demographic <- read_csv(synGet("syn26445516")$path) %>% 
  filter(!str_detect(patient_id, "ACK"))
dsc_ihq <- read_csv(synGet("syn26445517")$path) %>% 
  filter(!str_detect(patient_id, "ACK"))

#v3 harmonized data
v3_dsc_harmonized <- read_csv(synGet("syn51216460")$path) %>% 
  filter(`Study Code` == "DSC")
```


Isolate IDs
```{r}
dsc_demo_ids <- dsc_demographic %>% select(patient_id, patient_id_deid, guid)
dsc_ihq_ids <- dsc_ihq %>% select(patient_id, patient_id_deid, guid)

v3_dsc_ids <- v3_dsc_harmonized %>% 
  select(participant_global_id = `Participant Global ID`, 
         participant_external_id = `Participant External ID`)
```


Check for dupes
- Patient 5458 has 2 rows in demo & ihq that appear to be identical
- There are 12 duplicate GUIDs (not including 5458)
- 181 rows where GUID is blank or #ERROR#
```{r}
get_dupes(dsc_demo_ids, patient_id) #5458
get_dupes(dsc_ihq_ids, patient_id)

get_dupes(dsc_demo_ids, patient_id_deid)
get_dupes(dsc_ihq_ids, patient_id_deid)

dsc_demo_ids %>% filter(str_detect(guid, "NDAR")) %>% get_dupes(guid)
dsc_ihq_ids %>% filter(str_detect(guid, "NDAR")) %>% get_dupes(guid)

dsc_demo_ids %>% filter(!str_detect(guid, "NDAR") | is.na(guid)) #181

get_dupes(v3_dsc_ids, participant_global_id) 
```

Compare IDs between demo/ihq and v3
- no differences in demo/ihq
```{r}
setdiff(dsc_demo_ids$patient_id, dsc_ihq_ids$patient_id)
setdiff(dsc_ihq_ids$patient_id, dsc_demo_ids$patient_id)
```


Set up table of unique IDs 
(filters out duplicate #5458 rows for now - delete in data later)
```{r}
dsc_id_guid_map <- dsc_demo_ids %>% unique()
```

Check for patient_id_deid that are also in patient_id 
(would not have converse b/c none of the patient_id_deid have 4 digits)
```{r}
overlap <- dsc_id_guid_map %>% 
  filter(patient_id_deid %in% dsc_id_guid_map$patient_id) %>% 
  pull(patient_id_deid)

dsc_id_guid_map %>% 
  filter(patient_id %in% overlap)
#92
#NOOOO
```



Compare with v3
- 9 patients deleted from v3
- 159 added
- N unique in v5: 3634

```{r}
#deleted from v3 (use mapping file with unique IDs)
setdiff(v3_dsc_ids$participant_external_id, dsc_id_guid_map$patient_id) #9

#added to latest export
length(setdiff(dsc_id_guid_map$patient_id, v3_dsc_ids$participant_external_id))

n_distinct(dsc_id_guid_map$patient_id) #3634
```


Add existing Global IDs
```{r}
dsc_id_guid_map <- dsc_id_guid_map %>% 
  left_join(v3_dsc_ids, by = c("patient_id" = "participant_external_id"))
```

Check v3 against global IDs from dewrangle, just in case
Download "newest descriptor" from dewrangle.com to "data" folder
```{r}
dsc_dewrangle_output <- read_csv(here::here("data", "global-descriptors-ds-connect.csv"))

setdiff(v3_dsc_ids$participant_global_id, dsc_dewrangle_output$globalId) #0
setdiff(dsc_dewrangle_output$globalId, v3_dsc_ids$participant_global_id)
#just the Study ID
```

Prepare dewrangle input
1. upload ORIGINAL patient_id for 159 new participants
(so every participant has original ID associated with global ID)
2. upload global_id + patient_id_deid for all participants
(so every participant has original ID as old descriptor and patient_id_deid
as new descriptor) -> use new descriptors going forward
```{r}
dsc_dewrangle_input_1 <- dsc_id_guid_map %>% 
  filter(is.na(participant_global_id)) %>% 
  select(descriptor = patient)

```

##TODO: Might need to go back and re-generate patient_id_deid before further dewrangling
For now, upload dsc_id_guid_map to Synapse as DRAFT
```{r}
dsc_id_guid_map %>% 
  write_csv(here::here("output", "dsc_guids_DRAFT_2024-02-28.csv"))
```

UGH also duplicate GUIDS!!
```{r}
dsc_ihq_ids %>% filter(str_detect(guid, "NDAR")) %>% get_dupes(guid)

```




















