---
title: "GUID mapping"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)
library(janitor)

synLogin()

gs4_deauth()
```

Get GUIDs:
**NOTE** DSC GUIDs file is complete now - 
use this, not v3 all_cohorts_participant as some participants have been added/deleted
Latest HTP GUIDs file from Matt (2024-08-30) now includes all HTP participants
```{r}
#replace #Error# with NA
dsc_guids <- read_csv(synGet("syn53708715")$path) %>%
  mutate(guid = na_if(guid, "#ERROR#"),
         study_code = "DSC") %>% 
  select(guid, study_code, participant_global_id)

htp_guids <- read_csv(synGet("syn53707355")$path) %>% 
  select(participant_global_id = globalId,
         guid = GUID)
```

Check for dupes:
```{r}
get_dupes(htp_guids, guid) %>%
  filter(str_detect(guid, "NDAR")) #0
get_dupes(htp_guids, participant_global_id) #0

get_dupes(dsc_guids, guid) %>%
  filter(str_detect(guid, "NDAR")) #6
get_dupes(dsc_guids, participant_global_id)
```

Bind DSC/HTP
```{r}
htp_dsc_guids <- htp_guids %>% 
  mutate(study_code = "HTP", .before = everything()) %>% 
  bind_rows(dsc_guids)
```

Counts:
- Total HTP participants: 1055
- Total DSC participants: 3634
- HTP participants missing GUIDs: 520
- DSC participants missing GUIDs: 181
- 52 GUIDs in both DSC and HTP

```{r}
htp_dsc_guids %>% 
  filter(study_code == "DSC" & is.na(guid))

htp_dsc_guids %>%
  filter(!is.na(guid)) %>% 
  group_by(guid) %>% 
  #guids with both DSC and HTP study code
  filter(min(study_code) == "DSC" & max(study_code) == "HTP") %>% 
  #just to check if one of each study
  add_count(guid) %>% 
  arrange(desc(n), guid) %>% 
  pull(guid) %>% 
  n_distinct()

```

Write out & upload to Synapse
```{r}
htp_dsc_guids %>% 
  arrange(study_code) %>% 
  write_csv(here::here("output", "htp_dsc_guids_2024-08-30.csv"))
```



### check Matt's dewrangle IDs

```{r}
dewrangle <- read_csv(here::here("data", "global-descriptors-htp-all_2024-08-29.csv")) %>% 
  filter(fhirResourceType == "Patient") %>% 
  select(descriptor, globalId)

setdiff(dewrangle$globalId, htp_guids$participant_global_id)
setdiff(htp_guids$participant_global_id, dewrangle$globalId)

```



