---
title: "GUID mapping"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)
library(janitor)

synLogin()

gs4_deauth()
```

Get GUIDs:
**NOTE** DSC GUIDs file is DRAFT for now - need new deid IDs in order to get 
new global IDS
Per Joaquin, need to include ALL participants from cohort, even if they don't
have GUIDs
```{r}
htp_guids <- read_xlsx(synGet("syn53707355")$path) %>% 
  select(participant_external_id = `Stripped Lab ID`,
         guid = GUID)

dsc_guids <- read_csv(synGet("syn53708715")$path) %>%
  mutate(patient_id = as.character(patient_id)) %>% 
  select(participant_external_id = patient_id, guid)

#dsc participants from v3
dsc_all_participants <- read_csv(synGet("syn51216460")$path) %>% 
  filter(`Study Code` == "DSC") %>% 
  select(study_code = `Study Code`,
         participant_global_id = `Participant Global ID`,
         participant_external_id = `Participant External ID`)

#htp participants from v4 (includes VBR)
htp_all_participants <- read_csv(synGet("syn52399152")$path) %>% 
  select(study_code = `Study Code`,
         participant_global_id = `Participant Global ID`,
         participant_external_id = `Participant External ID`)

htp_dsc_all_participants <- bind_rows(dsc_all_participants,
                                      htp_all_participants)
```

Check for dupes:
```{r}
get_dupes(htp_guids, guid) #0
get_dupes(htp_guids, participant_external_id) #0

get_dupes(dsc_guids, guid) %>%
  filter(str_detect(guid, "NDAR"))
```

Add GUIDs
replace #Error# with NA
```{r}
htp_dsc_guids <- bind_rows(htp_guids, dsc_guids)

htp_dsc_guid_mapping <- htp_dsc_all_participants %>% 
  left_join(htp_dsc_guids) %>% 
  select(guid, study_code, participant_global_id) %>% 
  mutate(guid = na_if(guid, "#ERROR#"))
```

Counts:
- Total HTP participants: 1055
- Total DSC participants: 3484
- HTP participants missing GUIDs: 565
- DSC participants missing GUIDs: 68
- 52 GUIDs in both DSC and HTP

```{r}
htp_dsc_guid_mapping %>% 
  filter(study_code == "DSC" & is.na(guid))

htp_dsc_guid_mapping %>%
  filter(!is.na(guid)) %>% 
  group_by(guid) %>% 
  filter(min(study_code) == "DSC" & max(study_code) == "HTP") %>% 
  add_count(guid) %>% 
  arrange(desc(n), guid)
```

Write out & upload to Synapse
```{r}
htp_dsc_guid_mapping %>% 
  write_csv(here::here("output", "htp_dsc_guids_DRAFT_2024-02-29.csv"))
```


TODO: Delete 9 participants from DSC v3; add 159 new participants who don't have global IDs yet





