---
title: "Harmonize TEAM-DS clinical data, v5 April 2024"
---

```{r}
library(tidyverse)
library(synapser)
library(readxl)
library(googlesheets4)

synLogin()

gs4_deauth()
```

## MAKE DICTIONARY

Get dataset

```{r}
teamds_clinical_data <- read_xlsx(synGet("syn53058587")$path)
```

In Excel, identify columns that should be numeric but aren't - list below:

- height
- weight
- bmi

For the above cols, replace "missing" with `NA`
And convert to numeric

Then select all numeric columns and get ranges (min-max)

Also convert `studyid` to character to avoid future errors

```{r}
teamds_clinical_data_converted <- teamds_clinical_data %>% 
  mutate(height = as.numeric(na_if(height, "missing")), 
         weight = as.numeric(na_if(weight, "missing")), 
         bmi = as.numeric(na_if(bmi, "missing")),
         studyid = as.character(studyid))

teamds_num_entries <- teamds_clinical_data_converted %>% 
  select(where(is.numeric))  %>% 
  broom::tidy(summary()) %>% 
  select(original_header = column, min, max) %>% 
  unite(min, max, col = "values", sep = "-")

```

Get unique values for character columns

```{r}
teamds_clin_entries <- teamds_clinical_data_converted %>% 
  select(where(is.character)) %>% 
  pivot_longer(cols = everything(),
               names_to = "original_header",
               values_to = "all_values") %>% 
  group_by(original_header) %>% 
  summarize(values = paste(unique(all_values), collapse = "; "))
```


Assemble dictionary

```{r}
teamds_all_entries <- bind_rows(teamds_num_entries, teamds_clin_entries)

teamds_dict <- tibble("original_header" = colnames(teamds_clinical_data),
       "working_header" = NA) %>% 
  left_join(teamds_all_entries)
```

Write out and paste into Google sheet 
(https://docs.google.com/spreadsheets/d/118amKSbrLRbmHrczFPhkLsxLKFu3GEr5goPMDvrlvxw)

In Google sheet, add original NA value to numeric ranges where NA was replaced 
(e.g. "36-69, missing")

```{r}
write_csv(teamds_dict, here::here("output", "teamds_dict.csv"))
```


Get "long" version for ETL sheet

```{r}
teamds_conditions_long <- teamds_clinical_data_converted %>% 
  select(where(is.character)) %>% 
  pivot_longer(cols = everything(),
               names_to = "original_header",
               values_to = "all_values") %>% 
  group_by(original_header) %>% 
  summarize(unique_values = unique(all_values)) %>% 
  arrange(unique_values, .by_group = TRUE) %>%
  ungroup() %>% 
  filter(unique_values != "missing" & unique_values != "No")
```

```{r}
write_csv(teamds_conditions_long, 
          here::here("output", "teamds_conditions_long.csv"))
```


## PRE-HARMONIZATION CHECKS

Checks
```{r}
# N unique participants
length(unique(teamds_clinical_data_converted$studyid)) #122

# duplicate ids?
teamds_clinical_data_converted %>% janitor::get_dupes(studyid) #0
```

Generate Global IDs
```{r}
teamds_clinical_data_converted %>%
  select(descriptor = studyid) %>% 
  mutate(fhirResourceType = "Patient") %>% 
  write_csv(here::here("output", "teamds_dewrangle_input.csv"))
```


Add global IDs
```{r}
teamds_global_ids <- read_csv(here::here("output", 
                                         "global-descriptors-team-ds.csv")) %>% 
  select(globalId, descriptor)

teamds_clinical_data_converted <- teamds_clinical_data_converted %>% 
  left_join(teamds_global_ids,
            by = c("studyid" = "descriptor")) %>% 
  select(globalId, everything()) 

#%>% filter(is.na(globalId))
```

## HARMONIZE PARTICIPANT

Participant ID, Sex, Ethnicity - rename, no other changes needed
```{r}
teamds_participant <- teamds_clinical_data_converted %>% 
  rename(participantGlobalId = globalId,
         participantExternalId = studyid,
         sex = gender)
```

Race

```{r}
teamds_etl_plan <- etl_to_linkml_plan %>% 
  filter(study == "TEAM-DS",
         model_slot == "race") %>% 
  select(model_slot, model_enum_label, original_value) %>% 
  unnest(original_value)

mtor_values <- teamds_etl_plan %>% 
  filter(model_enum_label == "More than one race") %>% 
  pull(original_value)

teamds_participant <- teamds_participant %>% 
  left_join(teamds_etl_plan %>% filter(model_slot == "race") %>% 
              select(model_enum_label, original_value), 
            by = c("race" = "original_value")) %>% 
  rename(race_harmonized = model_enum_label) %>%
  relocate(race_harmonized, .after = race) %>% 
  mutate(race_harmonized = case_when(race_other %in% mtor_values ~ "More than one race",
                                     TRUE ~ race_harmonized)) %>% 
  select(-race) %>% rename(race = race_harmonized)

```


Family ID: 101061 / 101062 & 101088 / 101089 sibling pairs - use min(Participant ID) as family ID
**use blanks ("") instead of `NA`**
```{r}
teamds_participant <- teamds_participant %>% 
  mutate(familyId = case_when(
    participantExternalId %in% c("101061", "101062") ~ "101061",
    participantExternalId %in% c("101088", "101089") ~ "101088"),
    .after = participantExternalId) %>% 
  mutate(familyType = case_when(
    familyId %in% c("101061", "101088") ~ "Other",
    TRUE ~ "Proband-only"), .after = familyId) %>% 
  mutate(fatherId = "", motherId = "", 
         siblingId = case_when(
           participantExternalId == "101061" ~ "101062",
           participantExternalId == "101062" ~ "101061",
           participantExternalId == "101088" ~ "101089",
           participantExternalId == "101089" ~ "101088"),
         otherFamilyMemberId = "",
         familyRelationship = "Proband",
         .after = familyType)
```


Final Participant:
Replace `NA` with blank throughout to avoid LinkML validation issues
```{r}
teamds_participant_harmonized <- teamds_participant %>% 
  select(participantGlobalId, participantExternalId, familyId, familyType,
         fatherId, motherId, siblingId, otherFamilyMemberId, familyRelationship,
         sex, race, ethnicity, ageAtFirstPatientEngagement = age) %>% 
  mutate(studyCode = "TEAM-DS",
         downSyndromeStatus = "T21", 
         ageAtFirstPatientEngagement = round(365.25 * ageAtFirstPatientEngagement, 0),
         firstPatientEngagementEvent = "Enrollment",
         outcomesVitalStatus = "Unknown or not available",
         ageAtLastVitalStatus = "") %>% 
  select(all_of(participant_schemalabels)) %>% 
  setNames(participant_displaynames) %>% 
  mutate(across(everything(), ~replace_na(.x, "")))

teamds_participant_harmonized %>% 
  write_csv(here::here("output", "teamds_participant_harmonized_DRAFT_2024-04-12.csv"))
```

**Trying to get summary stats, this needs work**

```{r}
teamds_summary <- teamds_participant_harmonized %>% 
  mutate(across(c(`Family Type`, `Family Relationship`, Sex, Race,
                  Ethnicity, `Down Syndrome Status`, 
                  `First Patient Engagement Event`, `Outcomes Vital Status`),
                as.factor)) %>% 
  skimr::skim_without_charts() 

write_csv(teamds_summary, here::here("output", "teamds_summary.csv"))
```


## CONDITIONS

Start with `dstype` and structured columns; add the rest after manual review of
Trish's annotations

```{r}
condition_lookup_teamds <- condition_codes_v5 %>% 
  filter(study == "TEAM-DS") %>% 
  unnest(source_column_value)

teamds_structured_cols <- condition_lookup_teamds %>% 
  pull(source_column) %>% unique()
```


```{r}
teamds_structured_conditions_harmonized <- teamds_participant %>%
  select(participantExternalId, all_of(teamds_structured_cols)) %>% 
  pivot_longer(cols = c(-participantExternalId),
               names_to = "source_column", 
               values_to = "source_column_value") %>%
  #pull(source_column_value) %>% unique()
  filter(!source_column_value %in% c("No", "NA", NA, "missing")) %>% 
  left_join(condition_lookup_teamds, 
            join_by(source_column, source_column_value)) %>% 
  #filter(is.na(study))
  mutate(studyCode = study,
         eventId = "",
         eventType = "",
         ageAtConditionMeasureObservation = "",
         conditionInterpretation = "Observed",
         conditionStatus = "History Of",
         conditionDataSource = "Self-reported",
         measureValue = "",
         measureUnit = "") %>%
  select(all_of(condition_schemalabels)) %>%
  setNames(condition_displaynames)
```
```{r}
teamds_structured_conditions_harmonized %>% 
  write_csv(here::here("output", "teamds_structured_conditions_2024-04-22.csv"))
```
















