---
title: "CDE basic stats"
author: "Pierrette Lo"
date: '2022-10-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(DT)

mcd_original <- read_csv(here::here("data", "mcd_original.csv"))

cde_tags <- read_csv(here::here("data", "cde_tags.csv"))%>% 
  # get CDE categories
  mutate(cde_category = str_extract(`Variable / Field Name`, "^[:alpha:]*(?=_)"), .before = `Used?`)

pheno_codes_cde <- read_csv(here::here("data", "pheno_codes_v2_cde.csv")) %>% 
  mutate(cde_category = str_extract(CDE_tag, "^[:alpha:]*(?=_)"), .after = CDE_tag)
```

<br>

How many CDEs in the [original worksheet](https://docs.google.com/spreadsheets/d/1gUiFUAke5vz-JId4mExoYxwx3X0pr2TdbTlMvtZtTdo/edit#gid=902968406)?
```{r, echo = FALSE}
n_distinct(mcd_original$`Variable / Field Name`)
```

Number of distinct [CDEs used for tagging](https://docs.google.com/spreadsheets/d/1gUiFUAke5vz-JId4mExoYxwx3X0pr2TdbTlMvtZtTdo/edit#gid=308350820). 

The original CDE list linked above was narrowed down because there were many redundant fields, as the CDEs were taken directly from the REDCap survey. E.g. `gi_other_gi_status`, `gi_any`, `gi_conditions`, and `gi_other_specify` are similar, so only `gi_other_gi_status` was used for tagging.
```{r, echo = FALSE}
n_distinct(cde_tags$`Variable / Field Name`)
```

<br>

### CDE tags used at least once in data?
```{r, echo = FALSE}
cde_tags %>% 
  count(`Used?`) %>% 
  ggplot(aes(y = fct_inorder(`Used?`), x = n)) +
  geom_col() +
  xlab(NULL) +
  ylab(NULL)
```

<br>

### Browse the 63 CDEs that weren't used (yet)
```{r, echo = FALSE}
cde_tags %>% 
  filter(`Used?` == "N") %>% 
  select(`Variable / Field Name`, cde_category) %>% 
  datatable()
```

<br>

### Breakdown of CDE categories in the [actual data](https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit#gid=1006243162)
```{r, echo = FALSE}
pheno_codes_cde %>% 
  count(cde_category) %>% 
  ggplot(aes(y = fct_reorder(cde_category, n), x = n)) +
  geom_col() +
  xlab(NULL) +
  ylab(NULL) +
  ggtitle("NA = no relevant CDE")
```

<br>

### Conditions without matching CDEs, and suggested categories
```{r, echo = FALSE}
pheno_codes_cde %>% 
  filter(is.na(cde_category)) %>% 
  select(cohort, display_text, CDE_tag) %>% 
  arrange(desc(CDE_tag)) %>% 
  datatable()
```


<br>

### Count of conditions in the "Other" buckets

Some of these catch-all buckets are quite large (e.g. `cardiac_other_chd`, `gi_other`) - consider adding more specific CDEs for recurrently observed conditions
```{r, echo = FALSE}
pheno_codes_cde %>% 
  filter(str_detect(CDE_tag, "other")) %>% 
  count(CDE_tag) %>% 
  ggplot(aes(y = fct_reorder(CDE_tag, n), x = n)) +
  geom_col() +
  xlab(NULL) +
  ylab(NULL) 
```

<br>

### Browse conditions in the "Other" buckets
```{r, echo = FALSE}
pheno_codes_cde %>% 
  filter(str_detect(CDE_tag, "other")) %>% 
  select(cohort, display_text, CDE_tag, cde_category) %>% 
  arrange(cde_category) %>% 
  datatable()
```

