---
title: "Harmonize DSC clinical data"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```

Read in data
Looks like read_csv converts the Acknowledgments rows to NA -> delete below
```{r}
dsc_demographic <- read_csv(synGet("syn26445516")$path)
dsc_ihq <- read_csv(synGet("syn26445517")$path)
```

##### DATA CLEANUP & CHECKS

From Dawn, 2022-08-29:
- We have updated the machine-readable headers
- Please use the IHQ as our main 'source of truth' for who you have data for (all de-identified of course, but the demographics file contains a few - that are not listed in the IHQ file) - we had to remove 40 (ish) people for having edited answers that didn't get recorded correctly
- We removed PID 12459 as we suspect they have an erroneous birthdate
- This export will likely differ from the last one as we ran the export from most recent, completed IHQs only, last time we provided 'historical' sessions as well. This should resolve your data curation questions from our last meeting

Also:
Delete duplicate demographics row for pt 5458
Delete pt 13061 - has IHQ but no demographics
Delete 59 pts with demographics but no IHQ

```{r}
dsc_demographic <- dsc_demographic %>% 
  filter(!is.na(patient_id)) %>% #del Acknowledgments
  distinct() #del 5458

dsc_ihq <- dsc_ihq %>% 
  filter(!is.na(patient_id)) %>% #del Acknowledgments
  filter(patient_id != 13061) #del 13061

n_distinct(dsc_demographic$patient_id) #3432
n_distinct(dsc_ihq$patient_id) #3373

# all pts with ihq have demographics
setdiff(dsc_ihq$patient_id, dsc_demographic$patient_id)

# 59 pts with demographics do not have ihq -> delete for now
no_demog <- setdiff(dsc_demographic$patient_id, dsc_ihq$patient_id)

dsc_demo_transformed <- dsc_demographic %>% 
  filter(!(patient_id %in% no_demog))

setdiff(dsc_demo_transformed$patient_id, dsc_ihq$patient_id)
```



##### PARTICIPANT

Family ID = parent_id if exists, otherwise = patient_id
Bring in parent_id from ihq
```{r}
dsc_demo_transformed <- dsc_demo_transformed %>% 
  left_join(dsc_ihq %>% select(patient_id, parent_id) %>% distinct()) %>% 
  relocate(parent_id, .after = patient_id) %>% 
  mutate(family_id = ifelse(!is.na(parent_id), parent_id, patient_id),
         .after = parent_id) 
```


Other cols that need to be transformed: 

Family Type:
"Proband-only" if 1 unique patient_id/family_id combo
"Other" if >1 patient_id/family_id
** may want to add "Siblings" to Family Type in schema

```{r}
dsc_demo_transformed <- dsc_demo_transformed %>% 
  group_by(family_id) %>% 
  mutate(n = n(), .after = family_id) %>% 
  mutate(family_type = case_when(n == 1 ~ "Proband-only",
                                 n == 2 ~ "Other"), .after = n) %>% 
  ungroup()
```

Race:
- A couple of participants (4068 & 6181) chose all the options including Unknown -> assume they are "more than one race"
- Not specifying the different races (if >1) for now - maybe in a future release
- "Prefer not to answer" if all race columns are blank
```{r}
dsc_race <- dsc_demo_transformed %>% 
  select(patient_id, starts_with("race")) %>% 
  pivot_longer(cols = -patient_id,
               names_to = "race_col",
               values_to = "status") %>%
  filter(status == 1) %>% 
  group_by(patient_id) %>%
  mutate(n = n()) %>% 
  arrange(desc(n)) %>%
  ungroup() %>% 
  mutate(race = case_when(n > 1 ~ "More than one race",
                          n == 1 ~ str_extract(race_col, "(?<=_).*$"))) %>% 
  mutate(race = case_when(race == "white" ~ "White",
                          race == "black" ~ "Black or African American",
                          race == "asian" ~ "Asian",
                          race == "pacific_islander" ~ "Native Hawaiian or Other Pacific Islander",
                          race == "american_indian" ~ "American Indian or Alaska Native",
                          race == "unknown" ~ "Unknown",
                          TRUE ~ race)) %>% 
  select(patient_id, race) %>% 
  distinct()

# join with main Demographic table
# and replace "NA" with "Prefer not to answer"
dsc_demo_transformed <- dsc_demo_transformed %>% 
  left_join(dsc_race) %>% 
  mutate(race = replace_na(race, "Prefer not to answer"))
```

Ethnicity - replace NA with Prefer not to answer

```{r}
dsc_demo_transformed <- dsc_demo_transformed %>% 
  mutate(ethnicity = replace_na(ethnicity, "Prefer not to answer"))
```

Down Syndrome Status - all participants have DS

Outcomes & Age at Vital Status - leave out for now due to below inconsistencies
```{r}
# 4 patients have vital_status = Yes (Alive) but deceased_year is not NA
# 6725 = survey_time 2015
# 6841 = 2016
# 11358 = 2019
# 11823 = 2020
dsc_demo_transformed %>% 
  left_join(dsc_ihq %>% select(patient_id, vital_status, survey_time)) %>% 
  select(patient_id, deceased_year, vital_status, survey_time) %>% 
  filter(vital_status == "Yes" & !is.na(deceased_year))

# 5 patients are dead but no deceased_year
dsc_demo_transformed %>% 
  left_join(dsc_ihq %>% select(patient_id, vital_status)) %>% 
  select(patient_id, deceased_year, vital_status) %>% 
  filter(vital_status == "No")

# 1 pt (3981) has vital_status NA but has deceased_year
dsc_demo_transformed %>% 
  left_join(dsc_ihq %>% select(patient_id, vital_status, survey_time)) %>% 
  select(patient_id, deceased_year, vital_status, survey_time) %>% 
  filter(!is.na(deceased_year) | vital_status == "No") %>% 
  arrange(vital_status) 

```


Create final Participant table - Rename/create empty cols to match DCA template, has to be in order - For empty columns, replace NA with "Not provided"

Finally, take only distinct Participant ID (filter out multiple visits)

```{r}
dsc_participant <- dsc_demo_transformed %>% 
  mutate(`Study Code` = "DSC",
         `Father ID` = "Not provided",
         `Mother ID` = "Not provided",
         `Family Relationship` = "Not provided",
         `Down Syndrome Status` = "T21",
         `Outcomes Vital Status` = "Not provided",
         `Age at Last Vital Status` = "Not provided") %>% 
  select(`Study Code`,
         `Participant ID` = patient_id,
         `External ID` = patient_id,
         `Family ID` = family_id,
         `Family Type` = family_type,
         `Father ID`,
         `Mother ID`,
         `Family Relationship`,
         Sex = gender,
         Race = race,
         Ethnicity = ethnicity,
         `Down Syndrome Status`,
         `Outcomes Vital Status`,
         `Age at Last Vital Status`)
```

```{r}
write_csv(dsc_participant, here::here("output", "dsc_participant.csv"))
```


##### CONDITION

Columns: 
Participant ID 
Condition Description 
Age at Condition Observation (Days) - blank for now 
Condition Interpretation - all "Observed" for now (ie. only positive assertions) 
Condition Source - all "Clinical" 
HPO Label 
HPO Code 
MONDO Label 
MONDO Code 
MAXO Label 
MAXO Code 
Other Label 
Other Code


Bring in previously annotated phenotypes csv downloaded from ETL Google sheet (https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit#gid=1006243162)

```{r}
pheno_lookup_dsc <- read_csv(here::here("data", "pheno_codes_v2.csv")) %>% 
  filter(cohort == "DSC")
```

Special cols (ie. NOT 1)
leukemia_ever_diagnosis
dementia_ad_cognitive_ever_diagnosis
dementia_ad_cognitive_ever_diagnosis
dementia_ad_cognitive_ever_diagnosis
celiac_ever_diagnosis -> gi_condition_celiac
chd_ever_diagnosis -> chd_condition_
skeletal_ever_diagnosis -> skeletal_condition_
thyroid_ever_diagnosis -> thyroid_condition_
seizures_epilepsy_ever_diagnosis
gi_ever_diagnosis -> gi_condition_
sleep_ever_diagnosis -> sleep_condition_
covid19_ever_diagnosis
diabetes_ever_diagnosis
eye_exam_ever
hearing_testing_ever
ds_specific_diagnosis


Data checks:
```{r}
# patients where chd_ever = Yes but doesn't have other chd_conditions: 17
# https://stackoverflow.com/a/62162600
dsc_ihq %>% 
  select(patient_id, chd_ever_diagnosis, starts_with("chd_condition_"), -chd_condition_none) %>% 
  rowwise() %>% 
  filter(chd_ever_diagnosis == "Yes" & 
           all(is.na(c_across(starts_with("chd_condition")))))

# patients where chd_ever = No but has other chd_conditions? 
dsc_ihq %>% 
  select(patient_id, chd_ever_diagnosis, starts_with("chd_condition_"), -chd_condition_none) %>% 
  rowwise() %>% 
  filter(chd_ever_diagnosis == "No" & 
           any(!is.na(c_across(starts_with("chd_condition")))))

# or patients where chd_ever = Yes but chd_condition_none = No
dsc_ihq %>% 
  filter(chd_ever_diagnosis == "Yes" & chd_condition_none == 1) %>% 
  pull(patient_id)

```

Check GI cols
```{r}
# are there any patients where gi_ever = Yes but doesn't have other gi_conditions? 37
dsc_ihq %>% 
  select(patient_id, gi_ever_diagnosis, starts_with("gi_condition_"), -gi_condition_none) %>% 
  rowwise() %>% 
  filter(gi_ever_diagnosis == "Yes" & 
           all(is.na(c_across(starts_with("gi_condition")))))

# or patients where chd_ever = No but has other chd_conditions? 
dsc_ihq %>% 
  select(patient_id, chd_ever_diagnosis, starts_with("chd_condition_"), -chd_condition_none) %>% 
  rowwise() %>% 
  filter(chd_ever_diagnosis == "No" & 
           any(!is.na(c_across(starts_with("chd_condition")))))

# or patients where chd_ever = Yes but chd_condition_none = No
dsc_ihq %>% 
  filter(chd_ever_diagnosis == "Yes" & chd_condition_none == 1) %>% 
  pull(patient_id)

```


CHD cols - just choose any positive assertion:
- chd_ever_diagnosis = Yes and no other selections, or chd_condition_unsure -> CHD
- chd_ever_diagnosis = Yes and chd_condition_none = 1 -> still CHD
- Otherwise go with the other selections (chd_)
- leave out chd_condition_unsure

GI cols
- gi_ever_diagnosis = Yes and no other selections, or gi_condition_unsure -> GI
- gi_ever_diagnosis = Yes and gi_condition_none = 1 -> still GI
- Otherwise go with the other selections (gi_condition_)
- leave out gi_condition_unsure
- celiac_ever_diagnosis = Yes OR gi_condition_celiac = 1 -> choose the positive assertion, ignore negative (will need to distinct() if both)

Sleep cols
- sleep_ever_diagnosis = Yes and no other selections, or sleep_condition_unsure -> sleep
- sleep_ever_diagnosis = Yes and sleep_condition_none = 1 -> still sleep
- Otherwise go with the other selections (sleep_)
- leave out sleep_condition_unsure

skeletal cols
- skeletal_ever_diagnosis = Yes and no other selections, or skeletal_condition_unsure -> skeletal
- skeletal_ever_diagnosis = Yes and skeletal_condition_none = 1 -> still skeletal
- Otherwise go with the other selections (skeletal_)
- leave out skeletal_condition_unsure

thyroid cols
- thyroid_ever_diagnosis = Yes and no other selections, or thyroid_condition_unsure -> thyroid
- thyroid_ever_diagnosis = Yes and thyroid_condition_none = 1 -> still thyroid
- Otherwise go with the other selections (thyroid_)
- leave out thyroid_condition_unsure


Pull conditions from ihq
```{r}
dsc_conditions_subset <- dsc_ihq %>% 
  select(patient_id, ds_specific_diagnosis,
         starts_with(c("chd", "cardiac", "gi", "skin", "seizures",
                       "neurologic", "sleep", "skeletal", "leukemia",
                       "cancer", "covid19", "thyroid", "diabetes", "celiac",
                       "metabolic", "hearing", "eye", "mental", "learning", "dementia"))) %>% 
  mutate(across(everything(), as.character)) %>% 
  pivot_longer(cols = -patient_id,
               names_to = "source_column",
               values_to = "status") %>% 
  filter(!is.na(status) & status != "No") %>% # delete blanks and _ever = No
  left_join(pheno_lookup_dsc, by = c("source_column", "status" = "source_column_value")) %>% 
  filter(!is.na(display_text))  # remove conditions like _none and _unsure that don't have annotations
```

Deal with _ever columns:

If a patient has ever_diagnosis AND _condition starting with the same -> delete ever, keep the more specific condition
```{r}
keep_specific_conditions <- dsc_conditions_subset %>% 
  filter(str_detect(source_column, "^chd|gi_|sleep|skeletal|thyroid") & !str_detect(source_column, "surgery")) %>% 
  mutate(prefix = str_extract(source_column, "^[^_]+(?=_)"), .before = source_column) %>% 
  group_by(patient_id, prefix) %>% # group by patient and condition category
  mutate(n = n(), .before = prefix) %>% 
  arrange(n) %>% 
  filter(n > 1) %>% # filter groups that only have 1 condition per category (could be _ever or something else) 
  filter(any(status == "Yes")) %>% # get only groups that include _ever
  ungroup() %>% 
  filter(status == 1) # remove evers, keep specific condition

#get back the groups with multiple conditions but no ever
keep_multiple_without_ever <- dsc_conditions_subset %>% 
  filter(str_detect(source_column, "^chd|gi_|sleep|skeletal|thyroid") & !str_detect(source_column, "surgery")) %>% 
  mutate(prefix = str_extract(source_column, "^[^_]+(?=_)"), .before = source_column) %>% 
  group_by(patient_id, prefix) %>% # group by patient and condition category
  mutate(n = n(), .before = prefix) %>% 
  arrange(n) %>% 
  filter(n > 1) %>% # filter groups that only have 1 condition per category (could be _ever or something else) 
  filter(all(status != "Yes")) %>% # get only groups that don't have _ever
  ungroup()

#get back the groups with 1 condition per category
keep_single_conditions <- dsc_conditions_subset %>% 
  filter(str_detect(source_column, "^chd|gi_|sleep|skeletal|thyroid") & !str_detect(source_column, "surgery")) %>% 
  mutate(prefix = str_extract(source_column, "^[^_]+(?=_)"), .before = source_column) %>% 
  group_by(patient_id, prefix) %>% # group by patient and condition category
  mutate(n = n(), .before = prefix) %>% 
  arrange(n) %>% 
  filter(n == 1) %>% 
  ungroup()

filtered_ever_conditions <- rbind(keep_specific_conditions, keep_multiple_without_ever, keep_single_conditions) %>% 
  select(-n, -prefix)

# check for duplicate rows
dsc_conditions_subset %>% 
  filter(!str_detect(source_column, "^chd|gi_ever|gi_condition|sleep|skeletal|thyroid")) %>% 
  rbind(filtered_ever_conditions) %>% 
  janitor::get_dupes() 

# add these back to the rest of the conditions
dsc_conditions_final <- dsc_conditions_subset %>% 
  filter(!str_detect(source_column, "^chd|gi_ever|gi_condition|sleep|skeletal|thyroid")) %>% 
  rbind(filtered_ever_conditions)
```


Create empty/all same value columns
Select columns in order of DCA template
```{r}
dsc_condition <- dsc_conditions_final %>% 
  mutate(`Study Code` = "DSC",
         `Participant ID` = patient_id,
         `Age at Condition Observation` = "Not provided",
         `Condition Interpretation` = "Observed",
         `Condition Source` = "Self-reported") %>% 
  select(`Study Code`,
         `Participant ID`,
         `Condition Description` = display_text,
         `Age at Condition Observation`,
         `Condition Interpretation`,
         `Condition Source`,
         `HPO Label` = HPO_label,
         `HPO Code` = HPO_code,
         `MONDO Label` = MONDO_label,
         `MONDO Code` = MONDO_code,
         `MAXO Label` = MAXO_label,
         `MAXO Code` = MAXO_code,
         `Other Label` = Other_label,
         `Other Code` = Other_code)
```

```{r}
write_csv(dsc_condition, here::here("output", "dsc_condition.csv"))
```
