---
title: "Harmonize ABCDS clinical data"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```

Read in data
```{r}
# control & DS demographics
demo_ctl <- read_csv(synGet("syn33401283")$path)
demo_ds <- read_csv(synGet("syn33401291")$path)

# control & DS health histories
health_ctl <- read_csv(synGet("syn33401148")$path)
health_ds <- read_csv(synGet("syn33401163")$path)

# DS karyotypes
karyo_ds <- read_csv(synGet("syn33401175")$path)

# link between controls & DS siblings
control_linkage <- read_xlsx(synGet("syn34654889")$path)
```


Combine DS and control subjects
```{r}
health_all <- rbind(health_ds, health_ctl)
demo_all <- rbind(demo_ds, demo_ctl)
```


Data checks
```{r}
#417 unique subjects, no setdiff between health & demo
n_distinct(health_all$subject_label)
n_distinct(demo_all$subject_label)
setdiff(health_all$subject_label, demo_all$subject_label)

# most patients have both Baseline & Cycle 2 health & demo
health_all %>% count(subject_label) %>% arrange(n)
demo_all %>% count(subject_label) %>% arrange(desc(n))

# the demo cols we're interested in are the same at both timepoints
demo_all %>% select(subject_label, de_gender, de_race, de_ethnicity, ds_vs_control_flag) %>% 
  distinct() %>% count(subject_label) %>% arrange(desc(n))

# there are changes in hh between the 2 timepoints - asked ABCDS team what to do (FOR NOW - use baseline data only)
health_all %>% select(subject_label, starts_with("hh")) %>% 
  distinct() %>% count(subject_label) %>% arrange(desc(n))

# controls - how many have DS sibs
n_distinct(demo_ctl$subject_label)
n_distinct(control_linkage$ABCDS_ID_SC)

# 10233 10224 10237 10235 are sibs in the linkage sheet who don't have any data
setdiff(control_linkage$ABCDS_ID_SC, demo_ctl$subject_label)
```

FOR NOW: limit to Baseline data only
```{r}
health_BL <- health_all %>% filter(event_code == "bl")
demo_BL <- demo_all %>% filter(event_code == "bl")

setdiff(demo_BL$subject_label, health_BL$subject_label)
# 2 subjects, 23001 and 23002, don't have baseline health histories
```

##### PARTICIPANT

Study Code = "ABC-DS"
Participant ID = subject_label
External ID = subject_label
Family ID
Family Type = Other
Father ID = Not provided
Mother ID = Not provided
Family Relationship
Sex
Race
Ethnicity
Down Syndrome Status
Outcomes Vital Status
Age at Last Vital Status


Bring in ABCDS_ID_DS from linkage table
Family ID = ABCDS_ID_DS if exists, otherwise = subject_label
```{r}
demo_BL_transformed <- demo_BL %>%  
  left_join(control_linkage %>% select(ABCDS_ID_SC, ABCDS_ID_DS),
            by = c("subject_label" = "ABCDS_ID_SC")) %>% 
  relocate(ABCDS_ID_DS, .after = subject_label) %>% 
  mutate(family_id = ifelse(!is.na(ABCDS_ID_DS), ABCDS_ID_DS, subject_label),
         .after = subject_label) 
```

Other cols that need to be transformed:

Family Type:
"Proband-only" if 1 unique patient_id/family_id combo
"Other" if >1 patient_id/family_id
** may want to add "Siblings" to Family Type in schema

Sex, Race, Ethnicity - need to be decoded
DS Status 

```{r}
demo_BL_transformed <- demo_BL_transformed %>% 
  group_by(family_id) %>% 
  mutate(n = n(), .after = family_id) %>%
  mutate(family_type = case_when(n == 1 ~ "Proband-only",
                                 n == 2 ~ "Other"), .after = n) %>% 
  ungroup() %>% 
  mutate(sex = case_when(de_gender == 1 ~ "Male",
                         de_gender == 2 ~ "Female"),
         .after = de_gender) %>% 
  mutate(race = case_when(de_race == 1 ~ "White",
                          de_race == 2 ~ "Black or African American",
                          de_race == 3 ~ "American Indian or Alaska Native",
                          de_race == 4 ~ "Asian",
                          de_race == 5 ~ "Native Hawaiian or Other Pacific Islander",
                          de_race == 6 ~ "Unknown",
                          str_detect(de_race, "\\|") ~ "More than one race"),
         .after = de_race) %>% 
  mutate(ethnicity = case_when(de_ethnicity == 0 ~ "Not Hispanic or Latino",
                               de_ethnicity == 1 ~ "Hispanic or Latino",
                               de_ethnicity == 2 ~ "Unknown"),
         .after = de_ethnicity) %>% 
  mutate(ds_status = ifelse(ds_vs_control_flag == "DS", "T21", "D21"),
         .after = ds_vs_control_flag)

```


Create final Participant table - Rename/create empty cols to match DCA template, has to be in order - For empty columns, replace NA with "Not provided"

```{r}
abcds_participant <- demo_BL_transformed %>% 
  mutate(`Study Code` = "ABC-DS",
         `Father ID` = "Not provided",
         `Mother ID` = "Not provided",
         `Family Relationship` = "Not provided",
         `Outcomes Vital Status` = "Not provided",
         `Age at Last Vital Status` = "Not provided") %>% 
  select(`Study Code`,
         `Participant ID` = subject_label,
         `External ID` = subject_label,
         `Family ID` = family_id,
         `Family Type` = family_type,
         `Father ID`,
         `Mother ID`,
         `Family Relationship`,
         Sex = sex,
         Race = race,
         Ethnicity = ethnicity,
         `Down Syndrome Status` = ds_status,
         `Outcomes Vital Status`,
         `Age at Last Vital Status`)
```

```{r}
write_csv(abcds_participant, here::here("output", "abcds_participant.csv"))
```


##### CONDITION

Columns: 
Participant ID 
Condition Description 
Age at Condition Observation (Days) - blank for now 
Condition Interpretation - all "Observed" for now (ie. only positive assertions) 
Condition Source - all "Clinical" 
HPO Label 
HPO Code 
MONDO Label 
MONDO Code 
MAXO Label 
MAXO Code 
Other Label 
Other Code


Bring in previously annotated phenotypes csv downloaded from ETL Google sheet (https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit#gid=1006243162)

```{r}
pheno_lookup_abcds <- read_csv(here::here("data", "pheno_codes_v2.csv")) %>% 
  filter(cohort == "ABCDS")
```

Remove columns that won't be harmonized (either for Sept, or ever)
ie. highlighted in grey or blue in `abcds_dic_harmonizing.xlsx`

Pull in karyotype
```{r}
health_BL_transformed <- health_BL %>% 
  select(-c(site_id, event_code, event_label, hh_cancer, hh_diabetes, hh_done, hh_fracture_5yr, hh_fracture_loc, ends_with("status"), ends_with("treated"), hh_hep_b_carrier, hh_hep_b_immune, hh_hep_b_vaccine, starts_with("hh_hrt_"), hh_menopausal, hh_menopausal_onset, hh_menopausal_status, hh_menstruated, hh_ndreason, update_stamp)) %>% 
  left_join(karyo_ds %>% select(subject_label, karyotype))
```

Get conditions:

Cols where NA matters:
- if hh_anemia = 1 AND hh_anemia_type = NA -> "blank AND hh_anemia = 1"

- if hh_chd = 1 AND hh_chd_type = NA ~ hh_chd_type -> "blank AND hh_chd = 1"
10065
20441
21306
21381
22768

Make all cols char so I can pivot longer

pivot longer

split cells with | using separate_rows
```{r}
# first fix the special cases for hh_anemia and hh_chd
health_BL_conditions <- health_BL_transformed %>%
  mutate(hh_chd_type = ifelse(hh_chd == 1 & is.na(hh_chd_type), "blank AND hh_chd = 1", hh_chd_type)) %>% 
  mutate(hh_anemia_type = ifelse(hh_anemia == 1 & is.na(hh_anemia_type), "blank AND hh_anemia = 1", hh_anemia_type)) %>% 
  select(-hh_chd, -hh_anemia) %>% 
  # then pivot longer
  mutate(across(everything(), as.character)) %>% 
  pivot_longer(cols = -subject_label,
               names_to = "condition",
               values_to = "status") %>% 
  # keep positive assertions
  filter(status != "0") %>% 
  # separate out multi-responses
  separate_rows(status, sep = "\\|") %>% 
  # look up annotations
  left_join(pheno_lookup_abcds, by = c("condition" = "source_column", "status" = "source_column_value"))
```


Create empty/all same value columns
Select columns in order of DCA template
```{r}
abcds_condition <- health_BL_conditions %>% 
  mutate(`Study Code` = "ABC-DS",
         `Age at Condition Observation` = "Not provided",
         `Condition Interpretation` = "Observed",
         `Condition Source` = "Clinical") %>% 
  select(`Study Code`,
         `Participant ID` = subject_label,
         `Condition Description` = display_text,
         `Age at Condition Observation`,
         `Condition Interpretation`,
         `Condition Source`,
         `HPO Label` = HPO_label,
         `HPO Code` = HPO_code,
         `MONDO Label` = MONDO_label,
         `MONDO Code` = MONDO_code,
         `MAXO Label` = MAXO_label,
         `MAXO Code` = MAXO_code,
         `Other Label` = Other_label,
         `Other Code` = Other_code)
```

```{r}
write_csv(abcds_condition, here::here("output", "abcds_condition.csv"))
```
