---
title: "Harmonize Nexus clinical data"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)

synLogin()
```

Read in data:

### Need to set up Synapse project for Nexus

```{r}
nexus_clinical_data <- read_xlsx(synGet("syn25765850")$path)
```

##### PARTICIPANT

Clean up headers, delete everything in parens
Add Study Code = "Nexus"

### need to get KF Patient IDs

```{r}
nexus_transformed <- nexus_clinical_data %>% 
  rename_with(~ str_trim(gsub("\\s\\(.*", "", .x))) %>% 
  mutate(`Study Code` = "Nexus",
         `External ID` = `Participant ID`, .before = everything())
```

Family ID - as is

Family Type = "Proband-only"

Sex = as is

Race
```{r}
unique(nexus_transformed$Race)

nexus_transformed <- nexus_transformed %>%
  mutate(Race = case_when(
    Race == "Other (White and Hispanic)" ~ "White",
    Race == "Other (Hispanic)" ~ "Other",
    Race == "Black" ~ "Black or African American",
    TRUE ~ Race
  ))

```

Ethnicity - as is

Down Syndrome Status = all T21

Outcomes Vital Status - N/A 
Age at the Last Vital Status (Days) - N/A

Create final Participant table - Rename/create empty cols to match DCA template, has to be in order - For empty columns, replace NA with "Not provided"

```{r}
nexus_participant <- nexus_transformed %>% 
  mutate(`Down Syndrome Status` = "T21",
         `Family Type` = "Proband-only",
         `Father ID` = "Not provided",
         `Mother ID` = "Not provided",
         `Family Relationship` = "Not provided",
         `Outcomes Vital Status` = "Not provided",
         `Age at Last Vital Status` = "Not provided") %>% 
  select(`Study Code`,
         `Participant ID`,
         `External ID`,
         `Family ID`,
         `Family Type`,
         `Father ID`,
         `Mother ID`,
         `Family Relationship`,
         Sex,
         Race,
         Ethnicity,
         `Down Syndrome Status`,
         `Outcomes Vital Status`,
         `Age at Last Vital Status`)

```

```{r}
write_csv(nexus_participant, here::here("output", "nexus_participant.csv"))
```


##### CONDITION

New table! Columns: 
Participant ID 
Condition Description 
Age at Condition Observation (Days) - blank for now 
Condition Interpretation - all "Observed" for now (ie. only positive assertions) 
Condition Source - all "Clinical" 
HPO Label 
HPO Code 
MONDO Label 
MONDO Code 
MAXO Label 
MAXO Code 
Other Label 
Other Code


Bring in previously annotated phenotypes csv downloaded from ETL Google sheet (https://docs.google.com/spreadsheets/d/1cHSburUDg6CR4az5FZR82FI992h7cyv-22pLVTFLZvE/edit#gid=1006243162)

```{r}
pheno_lookup <- read_csv(here::here("data", "pheno_codes_v2.csv")) %>% 
  filter(cohort == "Nexus")
```


Pull conditions from clinical data
Add annotations from lookup table
Create empty/all same value columns
Select columns in order of DCA template
```{r}

nexus_condition <- nexus_transformed %>% 
  select(`Study Code`,
         `Participant ID`,
         `Condition Description` = `Diagnoses Text`) %>% 
  left_join(pheno_lookup, 
            by = c("Condition Description" = "source_column_value")) %>% 
  mutate(`Age at Condition Observation` = "Not provided",
         `Condition Interpretation` = "Observed",
         `Condition Source` = "Clinical") %>% 
  select(`Study Code`,
         `Participant ID`,
         `Condition Description`,
         `Age at Condition Observation`,
         `Condition Interpretation`,
         `Condition Source`,
         `HPO Label` = HPO_label,
         `HPO Code` = HPO_code,
         `MONDO Label` = MONDO_label,
         `MONDO Code` = MONDO_code,
         `MAXO Label` = MAXO_label,
         `MAXO Code` = MAXO_code,
         `Other Label` = Other_label,
         `Other Code` = Other_code)
```

```{r}
write_csv(nexus_condition, here::here("output", "nexus_condition.csv"))
```
