---
title: "March 2023 data cleaning"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(synapser)
library(readxl)
library(skimr)

synLogin()
```

Hakon:
- DNA: syn44291080
- RNA: syn44291064
de Smith: syn44271573
DS-Sleep:
- demo (DCG): syn42114527

**should we share the clinical data if sample failed seq? I think so**


### de Smith:

```{r}
desmith_clin <- read_xlsx(synGet("syn44271573")$path)
```

```{r}
skim(desmith_clin) %>% view()
```

Check that IDs are unique:
```{r}
n_distinct(desmith_clin$WPSID) #476
```


Cols that are <chr> but should contain numbers:
Age at enrollment
Gestational age at birth
Birthweight
Age at last known vital status
Platelets at birth
Peripheral blood blasts at birth

```{r}
# use datapasta to make vector
misclassified <- c("Age at enrollment", "Gestational age at birth", "Birthweight", "Age at last known vital status", "Platelets at birth", "Peripheral blood blasts at birth")

unique(desmith_clin$`Age at enrollment (days)`)

desmith_clin %>% 
  select(matches(misclassified)) %>% 
  filter(if_any(everything(), ~ !str_detect(., "[:digit:]")))
``` 

- let Age at enrollment be `NA` since it is all <15
- replace nk with `NA`
- convert misclassified cols to numeric

```{r}
desmith_cleaned <- desmith_clin %>% 
  mutate(across(matches(misclassified), ~na_if(., "nk"))) %>% 
  mutate(across(matches(misclassified), ~as.numeric(.)))
```


```{r}
character_cols <- desmith_cleaned %>% 
  select(where(is.character)) %>% 
  colnames()


```

Sample info for CHOP:

- find out age at sample collection - how to annotate if <15? 
* should harmonize to data model before sending to CHOP
```{r}
desmith_sample_catalog <- desmith_clin %>% 
  select(WPSID, Sample, `Analyte type`, `Experimental strategy`, `Age at enrollment (days)`)
```



### Hakon:

```{r}
hakon_clin_dna <- read_xlsx(synGet("syn44291080")$path)
hakon_clin_rna <- read_xlsx(synGet("syn44291064")$path)

#delete the last few cols since they are for solid tumors
hakon_clin_all <- hakon_clin_dna %>% 
  rbind(hakon_clin_rna) %>% 
  select(-(`Days to First Event`:last_col()))
```

```{r}
skim(hakon_clin_all) %>% view()
```

Check that IDs are unique:
```{r}
n_distinct(hakon_clin_all$`Participant ID`) #1152 (1140 DNA + 12 that only have RNA)
```

How many participants have both DNA & RNA
```{r}
hakon_clin_all %>% 
  count(`Participant ID`) %>% 
  filter(n == 2) #372
```

Has RNA but not DNA:
```{r}
setdiff(hakon_clin_rna$`Participant ID`, hakon_clin_dna$`Participant ID`) #12
```


### HTP X01

```{r}
htp_x01_wgs <- read_tsv(synGet("syn50912055")$path)
htp_x01_rna <- read_tsv(synGet("syn50912054")$path)
htp_clin <- read_xlsx(synGet("syn25574963")$path)
```


```{r}
setdiff(htp_x01_wgs$collaborator_participant_id, htp_clin$LabID)
# "HTP0721A2" "HTP0760A"  "HTP0888A"  "HTP0748A"  "HTP0755A"

setdiff(htp_x01_rna$individual_alias, htp_clin$LabID) %>% sort()
# "HTP0040B8" "HTP0749B"  "HTP0739A"  "HTP0746A"  "HTP0727A"  "HTP0728A"  "HTP0729A"  "HTP0730A"  "HTP0731A"  "HTP0737A"  "HTP0738A"
```

